### 2008-07-01 doc/documentation/appendix.tex
sorry, forgot to check in
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@367 3e02f596-3131-0410-8352-e04eeaec37cc

		
\begin{verbatim}
WS              = ' ' | ? ISO 6429 character Horizontal Tabulation ?;
NL              = ? ISO 6429 character Carriage Return ?;
IF              = ":-";
NOT             = 'not';
DIGIT           = '0' | '1' | '2' | '3' | '4' | '5' | '6' | '7' | '8' | '9';
DIGIT1          = DIGIT-'0';
UNDERSCORE      = '_';
LETTER          = 'a' | 'b' | 'c' | 'd' | 'e' | 'f' | 'g' | 'h'
                | 'i' | 'j' | 'k' | 'l' | 'm' | 'n' | 'o' | 'p' | 'q'
                | 'r' | 's' | 't' | 'u' | 'v' | 'w' | 'x' | 'y' | 'z';
CAP_LETTER      = 'A' | 'B' | 'C' | 'D' | 'E' | 'F' | 'G' | 'H' | 'I'
                | 'J' | 'K' | 'L' | 'M' | 'N' | 'O' | 'P' | 'Q' | 'R'
                | 'S' | 'T' | 'U' | 'V' | 'W' | 'X' | 'Y' | 'Z';
IN_STRING          = LETTER | CAP_LETTER | UNDERSCORE | DIGIT;

IDENTIFIER      = [LETTER | UNDERSCORE], {IN_STRING};
STRING          = '"' {INSTRING-('"' | 'NL')} '"';
VARIABLE        = [CAP_LETTER],  {IN_STRING};
COMMENT         = '%' {IN_STRING-NL}, NL;
SHOW            = ['#'], 'show';
HIDE            = ['#'], 'hide';
CONST           = ['#'], 'const';
DOMAIN          = ['#'], 'domain';
MINIMIZE        = ['#'], 'minimize';
MAXIMIZE        = ['#'], 'maximize';
COMPUTE         = ['#'], 'compute';
DISJUNCTION     = '|';
LPARA           = '(';
RPARA           = ')';
LBRAC           = '{';
RBRAC           = '}';
LSBRAC          = '[';
RSBRAC          = ']';
DOT             = '.';
DOTS            = '..';
SEMI            = ';';
DDOT            = ':';
COMMA           = ',';
PLUS            = '+';
TILDE           = '~';
OR              = '?';
AND             = '&';
XOR             = '^';
MINUS           = '-';
TIMES           = '*';
DIVIDE          = '/';
ASSIGN          = '=';
EQ              = '==';
NE              = '!=';
GT              = '>';
LT              = '<';
GE              = '>=';
LE              = '<=';

START = PROGRAM;

PROGRAM = PROGRAM, RULE, DOT;
PROGRAM = PROGRAM, SHOW, SHOW_LIST, DOT;
PROGRAM = PROGRAM, HIDE, HIDE_LIST, DOT;
PROGRAM = PROGRAM, DOMAIN, DOMAIN_LIST, DOT;
PROGRAM = PROGRAM, CONST, IDENTIFIER, ASSIGN, CONST_TERM, DOT;
PROGRAM = DOT;

SHOW_LIST = SHOW_LIST, COMMA, SHOW_PREDICATE;
SHOW_LIST = SHOW_PREDICATE;

HIDE_LIST = DOT;
HIDE_LIST = NHIDE_LIST;

NHIDE_LIST = NHIDE_LIST COMMA HIDE_PREDICATE;
NHIDE_LIST = HIDE_PREDICATE;

DOMAIN_LIST = DOMAIN_LIST COMMA DOMAIN_PREDICATE;
DOMAIN_LIST = DOMAIN_PREDICATE;

NEG_PRED = IDENTIFIER;
NEG_PRED = MINUS, IDENTIFIER;

SHOW_PREDICATE = NEG_PRED, LPARA, NEG_PRED, RPARA;
SHOW_PREDICATE = NEG_PRED;
SHOW_PREDICATE = NEG_PRED, DIVIDE, NUMBER;
HIDE_PREDICATE = NEG_PRED, LPARA VARIABLE_LIST, RPARA;
HIDE_PREDICATE = NEG_PRED;
HIDE_PREDICATE = NEG_PRED, DIVIDE, NUMBER;

DOMAIN_PREDICATE = NEG_PRED, LPARA, DOMAIN_LIST, RPARA;

VARIABLE_LIST = VARIABLE_LIST, COMMA, VARIABLE;
VARIABLE_LIST = VARIABLE;

DOMAIN_VAR = VARIABLE;
DOMAIN_VAR = DOMAIN_VAR, SEMI, VARIABLE;

DOMAIN_LIST = DOMAIN_LIST, COMMA, DOMAIN_VAR;
DOMAIN_LIST = DOMAIN_VAR(var);

RULE = HEAD_ATOM IF BODY;
RULE = HEAD_ATOM IF;
RULE = HEAD_ATOM;
RULE = IF BODY;
RULE = IF;
RULE = MAXIMIZE;
RULE = MINIMIZE;
RULE = COMPUTE;

BODY = BODY, COMMA, BODY_LITERAL;
BODY = BODY_LITERAL;

CONDITIONAL = PREDICATE;
CONDITIONAL = RELATION_LITERAL;

CONDITIONAL_LIST = CONDITIONAL_LIST, DDOT, CONDITIONAL;
CONDITIONAL_LIST = DOT;

BODY_LITERAL = BODY_ATOM;
BODY_LITERAL = NOT, BODY_ATOM;
BODY_LITERAL = RELATION_LITERAL;
BODY_LITERAL = VARIABLE, ASSIGN, AGGREGATE;
BODY_LITERAL = AGGREGATE_ATOM;

CONSTRAINT_LITERAL = CONSTRAINT_ATOM;
CONSTRAINT_LITERAL = NOT, CONSTRAINT_ATOM;

BODY_ATOM = PREDICATE, CONDITIONAL_LIST;

(* assignment literals are not realy relation literals but they are used like them *)
RELATION_LITERAL = VARIABLE, ASSIGN, TERM;
RELATION_LITERAL = TERM EQ TERM;
RELATION_LITERAL = TERM NE TERM;
RELATION_LITERAL = TERM GT TERM;
RELATION_LITERAL = TERM GE TERM;
RELATION_LITERAL = TERM LT TERM;
RELATION_LITERAL = TERM LE TERM;

HEAD_ATOM = AGGREGATE_ATOM;
HEAD_ATOM = DISJUNCTION;

DISJUNCTION = DISJUNCTION_LIST;
DISJUNCTION_LIST = DISJUNCTION_LIST, DISJUNCTION, CONSTRAINT_ATOM;
DISJUNCTION_LIST = CONSTRAINT_ATOM;

CONSTRAINT_ATOM = PREDICATE, CONDITIONAL_LIST;

PREDICATE = IDENTIFIER, LPARA, TERM_LIST, RPARA;
PREDICATE = IDENTIFIER;
PREDICATE = MINUS, IDENTIFIER, LPARA, TERM_LIST, RPARA;
PREDICATE = MINUS, IDENTIFIER;
AGGREGATE_ATOM = TERM, AGGREGATE, TERM;
AGGREGATE_ATOM = AGGREGATE, TERM;
AGGREGATE_ATOM = TERM, AGGREGATE;
AGGREGATE_ATOM = AGGREGATE;

TERM_LIST = TERM_LIST, COMMA, TERM;
TERM_LIST = TERM;

TERM = VARIABLE;
TERM = IDENTIFIER;
TERM = STRING;
TERM = NUMBER;
TERM = LPARA, TERM, RPARA;
TERM = TERM, MOD, TERM;
TERM = TERM, PLUS, TERM;
TERM = TERM, TIMES, TERM;
TERM = TERM, MINUS, TERM;
TERM = TERM, DIVIDE, TERM;
TERM = TERM, XOR, TERM;
TERM = TERM, AND, TERM;
TERM = TERM, OR, TERM;
TERM = TILDE, TERM;
TERM = TERM DOTS TERM;
TERM = MINUS TERM; [UMINUS] // TODO: Whats this
TERM = ABS, LPARA, TERM, RPARA;
TERM = TERM, SEMI, TERM;
TERM = IDENTIFIER, LPARA, TERM_LIST, RPARA;

CONST_TERM_LIST = CONST_TERM_LIST COMMA CONST_TERM;
CONST_TERM_LIST = CONST_TERM;

CONST_TERM = IDENTIFIER;
CONST_TERM = STRING;
CONST_TERM = NUMBER;
CONST_TERM = LPARA CONST_TERM, RPARA;
CONST_TERM = CONST_TERM, MOD, CONST_TERM;
CONST_TERM = CONST_TERM, PLUS, CONST_TERM;
CONST_TERM = CONST_TERM, TIMES, CONST_TERM;
CONST_TERM = CONST_TERM, MINUS, CONST_TERM;
CONST_TERM = CONST_TERM, DIVIDE, CONST_TERM;
CONST_TERM = MINUS, CONST_TERM; [UMINUS]  //TODO: whats this
CONST_TERM = ABS, LPARA, CONST_TERM, RPARA;
CONST_TERM = IDENTIFIER, LPARA, CONST_TERM_LIST, RPARA;

AGGREGATE = SUM, LSBRAC WEIGHT_LIST, RSBRAC;
AGGREGATE = LSBRAC, WEIGHT_LIST, RSBRAC;
AGGREGATE = COUNT, LBRAC CONSTR_LIST, RBRAC;
AGGREGATE = LBRAC, CONSTR_LIST, RBRAC;
AGGREGATE = MIN, LSBRAC, WEIGHT_LIST, RSBRAC;
AGGREGATE = MAX, LSBRAC, WEIGHT_LIST, RSBRAC;

COMPUTE  = COMPUTE, LBRAC, CONSTR_LIST, RBRAC;
COMPUTE  = COMPUTE, NUMBER, LBRAC,  CONSTR_LIST, RBRAC;
MINIMIZE = MINIMIZE, LBRAC,  CONSTR_LIST, RBRAC;
MINIMIZE = MINIMIZE, LSBRAC, WEIGHT_LIST, RSBRAC;
MAXIMIZE = MAXIMIZE, LBRAC,  CONSTR_LIST, RBRAC;
MAXIMIZE = MAXIMIZE, LSBRAC, WEIGHT_LIST, RSBRAC;

WEIGHT_LIST = NWEIGHT_LIST;
WEIGHT_LIST = DOT;
NWEIGHT_LIST = NWEIGHT_LIST, COMMA, WEIGHT_TERM;
NWEIGHT_LIST = WEIGHT_TERM;

WEIGHT_TERM = CONSTRAINT_LITERAL, ASSIGN TERM;
WEIGHT_TERM = CONSTRAINT_LITERAL;

CONSTR_LIST = NCONSTR_LIST;
CONSTR_LIST = DOT;
NCONSTR_LIST = NCONSTR_LIST, COMMA CONSTR_TERM;
NCONSTR_LIST = CONSTR_TERM;

CONSTR_TERM = CONSTRAINT_LITERAL;
\end{verbatim}

### 2008-07-01 doc/documentation/paper.tex
sorry, forgot to check in
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@367 3e02f596-3131-0410-8352-e04eeaec37cc

\documentclass[a4paper,10pt,titlepage]{article}

\usepackage[latin1]{inputenc}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage[all]{xy}
\usepackage{longtable}
\usepackage{multirow}

\author{Max Ostrowski}
\title{Gringo 2.0}
\date{\today}

\begin{document}
\maketitle
\tableofcontents
\section{Introduction}
Answer Set Programming (ASP) \cite{baral02a} is a declarative approach for modeling and solving search problems, represented as logic programs.
An ASP system usually consists of a grounder and a solver.
First, a grounder translates a non-ground problem description into a propositional program, which can be processed by the second component of the system, the solver. This is illustrated in figure (\ref{fig:ASP}).
\begin{figure}[h]
\[
\xymatrix{*+[F]\txt{Problem\\representation}\ar[rr]^{Grounder}&&
		*+[F]\txt{Intermediate\\ representation} \ar[rr]^{Solver}&&
		*+[F]\txt{Variable\\ assignment}}
\]
\caption{Architecture of an ASP System}
\label{fig:ASP}
\end{figure}
Gringo2.0 is a state of the art and has many features that makes writing logic programs easy.
It was developed by Roland Kaminski and is orientated on the ideas of gringo1.0.
The goal of this work is to give an overview of the features supported by gringo2.0. First, the general restrictions to the input language are stated. A broad set of potentialities for writing encodings for logic programs exists. Currently, no other grounder support such a rich featured, less restricted input language.
There exist several output formats which can be used be different grounders and even a human readable format for debugging purposes.
Gringo also has an internal interface to the solver clasp \cite{gekanesc07b}, so the output can directly by parsed to the solver and answersets can be computed.
The syntax of gringo2.0 can be found in the appendix \ref{appendix}.

\section{Interface}
Gringo can be started on the command line and has some parameters that can be useful.
The first parameters are gringo parameters and and arbritary number of filenames of the logic program.
\paragraph{--help}
Gives a short overview of all parameters.
\paragraph{--const}
You can set the value of constants defined in you logic program via the command line of gringo.
After the parameter ``--const'' you have to give the name of the constant you want to set followed by a ``='' and the value of that constant. The integer value can even be an arithmetic expression. For example:
\begin{verbatim}
 gringo --const n=5*2
\end{verbatim}
the atom $n$ is replaced by $10$ in the logic program.

\paragraph{--convert}
If this parameter is used, the grounder expects an already grounded program as input.
For converting an already grounded logic program into an output format supported by gringo this is the right option. This mode saves a lot of memory but does not ground the program.
\paragraph{-l}
This mode prints smodels conform output. For a definition of the output format see \cite{lparseManual}.
Since some features of gringo are not natively supported for this output language, new atoms are introduced to create an equivalent program.\newline
This is the current default output format of gringo.
Sum aggregates with negative weights are handled like weight constraint in lparse for this output format. So a transformation as explaned in \cite{lparseManual} is used to get rid of the negative weights.
\paragraph{-c}
In this mode, the output is directly parsed to clasp, an ASP solver. For solving logic programs, no intermediate files are produced and the answersets are printed directly.
With this option enabled, all parameters for clasp are valid and can be declared in the command line. Therefore adouble minus ``--'' must be given after all gringo related parameters are set. After them die clasp parameters follow.
For example:
\begin{verbatim}
 gringo -c --const n=5 -- --supp-models 3
\end{verbatim}
Furthermore all restriction on smodels output al hold as clasps internal interface currently reads smodels format.
\paragraph{-p}
Parameter ``p'' invokes the plain output method, so the output is printed in human readable format, which is also the input format for gringo.
\paragraph{-g}
This parameter invokes the new output format for intermediate languages ASPils \cite{atleastthisreferenceishouldgive}. Currently only the ``Conglomeration'' normal form is supported.
\paragraph{default parameters}
The default parameter is ``-l'' for smodels output.
\section{Language Constructs}
The basic elements of a logic program are atoms, constants, variables and rules.
\paragraph{Constant}
A constant is a representation of an entity in the problem domain.
A constant is either a name, starting with a lower case letter or an underscore, or a positive integer.
It also can be encompassed by double quotation marks. When using quotation marks, which is sometimes helpful when using constants with spaces, it has to be remarked that a constant with quotation marks is never equal to a constant without, so ``$a$'' is not $a$.
\paragraph{Variable}
A variable is programming construct to simplify writing logic programs.
It always starts with a capitalized letter and can contain digits and letters.
In the grounding process a variable is replaced by one or more constants.
Each variable occurring in a rule must be domain restricted as described in section \ref{domainrestricted}.
\paragraph{Atom}

An atom is a predicate with a name. The name can consist of letters, digits and the underscore.
It has to start with a lower case letter or an underscore.
An atom can have several parameters, encapsulated by parenthesis and separated with commas.
The parameters are either constants, variables or function symbols.
A function symbols do have the same syntax as atoms. This way they can be finitely nested.
As an example, an atom with three parameters is given:
The first parameter being a constant, the others being function symbols.
\begin{verbatim}
 p(a,f(b,X), g(f(X),Y)).
\end{verbatim}

\paragraph{Rule}
A rule normally consists of a head and a body and ands with a dot.
The body is separated from the using ``\:\-'' as in
\begin{verbatim}
 head :- body.
\end{verbatim}
The body is a comma separated conjunction of expressions.
The supported expressions are introduced in the next section.
The variables used in the head must be restricted to a domain in the body.
Check (\ref{domainrestricted}) to see the restrictions on domains.
\newline
A rule can occur without a body as a fact, therefore the body is omitted.
\begin{verbatim}
 head.
\end{verbatim}
The same can be done with the body to express an integrity constraint.
\begin{verbatim}
 :- body.
\end{verbatim}

\paragraph{Comments}
Comments included in the input are line based.
Every line starting with ``\%'' is a comment. The whole line is omitted and not to be handled by the grounder.
\paragraph{Warnings}
Several warnings are produced by gringo to give hints on common mistakes.
They are all printed to standard error output.

\begin{itemize}
 \item 
\begin{verbatim}
 "Warning: sum with negative bounds in the body of a rule"
\end{verbatim}
This warning occurs if a sum aggregate with negative weights has been found and the output format smodels is used or the output is redirected to clasp.
 \item 
\begin{verbatim}
 "Warning: multiple definitions of #const n"
\end{verbatim}
This warning is printed if more than one \#const statement is used to set the same constant.
The last attached value is used for grounding.
\item
\begin{verbatim}
 "Warning: p/3 is never defined"
\end{verbatim}
This warnings states that an atom is used in the body of a rule that never occurs in the head.
``p'' is the name of the atom and ``3'' is the arity.
\item
\begin{verbatim}
 "got ground program i hope u have enough memory :)"
\end{verbatim}
Use the command line option ``--convert'' to convert a ground program into another format.

\end{itemize}
Furthermore all warnings from clasp are included.

\paragraph{Errors}
Gringo2.0 stops the execution and prints an error mesaage to standard error output.
\begin{itemize}
 \item 
\begin{verbatim}
 "the following rule cannot not be grounded,
  non domain predicates : { ... }"
\end{verbatim}
This error message gives a list of atoms that are not omega restricted.
 \item 
\begin{verbatim}
 "the following rule cannot not be grounded,
  weakly restricted varibles: { ... }"
\end{verbatim}
This error message gives a list of atoms that are not lambda restricted. 
\item
\begin{verbatim}
 "syntax error on line Y column Z unexpected token: X"
\end{verbatim}
The parser can not read the input format.
\end{itemize}

\subsection{Meta Constructs}
The ``\#'' sign in front of all meta expressions is optional and can therefore be omitted.
\paragraph{hide}
The hide construct is used to hide atoms from the output of the ASP solver.
Two different syntaxes are supported by gringo.
The first 
\begin{verbatim}
 #hide name/arity.
\end{verbatim}
and the second
\begin{verbatim}
 #hide name(X1, ... , Xn).
\end{verbatim}
Both variants hide the predicate ``name''.
The arity of the predicate can be given explicit or implicit with a list of variables.
The variables are just placeholder and do not unify. The second syntax is implemented for compatibility reasons only.
If the predicate that has to be hidden has arity zero, meaning it is an atom, the syntax is
\begin{verbatim}
 #hide name.
\end{verbatim}
Not to give explicitly all predicates that are hidden, only the ones which are visible the statement
\begin{verbatim}
 #hide.
\end{verbatim}
in conjunction with the ``show'' statement is used.
The ``show'' statement has the same syntax as ``hide'' but uses the keyword ``show''.
\begin{verbatim}
 #hide.
 #show p/3.
\end{verbatim}
The example hides all atoms except the predicate $p$ with arity $3$.
\paragraph{const}
Within the ``const'' statement, a default value for constants can be defined.
Constants are atoms used in the logic program and are normally defined via the command line options.
A default value for a constant $n$ can be easily be set.
\begin{verbatim}
 #const n=5.
\end{verbatim}
\paragraph{domain}
A domain statement states that a specific variable has a global domain restriction.
The statement
\begin{verbatim}
 #domain time(T).
\end{verbatim}
makes sure that the variable ``T'' is always bound to the domain of ``time''.
This has the same effect as adding the domain ``time(T)'' to every rule the variable occurs in.
\paragraph{compute}
The compute statement is a hint for the solver which models to compute.
Is has the syntax
\begin{verbatim}
 #compute number {a1, ..., aN, not aN+1, ..., not aM}.
\end{verbatim}
$number$ is the number of models that shall be computed by the answerset solver and can be omitted. If it is zero, no restriction to the number of models is applied.
All positive atoms in the compute statement must be in the computed models, and all negative atoms do not have to be in the models of the program.
The compute statement is solver dependent and supported by every solver.
\subsection{Features}
\paragraph{Expanding Terms}
\subparagraph{Range Operator}
To define several atoms that stretch on one or more parameters over a certain range, two dots can used to express an increasing integer range.
Using the range operator in the head of a rule, it works as a conjunction of all the members in the range, so for every number the rule is duplicated.
\begin{verbatim}
 times(1..3, 1..2).
\end{verbatim}
will result in the facts
\begin{verbatim}
 times(1,1). times(2,1). times(3,1).
 times(1,2). times(2,2). times(3,2).
\end{verbatim}

Using the range operator in the body of a rule, a disjunction of all members in the range is created. So the rules will be duplicated for every member.
\begin{verbatim}
 a :- p(1..3).
\end{verbatim}
will result in the rules
\begin{verbatim}
 a :- p(1).
 a :- p(2).
 a :- p(3).
\end{verbatim}
If the range operator is used in an aggregate, it creates a a comma separated list of atoms, so
\begin{verbatim}
 a :- {b(1..3)}.
\end{verbatim}
results in
\begin{verbatim}
 a :- {b(1), b(2), b(3)}.
\end{verbatim}

\paragraph{Conjunctor ;}
The conjunctor ``;'' will create a conjunction of the its elements.
If used in the head of a rule this means that the rule is duplicated.
If used in the body instead, the atoms will be logically combined using a conjunction.
\begin{verbatim}
 a(X;1..2) :- b(X;a;b).
 b(foo). b(bar).
\end{verbatim}
will results in the rules
\begin{verbatim}
 a(foo,1) :- b(foo), b(a), b(b).
 a(foo,2) :- b(foo), b(a), b(b).
 a(bar,1) :- b(bar), b(a), b(b).
 a(bar,2) :- b(bar), b(a), b(b).
\end{verbatim}
The conjunctor can not be used directly in aggregates, but for example in combination with the expander.
\begin{verbatim}
 { a(X,Y) : p(X;Y) }.
 p(1..2).
\end{verbatim}
Results in
\begin{verbatim}
 {a(1,1), a(1,2), a(2,1), a(2,2)}.
 p(1).
 p(2).
\end{verbatim}

\paragraph{Expander :}
With the expander ``:'' a domain for an atom can be set.
On the left side of the operator an atom with variables is expected and on the right side the domain for that variables is given.
In the head, the atoms get connected via a disjunction, in the body they get connected via a conjunction, so no new rules are created with this operator.
\begin{verbatim}
 a(X,Y) : b(X) : c(Y) :- d(X) : e(X).
 b(1..2).
 c(3).
 e(a;b;c).
\end{verbatim}
would become
\begin{verbatim}
 a(1,3) | a(2,3) :- d(a), d(b), d(c).
\end{verbatim}
This operator can be used in aggregates to create a comma separated list of atoms.

\paragraph{Arithmetic Expressions}
The build in arithmetics can be used for integer arithmetic.
As operators addition (+), subtraction (-), multiplication (*), division (/), modulo (mod) and the absolute value operator (abs) can be used.
Furthermore the bitwise operators and(\&), or(?), xor(\textasciicircum) and invert($\sim$) are supported.
Arithmetic expressions can occur everywhere in head and in the body, even in complex constructs.
As parameters, integer constants or expressions that evaluate to integers are used.
The simple example of using operators and range terms
\begin{verbatim}
 a(X-1 .. X+1) :- b(X,Y), c(abs(Y * 2 * -X)).
 b(1,2).
 c(4).
\end{verbatim}
would result in the grounded program\footnote{gringo would simplify the grounded rule to fact, since all literals in the body are facts}
\begin{verbatim}
 a(0) :- b(1,2), c(4).
 a(1) :- b(1,2), c(4).
 a(2) :- b(1,2), c(4).
 b(1,2).
 c(4).
\end{verbatim}

TODO: Bitwise operators !!!!

\subparagraph{Relational Operators}
The relational operators are ``less than'' ($<$),\\ ``greater than'' ($>$), ``less equal'' ($<=$), ``greater equal'' ($>=$) , ``equal'' ($==$) and ``not equal'' ($!=$).
They can be used to constrain certain variables.
For example, some rule is only generated for all unifications where X is greater than Y-1.
\begin{verbatim}
 a(X) :- b(X,Y), X >= Y-1.
 b(0,2).
 b(4,2).
\end{verbatim}
This would let $b(4,2)$ a valid substitution, since $4 >= 2-1$ but $b(0,2)$ not, so the grounded program could be:
\begin{verbatim}
 a(4) :- b(4,2).
 b(0,2).
 b(4,2).
\end{verbatim}

\subparagraph{Assignment Operator}
As assignment operator ``$=$'' is used.
The variable on the left side is set to the value of the right side.
\begin{verbatim}
 a(X) :- X=Y*2, b(Y).
 b(1..2).
\end{verbatim}
becomes
\begin{verbatim}
 a(2) :- b(1).
 a(4) :- b(2).
 b(1..2).
\end{verbatim}
A cyclic assignment that results in an unbound variable is not allowed and is rejected with an error . For example
\begin{verbatim}
 a(X) :- X=Y, Y=X.
\end{verbatim}

\paragraph{True Negation}
True negation can be used. Therefore the atom is preceded by a minus sign.
\begin{verbatim}
 -a :- not  a.
  a :- not -a. 
\end{verbatim}
\paragraph{Disjunction}
A disjunction can be expressed using the pipe symbol ``$\mid$''.
A disjunction can only occur in the head of a rule.
\begin{verbatim}
 red | green | blue :- colored.
\end{verbatim}
Using disjunctions in logic programs increase expressiveness and complexity of the logic program.
A solver for disjunctive logic programs such as claspD \cite{claspD} or cmodels \cite{cmodels} is needed to solve such programs.

\paragraph{Aggregates}
An aggregate is a set or a multiset of atoms and has a lower and an upper bound.
The general syntax is
\begin{verbatim}
 lower aggregate_name [atom1=w1, ..., atomN=wN] upper
\end{verbatim}
If squared brackets are used, the atom list is interpretated as a set, otherwise, for curly brackets, as a multiset of atoms.
The lower and upper bound is optional, and can be omitted.

\subparagraph{Count}
A ``count'' aggregate is a set of atoms and has an optional lower and upper bound.
It evaluates to true if the number of true atoms in the set is between the lower and upper bound. Only curly brackets may be used for a count aggregate as it represents a set of atoms.
Weights are not allowed in the count aggregate.
For this aggregate a shorthand syntax exists
\begin{verbatim}
 lower {atom1, ..., atomN} upper
\end{verbatim}
The count aggregate is equal to the cardinality constraint known from smodels \cite{smodels}.
\subparagraph{Sum}
The ``sum'' aggregate is a multiset of atoms with weight and has an optional lower and upper bound. It evaluates to true if the sum of the weights of all true atoms in the multiset is between the lower an upper bound.
The sum aggregate is equal to the weight constraints known from smodels, and has therefore a shorthand syntax
\begin{verbatim}
 lower sum [atom1=w1, ..., atomN=wN] upper
\end{verbatim}
If the command line option ``-l'' is used the semantic may change if negative weights exist in the aggregate. The transformation used is described in \cite{lparseManual}.

\subparagraph{Min}
The ``min'' aggregate is determined by the minimum of all weights of atoms that evaluate to true.
For the minimum aggregate a multiset with weighted atoms is used. It has no shorthand syntax.
\subparagraph{Max}
The ``max'' aggregate is similiar to the minimum aggregate but considers the maximum of the weights.

\paragraph{Optimize Statements}

\subparagraph{Minimize/Maximize}
An optimize statement is used to find a minimal or maximal answerset of the logic program.
The statement can be weighted or not, therefore weights can be omitted in the following syntax.
\begin{verbatim}
 minimize [atom1 = w1, ..., atomN = wN].
\end{verbatim}
The maximize statement uses the keyword ``maximize'' instead.

\section{$\Pi$-Restrictiveness (or something else)}
\label{domainrestricted}
\section{Examples}
In this section come encodings for common problems are presented and explained.
\paragraph{Fastfood}
This example needs a proper instance to work.
\begin{verbatim}
#hide.
#show altdepot/2.
#show costs/1.

% calculate the number of depots
depots(X) :- X = { depot(DN, DK) :  depot(DN, DK) }.

% calculate the costs for the initial depots
% (the grounder can calculate this on its own)
costs(C) :- C = [ serves(RN, Dist) : restaurant(RN, RK) : dist(Dist) = Dist ].

% choose the closest depot for a restaurant
% (the grounder can calculate this on its own)
serves(Rname,Dist) :- restaurant(Rname,RK),
   Dist = min [ depot(DN1, DK1) : depot(DN1, DK1) = abs(DK1-RK) ].

dist(abs(X-Y)) :- restaurant(A, X), restaurant(B, Y).

% Each restaurant may be an alternative depot or not.
X { altdepot(R,K) : restaurant(R,K) } X :- depots(X).

% choose the closest depot for a restaurant
altserves(Rname,Dist) :- restaurant(Rname,RK),
   Dist = min [ altdepot(DN1, DK1) : restaurant(DN1, DK1) = abs(DK1-RK) ].

% fail if we havnt found a better solution
 :- costs(C), C [ altserves(R,Dist) : dist(Dist) : restaurant(R,RK) = Dist ].

% thats the easiest way to show the calculated costs
minimize [ altserves(R,Dist) : dist(Dist) : restaurant(R,RK) = Dist ].
\end{verbatim}

\paragraph{Queens}
\begin{verbatim}
#hide.
#show q(X,Y).

#const queens = 200.
#const diags  = queens * 2 - 1.

d(1..queens).
diag(1..diags).

% enumerate the diagonals
diagX(X,Y,X - Y + queens) :- d(X), d(Y).
diagY(X,Y,X + Y - 1) :- d(X), d(Y).

% exactly 1 queen in each row/col
1 { q(X,Y) : d(Y) } 1 :- d(X).
1 { q(X,Y) : d(X) } 1 :- d(Y).

% at most 1 queen on each diagonal
{ q(X,Y) : diagX(X,Y,D) } 1 :- diag(D).
{ q(X,Y) : diagY(X,Y,D) } 1 :- diag(D).

\end{verbatim}

\section{Benchmarks}
TODO: run on what machine, give reference to encodings.

\input{benchmark2.result}
\appendixname
\begin{appendix}
\label{appendix}
\input{appendix}
\end{appendix}

\bibliographystyle{plain}
\bibliography{bibtex}
\appendix
%todo: input grammar in EBNF
\end{document}

### 2008-07-01 lib/gringo/include/statistic.h
last changes, do not work for lower bounds
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@368 3e02f596-3131-0410-8352-e04eeaec37cc

@@ -31,8 +31,10 @@ namespace NS_GRINGO
 		/// Constructor
 		Statistic();
 		virtual void print(std::ostream* out) const;
-		void startWeightList();
-		void endWeightList();
+		void startWeight();
+		void endWeight();
+		void startBound();
+		void endBound();
 		void possibleNegativeVariable();
 		void possibleNegative();
 		/// Destructor
@@ -59,7 +61,8 @@ namespace NS_GRINGO
 		// true if a variable exists in a weight
 		bool varInWeight_;
 	private:
-		bool weightList_;
+		bool weight_;
+		bool bound_;
 	};
 }
 

### 2008-07-01 lib/gringo/src/statistic.cpp

@@ -35,15 +35,17 @@ void Statistic::print(std::ostream* out) const
 
 void Statistic::possibleNegativeVariable()
 {
-	std::cout << "possNegativeVar" << std::endl;
-	if (weightList_)
+	if (weight_)
+	{
+		std::cout << "possNegativeVar" << std::endl;
 		varInWeight_ = true;
+	}
 }
 
 void Statistic::possibleNegative()
 {
 	std::cout << "possNegative" << std::endl;
-	if (weightList_)
+	if (weight_)
 	{
 		negative_ = true;
 	}
@@ -52,16 +54,28 @@ void Statistic::possibleNegative()
 }
 
 
-void Statistic::startWeightList()
+void Statistic::startWeight()
 {
 	std::cout << "startWeightList" << std::endl;
-	weightList_ = true;
+	weight_ = true;
 }
 
-void Statistic::endWeightList()
+void Statistic::endWeight()
 {
 	std::cout << "EndWeightList" << std::endl;
-	weightList_ = false;
+	weight_ = false;
+}
+
+void Statistic::startBound()
+{
+	std::cout << "startBound" << std::endl;
+	bound_ = true;
+}
+
+void Statistic::endBound()
+{
+	std::cout << "endBound" << std::endl;
+	bound_ = false;
 }
 
 Statistic::~Statistic()

### 2008-07-01 lib/gringo/include/aggregateliteral.h
- improved handling of aggregates   (at least i hope so)
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@369 3e02f596-3131-0410-8352-e04eeaec37cc

@@ -43,6 +43,7 @@ namespace NS_GRINGO
 		virtual double heuristicValue();
 		virtual void setBounds(Term *lower, Term *upper);
 		virtual void setEqual(Constant *equal);
+		virtual void setEqual(int bound);
 		virtual bool checkBounds(int lower, int upper);
 		/**
 		 * \brief This function returns if the aggregate is a fact
@@ -62,6 +63,11 @@ namespace NS_GRINGO
 		Term *upper_;
 		Constant *equal_;
 		bool fact_;
+	public:
+		int lowerBound_;
+		int upperBound_;
+		int minLowerBound_;
+		int maxUpperBound_;
 	};
 }
 

### 2008-07-01 lib/gringo/src/aggregateliteral.cpp

@@ -69,24 +69,30 @@ bool AggregateLiteral::checkO(LiteralVector &unsolved)
 	return true;
 }
 
+void AggregateLiteral::setEqual(int bound)
+{
+	lowerBound_ = bound;
+	upperBound_ = bound;
+}
+
 bool AggregateLiteral::checkBounds(int lower, int upper)
 {
-	int lowerBound = lower_ ? (int)lower_->getValue() : lower;
-	int upperBound = upper_ ? (int)upper_->getValue() : upper;
+	lowerBound_ = lower_ ? std::max((int)lower_->getValue(), lower) : lower;
+	upperBound_ = upper_ ? std::min((int)upper_->getValue(), upper) : upper;
 	// stupid bounds
-	if(lowerBound > upperBound)
+	if(lowerBound_ > upperBound_)
 		return getNeg();
 	// the aggregate lies completely in the intervall
 	// ---|         |--- <- Bounds
 	// ------|   |------ <- Aggregate
-	if(lower >= lowerBound && upper <= upperBound)
+	if(lower >= lowerBound_ && upper <= upperBound_)
 	{
 		return !getNeg();
 	}
 	// the intervals dont intersect
 	// ----------|   |--- <- Bounds
 	// ---|   |---------- <- Aggregate
-	if(upper < lowerBound || lower > upperBound)
+	if(upper < lowerBound_ || lower > upperBound_)
 		return getNeg();
 	// the intervalls intersect
 	return true;
@@ -191,6 +197,7 @@ namespace
 		if(current_ <= upper_)
 		{
 			g->g_->setValue(var_, Value(current_), binder);
+			l_->setEqual(current_); 
 			status = SuccessfulMatch;
 		}
 		else
@@ -203,6 +210,7 @@ namespace
 		{
 			current_++;
 			g->g_->setValue(var_, Value(current_), binder);
+			l_->setEqual(current_); 
 			status = SuccessfulMatch;
 		}
 		else

### 2008-07-01 lib/gringo/src/countaggregate.cpp

@@ -55,6 +55,8 @@ void CountAggregate::match(Grounder *g, int &lower, int &upper, int &fixed)
 	lower = 0;
 	upper = 0;
 	fixed = 0;
+	maxUpperBound_ = 0;
+	minLowerBound_ = 0;
 	for(ConditionalLiteralVector::iterator it = literals_->begin(); it != literals_->end(); it++)
 	{
 		ConditionalLiteral *p = *it;
@@ -79,6 +81,7 @@ void CountAggregate::match(Grounder *g, int &lower, int &upper, int &fixed)
 				fact_ = false;
 				upper++;
 			}
+			maxUpperBound_++;
 		}
 	}
 	lower+= fixed;
@@ -119,15 +122,17 @@ NS_OUTPUT::Object *CountAggregate::convert()
 			lits.push_back(p->convert());
 	}
 	NS_OUTPUT::Aggregate *a;
-	if(lower_ && upper_)
-		a = new NS_OUTPUT::Aggregate(getNeg(), NS_OUTPUT::Aggregate::COUNT, lower_->getValue(), lits, weights, upper_->getValue());
-	else if(lower_)
-		a = new NS_OUTPUT::Aggregate(getNeg(), NS_OUTPUT::Aggregate::COUNT, lower_->getValue(), lits, weights);
-	else if(upper_)
-		a = new NS_OUTPUT::Aggregate(getNeg(), NS_OUTPUT::Aggregate::COUNT, lits, weights, upper_->getValue());
+	bool hasUpper = upperBound_ < maxUpperBound_;
+	bool hasLower = lowerBound_ > minLowerBound_;
+
+	if(hasLower && hasUpper)
+		a = new NS_OUTPUT::Aggregate(getNeg(), NS_OUTPUT::Aggregate::COUNT, lowerBound_, lits, weights, upperBound_);
+	else if(hasLower)
+		a = new NS_OUTPUT::Aggregate(getNeg(), NS_OUTPUT::Aggregate::COUNT, lowerBound_, lits, weights);
+	else if(hasUpper)
+		a = new NS_OUTPUT::Aggregate(getNeg(), NS_OUTPUT::Aggregate::COUNT, lits, weights, upperBound_);
 	else
 		a = new NS_OUTPUT::Aggregate(getNeg(), NS_OUTPUT::Aggregate::COUNT, lits, weights);
-
 	return a;
 }
 

### 2008-07-01 lib/gringo/src/maxaggregate.cpp

@@ -37,6 +37,8 @@ void MaxAggregate::match(Grounder *g, int &lower, int &upper, int &fixed)
 	lower = INT_MAX;
 	upper = INT_MIN;
 	fixed = INT_MIN;
+	maxUpperBound_ = INT_MAX;
+	minLowerBound_ = INT_MIN;
 	for(ConditionalLiteralVector::iterator it = literals_->begin(); it != literals_->end(); it++)
 	{
 		ConditionalLiteral *p = *it;
@@ -57,6 +59,8 @@ void MaxAggregate::match(Grounder *g, int &lower, int &upper, int &fixed)
 				lower = std::min(lower, weight);
 				upper = std::max(upper, weight);
 			}
+			maxUpperBound_ = std::max(maxUpperBound_, weight);
+			minLowerBound_ = std::min(minLowerBound_, weight);
 		}
 	}
 
@@ -128,6 +132,7 @@ namespace
 
 		current_ = set_.begin();
 		g->g_->setValue(var_, Value(*current_), binder);
+		l_->setEqual(*current_); 
 		status = SuccessfulMatch;
 	}
 
@@ -137,6 +142,7 @@ namespace
 		if(current_ != set_.end())
 		{
 			g->g_->setValue(var_, Value(*current_), binder);
+			l_->setEqual(*current_); 
 			status = SuccessfulMatch;
 		}
 		else
@@ -183,12 +189,14 @@ NS_OUTPUT::Object *MaxAggregate::convert()
 		}
 	}
 	NS_OUTPUT::Aggregate *a;
-	if(lower_ && upper_)
-		a = new NS_OUTPUT::Aggregate(getNeg(), NS_OUTPUT::Aggregate::MIN, lower_->getValue(), lits, weights, upper_->getValue());
-	else if(lower_)
-		a = new NS_OUTPUT::Aggregate(getNeg(), NS_OUTPUT::Aggregate::MIN, lower_->getValue(), lits, weights);
-	else if(upper_)
-		a = new NS_OUTPUT::Aggregate(getNeg(), NS_OUTPUT::Aggregate::MIN, lits, weights, upper_->getValue());
+	bool hasUpper = upper_ && (upperBound_ < maxUpperBound_);
+	bool hasLower = lower_ && (lowerBound_ > minLowerBound_);
+	if(hasLower && hasUpper)
[…]

### 2008-07-01 lib/gringo/src/minaggregate.cpp

@@ -37,6 +37,8 @@ void MinAggregate::match(Grounder *g, int &lower, int &upper, int &fixed)
 	lower = INT_MAX;
 	upper = INT_MIN;
 	fixed = INT_MAX;
+	maxUpperBound_ = INT_MAX;
+	minLowerBound_ = INT_MIN;
 	for(ConditionalLiteralVector::iterator it = literals_->begin(); it != literals_->end(); it++)
 	{
 		ConditionalLiteral *p = *it;
@@ -57,6 +59,8 @@ void MinAggregate::match(Grounder *g, int &lower, int &upper, int &fixed)
 				lower = std::min(lower, weight);
 				upper = std::max(upper, weight);
 			}
+			maxUpperBound_ = std::max(maxUpperBound_, weight);
+			minLowerBound_ = std::min(minLowerBound_, weight);
 		}
 	}
 
@@ -109,6 +113,7 @@ namespace
 
 		current_ = set_.begin();
 		g->g_->setValue(var_, Value(*current_), binder);
+		l_->setEqual(*current_); 
 		status = SuccessfulMatch;
 	}
 
@@ -119,6 +124,7 @@ namespace
 		if(current_ != set_.end())
 		{
 			g->g_->setValue(var_, Value(*current_), binder);
+			l_->setEqual(*current_); 
 			status = SuccessfulMatch;
 		}
 		else
@@ -184,15 +190,17 @@ NS_OUTPUT::Object *MinAggregate::convert()
 		}
 	}
 	NS_OUTPUT::Aggregate *a;
-	if(lower_ && upper_)
-		a = new NS_OUTPUT::Aggregate(getNeg(), NS_OUTPUT::Aggregate::MIN, lower_->getValue(), lits, weights, upper_->getValue());
-	else if(lower_)
-		a = new NS_OUTPUT::Aggregate(getNeg(), NS_OUTPUT::Aggregate::MIN, lower_->getValue(), lits, weights);
-	else if(upper_)
-		a = new NS_OUTPUT::Aggregate(getNeg(), NS_OUTPUT::Aggregate::MIN, lits, weights, upper_->getValue());
+	bool hasUpper = upperBound_ < maxUpperBound_;
+	bool hasLower = lowerBound_ > minLowerBound_;
+
+	if(hasLower && hasUpper)
[…]

### 2008-07-01 lib/gringo/src/statistic.cpp

@@ -37,14 +37,14 @@ void Statistic::possibleNegativeVariable()
 {
 	if (weight_)
 	{
-		std::cout << "possNegativeVar" << std::endl;
+		std::cerr << "possNegativeVar" << std::endl;
 		varInWeight_ = true;
 	}
 }
 
 void Statistic::possibleNegative()
 {
-	std::cout << "possNegative" << std::endl;
+	std::cerr << "possNegative" << std::endl;
 	if (weight_)
 	{
 		negative_ = true;
@@ -56,25 +56,25 @@ void Statistic::possibleNegative()
 
 void Statistic::startWeight()
 {
-	std::cout << "startWeightList" << std::endl;
+	std::cerr << "startWeightList" << std::endl;
 	weight_ = true;
 }
 
 void Statistic::endWeight()
 {
-	std::cout << "EndWeightList" << std::endl;
+	std::cerr << "EndWeightList" << std::endl;
 	weight_ = false;
 }
 
 void Statistic::startBound()
 {
-	std::cout << "startBound" << std::endl;
+	std::cerr << "startBound" << std::endl;
 	bound_ = true;
 }
 
 void Statistic::endBound()
 {
-	std::cout << "endBound" << std::endl;
+	std::cerr << "endBound" << std::endl;
 	bound_ = false;
 }
 

### 2008-07-01 lib/gringo/src/sumaggregate.cpp

@@ -33,6 +33,8 @@ void SumAggregate::match(Grounder *g, int &lower, int &upper, int &fixed)
 	lower = 0;
 	upper = 0;
 	fixed = 0;
+	maxUpperBound_ = 0;
+	minLowerBound_ = 0;
 	for(ConditionalLiteralVector::iterator it = literals_->begin(); it != literals_->end(); it++)
 	{
 		ConditionalLiteral *p = *it;
@@ -55,6 +57,10 @@ void SumAggregate::match(Grounder *g, int &lower, int &upper, int &fixed)
 				else
 					lower+= weight;
 			}
+			if(weight > 0)
+				maxUpperBound_+= weight;
+			else
+				minLowerBound_+= weight;
 		}
 	}
 	lower+= fixed;
@@ -98,12 +104,15 @@ NS_OUTPUT::Object *SumAggregate::convert()
 		}
 	}
 	NS_OUTPUT::Aggregate *a;
-	if(lower_ && upper_)
-		a = new NS_OUTPUT::Aggregate(getNeg(), NS_OUTPUT::Aggregate::SUM, lower_->getValue(), lits, weights, upper_->getValue());
-	else if(lower_)
-		a = new NS_OUTPUT::Aggregate(getNeg(), NS_OUTPUT::Aggregate::SUM, lower_->getValue(), lits, weights);
-	else if(upper_)
-		a = new NS_OUTPUT::Aggregate(getNeg(), NS_OUTPUT::Aggregate::SUM, lits, weights, upper_->getValue());
+	bool hasUpper = upper_ && (upperBound_ < maxUpperBound_);
+	bool hasLower = lower_ && (lowerBound_ > minLowerBound_);
+
+	if(hasLower && hasUpper)
+		a = new NS_OUTPUT::Aggregate(getNeg(), NS_OUTPUT::Aggregate::SUM, lowerBound_, lits, weights, upperBound_);
+	else if(hasLower)
+		a = new NS_OUTPUT::Aggregate(getNeg(), NS_OUTPUT::Aggregate::SUM, lowerBound_, lits, weights);
+	else if(hasUpper)
+		a = new NS_OUTPUT::Aggregate(getNeg(), NS_OUTPUT::Aggregate::SUM, lits, weights, upperBound_);
 	else
 		a = new NS_OUTPUT::Aggregate(getNeg(), NS_OUTPUT::Aggregate::SUM, lits, weights);
 

### 2008-07-06 lib/clasp/src/program_rule.cpp
- renamed several classes - splitted the scc into scc and program - the statementdependencygraph will be deleted   directly after it was build now
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@374 3e02f596-3131-0410-8352-e04eeaec37cc

@@ -97,6 +97,9 @@ PrgRule::RData PrgRule::simplify(RuleState& rs) {
 	if (type_ != CONSTRAINTRULE && type_ != WEIGHTRULE) bound_ = (weight_t)body.size();
 	if (bound_ <= 0) {
 		body.clear();	// body is satisfied - ignore contained lits
+		if (type_ == CONSTRAINTRULE || type_ == WEIGHTRULE) {
+			type_ = BASICRULE;
+		}
 		r.value_ = value_true;
 	}
 	if (bound_ > r.sumWeight) {
@@ -260,6 +263,12 @@ uint32 PrgRule::transformAggregate(ProgramBuilder& prg, TransformationMode m) {
 uint32 PrgRule::transformAggregateQuad(ProgramBuilder& prg, const WeightVec& sumWeights) {
 	typedef std::pair<uint32, weight_t> Key;
 	typedef std::deque<Key> KeyQueue;
+	if (bound_ == 0) {
+		PrgRule r(BASICRULE);
+		r.addHead(heads[0]);
+		prg.addRule(r);
+		return 1;
+	}
 	VarVec atoms(body.size() * bound_, varMax);
 	atoms[bound_-1] = heads[0];
 	KeyQueue todo;
@@ -317,6 +326,7 @@ uint32 PrgRule::transformAggregateExp(ProgramBuilder& prg, const WeightVec& sumW
 	uint32		end		= (uint32)body.size();
 	weight_t	cw		= 0;
 	uint32		next	= 0;
+	if (next == end) { prg.addRule(r); return 1; }
 	while (next != end) {
 		r.addToBody(body[next].first.var(), body[next].first.sign() == false);
 		weights.push_back( body[next].second );

### 2008-07-06 lib/gringo/include/aggregateliteral.h

@@ -29,7 +29,7 @@ namespace NS_GRINGO
 	public:
 		AggregateLiteral(ConditionalLiteralVector *literals);
 		AggregateLiteral(const AggregateLiteral &a);
-		virtual Node *createNode(DependencyGraph *dg, Node *prev, DependencyAdd todo);
+		virtual SDGNode *createNode(SDG *dg, SDGNode *prev, DependencyAdd todo);
 		virtual void createNode(LDGBuilder *dg, bool head);
 		virtual void getVars(VarSet &vars) const;
 		virtual bool checkO(LiteralVector &unsolved);

### 2008-07-06 lib/gringo/include/assignmentliteral.h

@@ -28,7 +28,7 @@ namespace NS_GRINGO
 	public:
 		AssignmentLiteral(Constant *c, Term *t);
 		AssignmentLiteral(const AssignmentLiteral &r);
-		virtual Node *createNode(DependencyGraph *dg, Node *prev, DependencyAdd todo);
+		virtual SDGNode *createNode(SDG *dg, SDGNode *prev, DependencyAdd todo);
 		virtual void createNode(LDGBuilder *dg, bool head);
 		virtual void getVars(VarSet &vars) const;
 		virtual bool checkO(LiteralVector &unsolved);

### 2008-07-06 lib/gringo/include/computeliteral.h

@@ -29,7 +29,7 @@ namespace NS_GRINGO
 	public:
 		ComputeLiteral(ConditionalLiteralVector *literals, int number);
 		ComputeLiteral(const ComputeLiteral &a);
-		virtual Node *createNode(DependencyGraph *dg, Node *prev, DependencyAdd todo);
+		virtual SDGNode *createNode(SDG *dg, SDGNode *prev, DependencyAdd todo);
 		virtual void createNode(LDGBuilder *dg, bool head);
 		virtual void getVars(VarSet &vars) const;
 		virtual bool checkO(LiteralVector &unsolved);

### 2008-07-06 lib/gringo/include/gringo.h

@@ -80,13 +80,9 @@ namespace NS_GRINGO
 	typedef std::vector<std::string*> StringVector;
 	typedef std::vector<Constant*> ConstantVector;
 
-	class SCC;
-	typedef std::vector<SCC*> SCCVector;
-	typedef std::set<SCC*> SCCSet;
-
 	class AggregateLiteral;
 
-	class DependencyGraph;
+	class SDG;
 
 	class GrinGoParser;
 	class LparseParser;
@@ -101,8 +97,8 @@ namespace NS_GRINGO
 	typedef std::vector<Literal*> LiteralVector;
 	typedef std::set<Literal*> LiteralSet;
 
-	class Node;
-	typedef std::vector<Node*> NodeVector;
+	class SDGNode;
+	typedef std::vector<SDGNode*> SDGNodeVector;
 
 	class Statement;
 	typedef std::vector<Statement*> StatementVector;
@@ -125,6 +121,9 @@ namespace NS_GRINGO
 	{ 
 		return os.put(os.widen('\n'));
 	}
+
+	class Program;
+	typedef std::vector<Program*> ProgramVector;
 
 	class Domain;
 	typedef std::vector<Domain*> DomainVector;

### 2008-07-06 lib/gringo/include/node.h

@@ -1,60 +1,0 @@
-// Copyright (c) 2008, Roland Kaminski
-//
-// This file is part of GrinGo.
-//
-// GrinGo is free software: you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation, either version 3 of the License, or
-// (at your option) any later version.
-//
-// GrinGo is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-//
-// You should have received a copy of the GNU General Public License
-// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
-
-#ifndef NODE_H
-#define NODE_H
-
-#include <gringo.h>
-#include <value.h>
-
-namespace NS_GRINGO
-{
-	class Node
-	{
-		friend class DependencyGraph;
-	public:
-		enum Type {STATEMENTNODE = 0, PREDICATENODE = 1};
-	public:
-		Node(Statement *rule);
-		Node(Domain *dom);
-		NodeVector *getDependency() const;
-		NodeVector *getNegDependency() const;
-		Statement *getStatement() const;
-		Domain *getDomain() const;
-		Type getType() const;
-
-		void addDependency(Node *n, bool neg = false);
-		~Node();
-	private:
-		int lowlink_;
-		int index_;
-		unsigned int type_ : 1;
-		unsigned int stacked_ : 1;
-		unsigned int done_ : 1;
-		SCC *scc_;
-		union
-		{
-			Statement  *rule_;
-			Domain     *dom_;
-		};
-		NodeVector dependency_;
-		NodeVector negDependency_;
-	};
-}
-
-#endif
-

### 2008-07-06 lib/gringo/include/normalrule.h

@@ -38,7 +38,7 @@ namespace NS_GRINGO
 		virtual ~NormalRule();
 		// implemented from base class or interface
 		virtual void getVars(VarSet &vars) const;
-		virtual void buildDepGraph(DependencyGraph *dg);
+		virtual void buildDepGraph(SDG *dg);
 		virtual void print(std::ostream &out);
 		virtual bool checkO(LiteralVector &unsolved);
 		virtual bool check(VarVector &free);

### 2008-07-06 lib/gringo/include/program.h

@@ -15,34 +15,30 @@
 // You should have received a copy of the GNU General Public License
 // along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
 
-#ifndef DEPENDENCYGRAPH_H
-#define DEPENDENCYGRAPH_H
+#ifndef PROGRAM_H
+#define PROGRAM_H
 
 #include <gringo.h>
 #include <printable.h>
 
 namespace NS_GRINGO
 {
-	class DependencyGraph
+	class Program : public Printable
 	{
 	public:
-		DependencyGraph();
-		Node *createStatementNode(Statement *r, bool preserveOrder = false);
-		Node *createPredicateNode(PredicateLiteral *pred);
-		NodeVector &getPredNodes();
-		void calcSCCs();
+		enum Type {FACT=0, BASIC=1, NORMAL=2};
+	public:
+		Program(Type type, StatementVector &rules);
+		void print(std::ostream &out);
+		Evaluator *getEvaluator();
 		bool check(Grounder *g);
-		virtual ~DependencyGraph();
-	protected:
-		void calcSCCDep(Node *v1, SCC *scc, bool &root);
-		void tarjan(Node *v1, std::stack<Node*> &stack, int &index);
-	protected:
-		int uids_;
-		NodeVector ruleNodes_;
-		NodeVector predicateNodes_;
-		SCCVector sccs_;
-		SCCSet sccRoots_;
-		Node *last_;
+		StatementVector *getStatements();
+		Type getType();
+		~Program();
+	private:
+		Type type_;
+		StatementVector rules_;
+		Evaluator *eval_;
 	};
 }
 

### 2008-07-06 lib/gringo/include/relationliteral.h

@@ -30,7 +30,7 @@ namespace NS_GRINGO
 	public:
 		RelationLiteral(RelationType type, Term *a, Term *b);
 		RelationLiteral(const RelationLiteral &r);
-		virtual Node *createNode(DependencyGraph *dg, Node *prev, DependencyAdd todo);
+		virtual SDGNode *createNode(SDG *dg, SDGNode *prev, DependencyAdd todo);
 		virtual void createNode(LDGBuilder *dg, bool head);
 		virtual void getVars(VarSet &vars) const;
 		virtual bool checkO(LiteralVector &unsolved);

### 2008-07-06 lib/gringo/include/statement.h

@@ -33,10 +33,10 @@ namespace NS_GRINGO
 		/// Constructor
 		Statement();
 		/**
-		 * \brief Adds the needed dependencies to the DependencyGraph
-		 * \param dg The DependencyGraph
+		 * \brief Adds the needed dependencies to the SDG
+		 * \param dg The SDG
 		 */
-		virtual void buildDepGraph(DependencyGraph *dg) = 0;
+		virtual void buildDepGraph(SDG *dg) = 0;
 		/**
 		 * \brief Returns all variables occuring in the Statement
 		 * \param vars Reference to the result

### 2008-07-06 lib/gringo/include/statementdependencygraph.h

@@ -1,0 +1,98 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#ifndef DEPENDENCYGRAPH_H
+#define DEPENDENCYGRAPH_H
+
+#include <gringo.h>
+#include <printable.h>
+
+namespace NS_GRINGO
+{
+	class SDG
+	{
+		friend class SDGNode;
+	private:
+		class SCC;
+		typedef std::vector<SCC*> SCCVector;
+		typedef std::set<SCC*> SCCSet;
+
+		class SCC
+		{
+		public:
+			enum Type {FACT=0, BASIC=1, NORMAL=2};
+		public:
+			SCC();
+		public:
+			Type type_;
+			int edges_;
+			StatementVector rules_;
+			SCCSet sccs_;
+		};
+	public:
+		SDG();
+		SDGNode *createStatementNode(Statement *r, bool preserveOrder = false);
+		SDGNode *createPredicateNode(PredicateLiteral *pred);
+		void calcSCCs(Grounder *g);
+		~SDG();
+	private:
+		void calcSCCDep(SDGNode *v1, SCC *scc, bool &root);
+		void tarjan(SDGNode *v1, std::stack<SDGNode*> &stack, int &index);
+	private:
+		int uids_;
+		SDGNodeVector ruleNodes_;
+		SDGNodeVector predicateNodes_;
+		SCCVector sccs_;
+		SCCSet sccRoots_;
+		SDGNode *last_;
+	};
+	
+	class SDGNode
+	{
+		friend class SDG;
+	public:
+		enum Type {STATEMENTNODE = 0, PREDICATENODE = 1};
+	public:
+		SDGNode(Statement *rule);
+		SDGNode(Domain *dom);
[…]

### 2008-07-06 lib/gringo/src/conditionalliteral.cpp

@@ -19,7 +19,6 @@
 #include "predicateliteral.h"
 #include "grounder.h"
 #include "term.h"
-#include "node.h"
 #include "indexeddomain.h"
 #include "dlvgrounder.h"
 #include "value.h"
@@ -73,9 +72,9 @@ bool ConditionalLiteral::checkO(LiteralVector &unsolved)
 	return unsolved.size() == 0;
 }
 
-Node *ConditionalLiteral::createNode(DependencyGraph *dg, Node *prev, DependencyAdd todo)
+SDGNode *ConditionalLiteral::createNode(SDG *dg, SDGNode *prev, DependencyAdd todo)
 {
-	Node *n = pred_->createNode(dg, prev, todo);
+	SDGNode *n = pred_->createNode(dg, prev, todo);
 	if(conditionals_)
 	{
 		for(LiteralVector::iterator it = conditionals_->begin(); it != conditionals_->end(); it++)

### 2008-07-06 lib/gringo/src/grounder.cpp

@@ -20,8 +20,8 @@
 #include "constant.h"
 #include "predicateliteral.h"
 #include "normalrule.h"
-#include "dependencygraph.h"
-#include "scc.h"
+#include "statementdependencygraph.h"
+#include "program.h"
 #include "value.h"
 #include "evaluator.h"
 #include "domain.h"
@@ -29,7 +29,7 @@
 
 using namespace NS_GRINGO;
 
-Grounder::Grounder(NS_OUTPUT::Output *output) : internalVars_(0), depGraph_(new DependencyGraph()), output_(output), eval_(0)
+Grounder::Grounder(NS_OUTPUT::Output *output) : internalVars_(0), output_(output), eval_(0)
 {
 }
 
@@ -65,10 +65,19 @@ void Grounder::addDomains(std::string *id, std::vector<StringVector*>* list)
 
 void Grounder::buildDepGraph()
 {
+	SDG dg;
 	for(StatementVector::iterator it = rules_.begin(); it != rules_.end(); it++)
-	{
-		(*it)->buildDepGraph(depGraph_);
-	}
+		(*it)->buildDepGraph(&dg);
+	reset(true);
+	dg.calcSCCs(this);
+}
+
+bool Grounder::check()
+{
+	for(ProgramVector::iterator it = sccs_.begin(); it != sccs_.end(); it++)
+		if(!(*it)->check(this))
+			return false;
+	return true;
 }
 
 void Grounder::addDomains()
@@ -137,12 +146,10 @@ void Grounder::start()
 	std::cerr << "done" << std::endl;
 	std::cerr << "building dependencygraph ... " << std::endl;
 	buildDepGraph();
-	reset(true);
-	depGraph_->calcSCCs();
 	std::cerr << "done" << std::endl;
 	std::cerr << "checking ... " << std::endl;
-	if(!depGraph_->check(this))
-		return;
+	if(!check())
+		throw GrinGoException("Error: the program is not groundable.");
 	reset(false);
 	std::cerr << "done" << std::endl;
 	std::cerr << "grounding ... " << std::endl;
@@ -150,7 +157,7 @@ void Grounder::start()
 	std::cerr << "done" << std::endl;
 }
 
-void Grounder::addSCC(SCC *scc)
+void Grounder::addProgram(Program *scc)
 {
 	sccs_.push_back(scc);
 }
@@ -160,10 +167,10 @@ void Grounder::ground()
 	output_->initialize(getPred());
 	substitution_.resize(varMap_.size() + 2);
[…]

### 2008-07-06 lib/gringo/src/literalstatement.cpp

@@ -16,10 +16,9 @@
 // along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
 
 #include "literalstatement.h"
-#include "dependencygraph.h"
+#include "statementdependencygraph.h"
 #include "literaldependencygraph.h"
 #include "conditionalliteral.h"
-#include "node.h"
 #include "output.h"
 #include "grounder.h"
 #include "evaluator.h"
@@ -30,10 +29,10 @@ LiteralStatement::LiteralStatement(Literal *lit, bool preserveOrder) : lit_(lit)
 {
 }
 
-void LiteralStatement::buildDepGraph(DependencyGraph *dg)
+void LiteralStatement::buildDepGraph(SDG *dg)
 {
 	// the order of optimize statements is important
-	Node *prev = dg->createStatementNode(this, preserveOrder_);
+	SDGNode *prev = dg->createStatementNode(this, preserveOrder_);
 	lit_->createNode(dg, prev, Literal::ADD_NOTHING);
 }
 

### 2008-07-06 lib/gringo/src/node.cpp

@@ -1,73 +1,0 @@
-// Copyright (c) 2008, Roland Kaminski
-//
-// This file is part of GrinGo.
-//
-// GrinGo is free software: you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation, either version 3 of the License, or
-// (at your option) any later version.
-//
-// GrinGo is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-//
-// You should have received a copy of the GNU General Public License
-// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
-
-#include "node.h"
-#include "scc.h"
-
-using namespace NS_GRINGO;
-
-Node::Node(Domain *domain) : lowlink_(-1), index_(-1), type_(PREDICATENODE), stacked_(0), done_(0), scc_(0), dom_(domain)
-{
-}
-
-Node::Node(Statement *rule) : lowlink_(-1), index_(-1), type_(STATEMENTNODE), stacked_(0), done_(0), scc_(0), rule_(rule)
-{
-}
-
-NodeVector *Node::getNegDependency() const
-{
-	return const_cast<NodeVector *>(&negDependency_);
-}
-
-NodeVector *Node::getDependency() const
-{
-	return const_cast<NodeVector *>(&dependency_);
-}
-
-Domain* Node::getDomain() const
-{
-	if(type_ == PREDICATENODE)
-		return dom_; 
-	else
-		return 0;
-}
-
-Statement* Node::getStatement() const
-{ 
-	if(type_ == STATEMENTNODE)
-		return rule_; 
-	else
-		return 0;
-}
-
-Node::Type Node::getType() const
-{
-	return static_cast<Node::Type>(type_);
-}
-
-void Node::addDependency(Node *n, bool neg)
-{
-	assert(n);
-	if(neg)
-		negDependency_.push_back(n);
-	dependency_.push_back(n);
-}
-
-Node::~Node() 
-{
-}
-

### 2008-07-06 lib/gringo/src/normalrule.cpp

@@ -18,9 +18,8 @@
 #include "normalrule.h"
 #include "literal.h"
 #include "predicateliteral.h"
-#include "dependencygraph.h"
+#include "statementdependencygraph.h"
 #include "literaldependencygraph.h"
-#include "node.h"
 #include "grounder.h"
 #include "output.h"
 #include "dlvgrounder.h"
@@ -33,9 +32,9 @@ NormalRule::NormalRule(Literal *head, LiteralVector *body) : Statement(), head_(
 {
 }
 
-void NormalRule::buildDepGraph(DependencyGraph *dg)
+void NormalRule::buildDepGraph(SDG *dg)
 {
-	Node *rn = dg->createStatementNode(this);
+	SDGNode *rn = dg->createStatementNode(this);
 	if(head_)
 		head_->createNode(dg, rn, Literal::ADD_HEAD_DEP);
 	else

### 2008-07-06 lib/gringo/src/optimizeliteral.cpp

@@ -17,10 +17,9 @@
 
 #include "optimizeliteral.h"
 #include "conditionalliteral.h"
-#include "node.h"
 #include "value.h"
 #include "grounder.h"
-#include "dependencygraph.h"
+#include "statementdependencygraph.h"
 #include "literaldependencygraph.h"
 #include "output.h"
 
@@ -130,14 +129,14 @@ void OptimizeLiteral::getVars(VarSet &vars) const
 		(*it)->getVars(vars);
 }
 
-Node *OptimizeLiteral::createNode(DependencyGraph *dg, Node *prev, DependencyAdd todo)
+SDGNode *OptimizeLiteral::createNode(SDG *dg, SDGNode *prev, DependencyAdd todo)
 {
 	for(ConditionalLiteralVector::iterator it = literals_->begin(); it != literals_->end(); it++)
 	{
 		// since there cant be any cycles through these statements 
 		// its ok to add only positive dependencies
 		// even though negative dependencies wont hurt
-		Node *p = (*it)->createNode(dg, prev, ADD_NOTHING);
+		SDGNode *p = (*it)->createNode(dg, prev, ADD_NOTHING);
 		assert(p);
 		prev->addDependency(p);
 	}

### 2008-07-06 lib/gringo/src/predicateliteral.cpp

@@ -15,7 +15,7 @@
 // You should have received a copy of the GNU General Public License
 // along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
 
-#include "dependencygraph.h"
+#include "statementdependencygraph.h"
 #include "literaldependencygraph.h"
 #include "predicateliteral.h"
 #include "assignmentliteral.h"
@@ -23,7 +23,6 @@
 #include "grounder.h"
 #include "constant.h"
 #include "term.h"
-#include "node.h"
 #include "domain.h"
 #include "rangeterm.h"
 #include "rangeliteral.h"
@@ -57,10 +56,9 @@ bool PredicateLiteral::checkO(LiteralVector &unsolved)
 	return true;
 }
 
-Node *PredicateLiteral::createNode(DependencyGraph *dg, Node *prev, DependencyAdd todo)
+SDGNode *PredicateLiteral::createNode(SDG *dg, SDGNode *prev, DependencyAdd todo)
 {
-	Node *n = dg->createPredicateNode(this);
-	predNode_ = n->getDomain();
+	SDGNode *n = dg->createPredicateNode(this);
 	switch(todo)
 	{
 		case ADD_BODY_DEP:

### 2008-07-06 lib/gringo/src/program.cpp

@@ -15,7 +15,7 @@
 // You should have received a copy of the GNU General Public License
 // along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
 
-#include "scc.h"
+#include "program.h"
 #include "statement.h"
 #include "grounder.h"
 #include "literal.h"
@@ -24,23 +24,24 @@
 
 using namespace NS_GRINGO;
 
-SCC::SCC() : type_(FACT), edges_(0), eval_(0)
+Program::Program(Type type, StatementVector &rules) : type_(type), eval_(0)
 {
+	std::swap(rules, rules_);
 }
 
-Evaluator *SCC::getEvaluator()
+Evaluator *Program::getEvaluator()
 {
 	if(!eval_)
 	{
 		switch(type_)
 		{
-			case SCC::FACT:
+			case Program::FACT:
 				eval_ = new Evaluator();
 				break;
-			case SCC::BASIC:
+			case Program::BASIC:
 				eval_ = new BasicProgramEvaluator();
 				break;
-			case SCC::NORMAL:
+			case Program::NORMAL:
 				eval_ = new Evaluator();
 				break;
 		}
@@ -48,19 +49,19 @@ Evaluator *SCC::getEvaluator()
 	return eval_;
 }
 
-void SCC::print(std::ostream &out)
+void Program::print(std::ostream &out)
 {
 	if(rules_.size() > 0)
 	{
 		switch(type_)
 		{
-			case SCC::FACT:
+			case Program::FACT:
 				out << "% fact program:" << std::endl;
 				break;
-			case SCC::BASIC:
+			case Program::BASIC:
 				out << "% basic program:" << std::endl;
 				break;
-			case SCC::NORMAL:
+			case Program::NORMAL:
 				out << "% normal program:" << std::endl;
 				break;
 		}
@@ -71,7 +72,7 @@ void SCC::print(std::ostream &out)
 	}
 }
 
-bool SCC::check(Grounder *g)
+bool Program::check(Grounder *g)
 {
 	VarVector free;
 	LiteralVector unsolved;
@@ -141,17 +142,17 @@ bool SCC::check(Grounder *g)
 	return true;
 }
 
-StatementVector *SCC::getStatements()
+StatementVector *Program::getStatements()
 {
 	return &rules_;
 }
 
-SCC::SCCType SCC::getType()
+Program::Type Program::getType()
 {
 	return type_;
 }
 
-SCC::~SCC()
+Program::~Program()
 {
 	if(eval_)
 		delete eval_;

### 2008-07-06 lib/gringo/src/rangeliteral.cpp

@@ -28,7 +28,7 @@ RangeLiteral::RangeLiteral(Constant *var, Term *lower, Term *upper) : Literal(),
 {
 }
 
-Node *RangeLiteral::createNode(DependencyGraph *dg, Node *prev, DependencyAdd todo)
+SDGNode *RangeLiteral::createNode(SDG *dg, SDGNode *prev, DependencyAdd todo)
 {
 	return 0;
 }

### 2008-07-06 lib/gringo/src/relationliteral.cpp

@@ -28,7 +28,7 @@ RelationLiteral::RelationLiteral(RelationType type, Term *a, Term *b) : Literal(
 
 }
 
-Node *RelationLiteral::createNode(DependencyGraph *dg, Node *prev, DependencyAdd todo)
+SDGNode *RelationLiteral::createNode(SDG *dg, SDGNode *prev, DependencyAdd todo)
 {
 	return 0;
 }

### 2008-07-07 doc/documentation/appendix.tex
some documentation changes
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@375 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 
-HIDE_LIST = DOT;
 HIDE_LIST = NHIDE_LIST;
@@
 CONDITIONAL_LIST = CONDITIONAL_LIST, DDOT, CONDITIONAL;
-CONDITIONAL_LIST = DOT;
 
@@
 WEIGHT_LIST = NWEIGHT_LIST;
-WEIGHT_LIST = DOT;
 NWEIGHT_LIST = NWEIGHT_LIST, COMMA, WEIGHT_TERM;
@@
 CONSTR_LIST = NCONSTR_LIST;
-CONSTR_LIST = DOT;
 NCONSTR_LIST = NCONSTR_LIST, COMMA CONSTR_TERM;

### 2008-07-07 doc/documentation/paper.tex
some documentation changes
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@375 3e02f596-3131-0410-8352-e04eeaec37cc

@@
-\documentclass[a4paper,10pt,titlepage]{article}
+\documentclass[a4paper,10pt]{article}
 
@@
 Parameter ``p'' invokes the plain output method, so the output is printed in human readable format, which is also the input format for gringo.
-\paragraph{-g}
-This parameter invokes the new output format for intermediate languages ASPils \cite{atleastthisreferenceishouldgive}. Currently only the ``Conglomeration'' normal form is supported.
+\paragraph{-g X}
+This parameter invokes the new output format for the intermediate languages ASPils \cite{atleastthisreferenceishouldgive}.
+The X can be any number from $1$ to $7$ to refer to one of the following normalforms (Simple=1, SModels=2, DisjunctiveLogicProgram=3, CModels=4, CModelsExtended=5, DLV=6, Conglomeration=7).
+If no normalform is specified, conglomeration(7)  is assumed.
+
 \paragraph{default parameters}
@@
 The parser can not read the input format.
+\begin{verbatim}
+ "Disjunction not allowed on this normalform."
+ "Count aggregate/cardinality constraint not allowed
+  in this normalform."
+ "Count aggregate/cardinality constraint not allowed 
+  with non trivial upper bound in this normalform."
+ "Count aggregate/cardinality constraint not allowed 
+  with negative lower bound in this normalform."
+ "Sum aggregate/weight constraint not allowed
+  in this normalform."
+ "Sum aggregate/weight constraint not allowed 
+  with non trivial upper bound in this normalform."
+ "Sum aggregate/weight constraint not allowed 
+  with negative lower bound in this normalform."
+ "Sum aggregate/weight constraint not allowed
+  with negative weights in this normalform."
+ "Sum aggregate/weight constraint not allowed
+  with negative weights in this normalform."
+ "Max aggregate not allowed in this normalform."
+ "Max aggregate not allowed
+  with negative weights in this normalform."
+ "Min aggregate not allowed in this normalform."
+ "Min aggregate not allowed 
+  with negative weights in this normalform."
+ "Optimize statement not allowed in this normalform."
+\end{verbatim}
+%\begin{verbatim}
+% "Times aggregate not allowed in this normalform."
+% "Times aggregate not allowed
+%  with negative weights in this normalform."
+%\end{verbatim}
+These errors belong to the ASPils output format. Another normal form as to be chosen.
 \end{itemize}
@@
 Furthermore the bitwise operators and(\&), or(?), xor(\textasciicircum) and invert($\sim$) are supported.
+The bitwise operators are executed on signed integers using the binary complement for negative numbers.
 Arithmetic expressions can occur everywhere in head and in the body, even in complex constructs.
@@
 \end{verbatim}
-would result in the grounded program\footnote{gringo would simplify the grounded rule to fact, since all literals in the body are facts}
+would result in the grounded program\footnote{gringo would simplify the grounded rules to facts, since all literals in the body are facts}
 \begin{verbatim}
@@
 \end{verbatim}
-
-TODO: Bitwise operators !!!!
 
@@
 \section{Benchmarks}
-TODO: run on what machine, give reference to encodings.
+The benchmarks were run on an Athlon XP 1700+ with 1GB of RAM.
+TODO: Official (non internal wiki) link to the instances.
 

### 2008-07-07 lib/gringo/src/conditionalliteral.cpp
- fixed some problems with negative literals in aggregates in the head of rules
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@377 3e02f596-3131-0410-8352-e04eeaec37cc

@@ -170,12 +170,14 @@ int ConditionalLiteral::count()
 
 void ConditionalLiteral::reset()
 {
-	pred_->reset();
+	if(!getNeg())
+		pred_->reset();
 }
 
 void ConditionalLiteral::finish()
 {
-	pred_->finish();
+	if(!getNeg())
+		pred_->finish();
 }
 
 bool ConditionalLiteral::solved()
@@ -194,8 +196,8 @@ void ConditionalLiteral::ground(Grounder *g)
 			//std::cerr << "creating grounder for: " << this << std::endl;
 			grounder_ = new DLVGrounder(g, this, conditionals_->size(), dg_, dg_->getGlobalVars());
 		}
-		
 		grounder_->ground();
+
 	}
 	else
 	{

### 2008-07-07 lib/gringo/src/output.cpp

@@ -144,9 +144,14 @@ Atom::Atom(bool neg, int predUid) : Object(neg), node_(0), predUid_(predUid)
 
 void Atom::addDomain(bool fact)
 {
-	if(fact)
-		node_->addFact(values_);
-	node_->addDomain(values_);
+	// atoms may occur negatively in aggregates
+	// no domain has to be added if the atom is negative
+	if(!neg_)
+	{
+		if(fact)
+			node_->addFact(values_);
+		node_->addDomain(values_);
+	}
 }
 
 void Atom::print_plain(Output *o, std::ostream &out)

### 2008-07-07 lib/gringo/include/grounder.h
- updated todolist (there is a small bug ill fix later) - added an example for negative heads - changed the statementdependencygraph/grounder slightly
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@378 3e02f596-3131-0410-8352-e04eeaec37cc

@@ -28,7 +28,7 @@ namespace NS_GRINGO
 	{
 	protected:
 		typedef std::pair<std::string*, StringVector*> DomainPredicate;
-		typedef std::vector<DomainPredicate> DomainVector;
+		typedef std::vector<DomainPredicate> DomainPredicateVector;
 		typedef std::map<std::string*, int> VariableMap;
 		typedef std::map<std::string*, std::pair<bool, Term*> > ConstTerms;
 	public:
@@ -38,7 +38,7 @@ namespace NS_GRINGO
 		void buildDepGraph();
 		void start();
 		void addDomains();
-		void reset(bool warn);
+		void reset();
 		bool check();
 		void ground();
 		void addProgram(Program *scc);
@@ -58,6 +58,8 @@ namespace NS_GRINGO
 		void preprocess();
 		NS_OUTPUT::Output *getOutput();
 		Evaluator *getEvaluator();
+		/// Adds a domain that never occurs in any head
+		void addZeroDomain(Domain *d);
 		virtual ~Grounder();
 	protected:
 		void addDomains(std::string *id, std::vector<StringVector*>::iterator pos, std::vector<StringVector*>::iterator end, StringVector &list);
@@ -65,7 +67,7 @@ namespace NS_GRINGO
 		int internalVars_;
 		ProgramVector sccs_;
 		VariableMap varMap_;
-		DomainVector domains_;
+		DomainPredicateVector domains_;
 		StatementVector rules_;
 		NS_OUTPUT::Output *output_;
 		Evaluator *eval_;

### 2008-07-07 lib/gringo/src/aggregateliteral.cpp

@@ -127,8 +127,9 @@ SDGNode *AggregateLiteral::createNode(SDG *dg, SDGNode *prev, DependencyAdd todo
 		SDGNode *p = (*it)->createNode(dg, prev, ADD_NOTHING);
 		assert(p);
 		prev->addDependency(p, true);
-		// if used in the head they it also depends cyclically on its literals
-		if(todo == ADD_HEAD_DEP)
+		// if used in the head it also depends cyclically on its literals
+		// but not if the literal is negative
+		if(todo == ADD_HEAD_DEP && !(*it)->getNeg())
 			p->addDependency(prev);
 	}
 	return 0;

### 2008-07-07 lib/gringo/src/grounder.cpp

@@ -68,7 +68,7 @@ void Grounder::buildDepGraph()
 	SDG dg;
 	for(StatementVector::iterator it = rules_.begin(); it != rules_.end(); it++)
 		(*it)->buildDepGraph(&dg);
-	reset(true);
+	reset();
 	dg.calcSCCs(this);
 }
 
@@ -87,7 +87,7 @@ void Grounder::addDomains()
 		Statement *rule = *it;
 		VarSet vars;
 		rule->getVars(vars);
-		for(DomainVector::iterator it = domains_.begin(); it != domains_.end(); it++)
+		for(DomainPredicateVector::iterator it = domains_.begin(); it != domains_.end(); it++)
 		{
 			DomainPredicate &dp = *it;
 			for(StringVector::iterator it = dp.second->begin(); it != dp.second->end(); it++)
@@ -107,24 +107,24 @@ void Grounder::addDomains()
 	}
 }
 
-void Grounder::reset(bool warn = false)
+void Grounder::addZeroDomain(Domain *d)
+{
+	d->setSolved(true);
+	for(int uid = 0; uid < (int)getDomains()->size(); uid++)
+	{
+		if(d == (*getDomains())[uid])
+		{
+			Signature &sig = (*getPred())[uid];
+			std::cerr << "Warning: " << *sig.first << "/" << sig.second << " is never defined." << std::endl;
+			break;
+		}
+	}
+}
+
+void Grounder::reset()
 {
 	for(StatementVector::iterator it = rules_.begin(); it != rules_.end(); it++)
 		(*it)->reset();
-	// all nodes with zero defines have to be false
-	for(int uid = 0; uid < (int)getDomains()->size(); uid++)
-	{
-		Domain *n = (*getDomains())[uid];
-		if(n->complete())
-		{
-			if(warn)
-			{
-				Signature &sig = (*getPred())[uid];
-				std::cerr << "Warning: " << *sig.first << "/" << sig.second << " is never defined." << std::endl;
-			}
-			n->setSolved(true);
-		}
-	}
 }
 
 void Grounder::preprocess()
@@ -150,7 +150,7 @@ void Grounder::start()
 	std::cerr << "checking ... " << std::endl;
 	if(!check())
 		throw GrinGoException("Error: the program is not groundable.");
-	reset(false);
+	reset();
 	std::cerr << "done" << std::endl;
 	std::cerr << "grounding ... " << std::endl;
 	ground();
[…]

### 2008-07-07 lib/gringo/src/statementdependencygraph.cpp

@@ -169,6 +169,8 @@ void SDG::calcSCCs(Grounder *g)
 		}
 		// set the type of the domain
 		v->getDomain()->setType(static_cast<Domain::Type>(v->scc_->type_));
+		if(v->dependency_.size() == 0)
+			g->addZeroDomain(v->getDomain());
 	}
 
 	// do a topological sort

### 2008-07-21 doc/documentation/paper.tex
dumb changes
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@376 3e02f596-3131-0410-8352-e04eeaec37cc
example and explaination for writing logic programs
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@406 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 The compute statement is solver dependent and supported by every solver.
+
+\subsection{How to write a logic program}
+Writing a logic program to solve a problem is rather describing a problem than describing an algorithm to solve a problem.
+Therefore the problem must be encoded as a set of rules.
+Typically these are divided into a problem description and an instance of the problem.
+The solution to a problem can then be read from answer sets of the rules.
+The procedure of writing a logic program shall now be explained with a little example.
+We want to write a problem encoding for the 3-colorability problem of graph coloring.
+Given an arbitraty graph, no two connected edges may have the same color.
+\newline
+Usually you start with the non instance domains, like what colors we do have.
+\begin{verbatim}
+ color(red). color(green). color(blue).
+\end{verbatim}
+After that we have to color the vertices, but we must give attention that each vertex only has one color.
+We can do this by simply saying that a vertex is colored with one color X, and that the same vertex does not have another color Y.
+\begin{verbatim}
+ colored(V,C) :- not colored(V,D), C != D, color(C;D), vertex(V).
+\end{verbatim}
+As you can see, you have to give the domain of the variables in the rule.
+The variable V is used for vertices and the variables C and D are used for colors.
+After every vertex now has a color we would get all possible solutions of all vertices colorfully painted.
+We now need to restrict our solutions again by saying that two vertices that are connected by an edge should not have the same color.
+We can do this by using an integrity constraint.
+An integrity constraint is a rule with an empty head, and if an assignment of variables fulfills the body, the assignment is not a solution of our problem.
+\begin{verbatim}
+ :- edge(V,U), colored(V,C), colored(U,C), vertex(V;U), color(C).
+\end{verbatim}
+Here we have the edge between two vertices V and U, and if both are colored with the same color C the assignment is not a solution to our problem.
+\newline
+We have finally written our problem description, all we need now is a problem instance. This step is fairly simple.
+We just need to define several vertices and edges that do connect the vertices.
+\begin{verbatim}
+ vertices(1..4).
+ edge(1,2). edge(1,4). edge(2,4).
+ edge(3,1). edge(4,1). edge(3,2).
+\end{verbatim}
+
+If we give the problem description and instance to the grounder, it will generate a variable free version of the problem.
+This can then be solved by a solver and in the answer sets of the solver the colored predicate states the color of a valid coloring of all vertices.
+\newline
+To ease the process writing logic programs some features in the input syntax were introduced and are now explained in detail.
 \subsection{Features}
@@
 
+\paragraph{Expander :}
+With the expander ``:'' a domain for an atom can be set.
+On the left side of the operator an atom with variables is expected and on the right side the domain for that variables is given.
+In the head, the atoms get connected via a disjunction, in the body they get connected via a conjunction, so no new rules are created with this operator.
+\begin{verbatim}
+ a(X,Y) : b(X) : c(Y) :- d(X) : e(X).
+ b(1..2).
+ c(3).
+ e(a;b;c).
+\end{verbatim}
+would become
+\begin{verbatim}
+ a(1,3) | a(2,3) :- d(a), d(b), d(c).
+\end{verbatim}
+This operator can be used in aggregates to create a comma separated list of atoms.
+
 \paragraph{Conjunctor ;}
@@
 \end{verbatim}
-
-\paragraph{Expander :}
-With the expander ``:'' a domain for an atom can be set.
-On the left side of the operator an atom with variables is expected and on the right side the domain for that variables is given.
-In the head, the atoms get connected via a disjunction, in the body they get connected via a conjunction, so no new rules are created with this operator.
-\begin{verbatim}
- a(X,Y) : b(X) : c(Y) :- d(X) : e(X).
- b(1..2).
- c(3).
- e(a;b;c).
-\end{verbatim}
-would become
-\begin{verbatim}
- a(1,3) | a(2,3) :- d(a), d(b), d(c).
-\end{verbatim}
-This operator can be used in aggregates to create a comma separated list of atoms.
 
@@
 
-\section{$\Pi$-Restrictiveness (or something else)}
+\section{$\lambda$-Restriction}
 \label{domainrestricted}
+Gringo verifies that the input program is $\lambda$-
+restricted \cite{gringopaper}, so that the existence of a finite equivalent ground instantiation is guaranteed.
+The definition of lambda restrictedness can be found in \cite{gringopaper} and was expanded for aggregates and disjunctions.
+
 \section{Examples}

### 2008-07-21 doc/documentation/paper.tex
smaller changes
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@407 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 Gringo also has an internal interface to the solver clasp \cite{gekanesc07b}, so the output can directly by parsed to the solver and answersets can be computed.
-The syntax of gringo2.0 can be found in the appendix \ref{appendix}.
+Furthermore gringo can be used with clasp incrementally. Detailed information about incremental solving can be found here \cite{IncrementalASP} 
+The syntax of the gringo2.0 input language can be found in the appendix \ref{appendix}.
+
+\section{$\lambda$-Restriction}
+\label{domainrestricted}
+Gringo verifies that the input program is $\lambda$-
+restricted \cite{gringopaper}, so that the existence of a finite equivalent ground instantiation is guaranteed.
+The definition of lambda restrictedness can be found in \cite{gringopaper} and was expanded for aggregates and disjunctions.
 
@@
 \end{verbatim}
-Furthermore all restriction on smodels output al hold as clasps internal interface currently reads smodels format.
+Furthermore all restriction on smodels output hold as clasps internal interface currently reads smodels format.
 \paragraph{-p}
@@
 \newline
-Usually you start with the non instance domains, like what colors we do have.
+Usually you start with the non instance domains, in our example the colors we use.
 \begin{verbatim}
@@
 \end{verbatim}
-After that we have to color the vertices, but we must give attention that each vertex only has one color.
+After that, we have to color the vertices, but we must give attention that each vertex has only one color.
 We can do this by simply saying that a vertex is colored with one color X, and that the same vertex does not have another color Y.
@@
 If we give the problem description and instance to the grounder, it will generate a variable free version of the problem.
-This can then be solved by a solver and in the answer sets of the solver the colored predicate states the color of a valid coloring of all vertices.
+This can then be solved by a solver and in the answer sets of the solver the ``colored'' predicate states a valid coloring of all vertices.
 \newline
@@
 \subsection{Features}
+The extended features of the input syntax are now explained in detail. For a full syntax description in EBNF see appendix (\ref{appendix}).
 \paragraph{Expanding Terms}
@@
 
-\section{$\lambda$-Restriction}
-\label{domainrestricted}
-Gringo verifies that the input program is $\lambda$-
-restricted \cite{gringopaper}, so that the existence of a finite equivalent ground instantiation is guaranteed.
-The definition of lambda restrictedness can be found in \cite{gringopaper} and was expanded for aggregates and disjunctions.
+\section{Compiling Gringo}
+For compiling gringo, a binary version of ``re2c'' is needed to compile the lexer and the parser.
+TODO: What about lemon.
+As the build environment ``cmake'' is used.
+TODO: How to build gringo, clingo, iclingo etc...
 
@@
 \begin{appendix}
+\section{EBNF Syntax}
 \label{appendix}

### 2008-07-28 lib/gringo/src/claspoutput.cpp
- set the empty name for the auxillary atom used during incremental grounding   this avoids some ugly bugs
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@410 3e02f596-3131-0410-8352-e04eeaec37cc

@@ -175,6 +175,7 @@ void IClaspOutput::reinitialize()
 		b_->setCompute(incUid_, false);
 	// create a new uid
 	incUid_ = newUid();
+	b_->setAtomName(incUid_, "");
 	IntVector empty, head;
 	head.push_back(incUid_);
 	printChoiceRule(head, empty, empty);

### 2008-07-28 doc/documentation/paper.tex
added some of svens remarks to the documentation
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@411 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 \end{figure}
-Gringo2.0 is a state of the art and has many features that makes writing logic programs easy.
+gringo2.0 is a state of the art grounder and has many features that makes writing logic programs easy.
 It was developed by Roland Kaminski and is orientated on the ideas of gringo1.0.
@@
 A broad set of potentialities for writing encodings for logic programs exists.
-Currently, no other grounder support such a rich featured, less restricted input language.
-There exist several output formats which can be used be different grounders and even a human readable format for debugging purposes.
-Gringo also has an internal interface to the solver clasp \cite{gekanesc07b}, so the output can directly by parsed to the solver and answersets can be computed.
+Currently, no other grounder supports such a rich featured, less restricted input language.
+gringo2.0 supports several output formats and even a human readable format for debugging purposes.
+It also has an internal interface to the solver clasp \cite{gekanesc07b}, so the output can directly by parsed to the solver and answersets can be computed.
 Furthermore gringo can be used with clasp incrementally. Detailed information about incremental solving can be found here \cite{IncrementalASP} 
@@
 Gringo can be started on the command line and has some parameters that can be useful.
-The first parameters are gringo parameters and and arbritary number of filenames of the logic program.
+The first parameters are gringo specific parameters and any arbritary number of filenames of the logic program.
 \paragraph{--help}
@@
 \paragraph{--const}
-You can set the value of constants defined in you logic program via the command line of gringo.
+You can set the value of constants defined in your logic program via the command line of gringo.
 After the parameter ``--const'' you have to give the name of the constant you want to set followed by a ``='' and the value of that constant.
@@
 Parameter ``p'' invokes the plain output method, so the output is printed in human readable format, which is also the input format for gringo.
-\paragraph{-g X}
+\paragraph{-g [1-7]}
 This parameter invokes the new output format for the intermediate languages ASPils \cite{atleastthisreferenceishouldgive}.
-The X can be any number from $1$ to $7$ to refer to one of the following normalforms (Simple=1, SModels=2, DisjunctiveLogicProgram=3, CModels=4, CModelsExtended=5, DLV=6, Conglomeration=7).
+The optional number can range from $1$ to $7$ and refers to one of the following normalforms (Simple=1, SModels=2, DisjunctiveLogicProgram=3, CModels=4, CModelsExtended=5, DLV=6, Conglomeration=7).
 If no normalform is specified, conglomeration(7)  is assumed.
@@
 The parameters are either constants, variables or function symbols.
-A function symbols do have the same syntax as atoms. This way they can be finitely nested.
+Function symbols do have the same syntax as atoms. This way they can be finitely nested.
 As an example, an atom with three parameters is given:
-The first parameter being a constant, the others being function symbols.
+The first parameter being a constant $a$, the others being function symbols $f(b,X)$ and $g(f(X), Y))$.
 \begin{verbatim}
@@
 \paragraph{Rule}
-A rule normally consists of a head and a body and ands with a dot.
-The body is separated from the using ``\:\-'' as in
+A rule normally consists of a head and a body and ends with a dot.
+The body is separated from the head using ``:-'' as in
 \begin{verbatim}
@@
 \end{verbatim}
-This warning occurs if a sum aggregate with negative weights has been found and the output format smodels is used or the output is redirected to clasp.
+This warning occurs if a sum aggregate with negative weights has been found and the output format smodels is used or the output is redirected to clasp. In this format negative bounds are not allowed.
  \item 
@@
 This warnings states that an atom is used in the body of a rule that never occurs in the head.
-``p'' is the name of the atom and ``3'' is the arity.
+``p'' is the name of the predicate and ``3'' is the arity.
 \item
@@
 \paragraph{Errors}
-Gringo2.0 stops the execution and prints an error mesaage to standard error output.
+gringo2.0 stops the execution and prints an error message to standard error output.
 \begin{itemize}
@@
 \end{verbatim}
-This error message gives a list of atoms that are not omega restricted.
+This error message gives a list of predicates that are not domain restricted.
  \item 
@@
  "the following rule cannot not be grounded,
-  weakly restricted varibles: { ... }"
+  weakly restricted variables: { ... }"
 \end{verbatim}
-This error message gives a list of atoms that are not lambda restricted. 
+This error message gives a list of predicates that are not lambda restricted. 
 \item
@@
 The arity of the predicate can be given explicit or implicit with a list of variables.
-The variables are just placeholder and do not unify. The second syntax is implemented for compatibility reasons only.
-If the predicate that has to be hidden has arity zero, meaning it is an atom, the syntax is
+The variables are just placeholder and do not unify with other variables. For example hide p(X,X) would also hide the fact p(a,b). The second syntax is implemented for compatibility reasons only.
+If the predicate that has to be hidden has arity zero the syntax is
 \begin{verbatim}
@@
 \paragraph{const}
-Within the ``const'' statement, a default value for constants can be defined.
-Constants are atoms used in the logic program and are normally defined via the command line options.
+Within the ``const'' statement, a default value for constants can be defined in the logic program.
+while they are normally defined via the command line options.
 A default value for a constant $n$ can be easily be set.
@@
 makes sure that the variable ``T'' is always bound to the domain of ``time''.
-This has the same effect as adding the domain ``time(T)'' to every rule the variable occurs in.
+This has the same effect as adding the domain ``time(T)'' to every rule the variable ``T'' occurs in.
 \paragraph{compute}
@@
 Typically these are divided into a problem description and an instance of the problem.
-The solution to a problem can then be read from answer sets of the rules.
+The solution to a problem can then be read from the answer sets of the program.
 The procedure of writing a logic program shall now be explained with a little example.
@@
 \subparagraph{Range Operator}
-To define several atoms that stretch on one or more parameters over a certain range, two dots can used to express an increasing integer range.
+To define several atoms that stretch on one or more parameters over a certain range, two dots can be used to express an increasing integer range.
 Using the range operator in the head of a rule, it works as a conjunction of all the members in the range, so for every number the rule is duplicated.
@@
 \end{verbatim}
-will result in the rules
+will result in the rules\footnote{gringo would simplify the grounded rules to facts, since all literals in the body are facts}
 \begin{verbatim}
@@
 \paragraph{Arithmetic Expressions}
-The build in arithmetics can be used for integer arithmetic.
+The build-in arithmetics can be used for integer arithmetic.
 As operators addition (+), subtraction (-), multiplication (*), division (/), modulo (mod) and the absolute value operator (abs) can be used.
@@
 \end{verbatim}
-would result in the grounded program\footnote{gringo would simplify the grounded rules to facts, since all literals in the body are facts}
+would result in the grounded program
 \begin{verbatim}
@@
 They can be used to constrain certain variables.
-For example, some rule is only generated for all unifications where X is greater than Y-1.
+For example, the following rule is only generated for all unifications where X is greater than Y-1.
 \begin{verbatim}
@@
 
-\paragraph{True Negation}
+\paragraph{Classical Negation}
 True negation can be used. Therefore the atom is preceded by a minus sign.

### 2008-07-30 doc/documentation/paper.tex
changes according to sven, unfinished
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@413 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 A broad set of potentialities for writing encodings for logic programs exists.
-Currently, no other grounder supports such a rich featured, less restricted input language.
+Currently, no other grounder supports such a rich featured language.
 gringo2.0 supports several output formats and even a human readable format for debugging purposes.
@@
 \label{domainrestricted}
+There exist two major grounders for ASP besides gringo, lParse \cite{lparseManual} and  the DLV grounding module\cite{DLV}.
+To ensure the existence of an equivivalent ground instantiation of a program both grounders except a subclass of logic programs.
+For lParse the logic programs must be $\omega$-restricted\cite{omega} while DLV can handle safe programs\cite{safe}.
 Gringo verifies that the input program is $\lambda$-
 restricted \cite{gringopaper}, so that the existence of a finite equivalent ground instantiation is guaranteed. 
-The definition of lambda restrictedness can be found in \cite{gringopaper} and was expanded for aggregates and disjunctions.
+$\lambda$-restrictedness extends the capabilities of $\omega$-restricted programs.
+The definition of $\lambda$-restrictedness can be found in \cite{gringopaper} and was expanded for aggregates and disjunctions.
+$\lambda$-restrictedness can not be compared with the safe programs of DLV since there exist programs that are safe and not $\lambda$-restricted and vice versa.
 
@@
 Gringo can be started on the command line and has some parameters that can be useful.
-The first parameters are gringo specific parameters and any arbritary number of filenames of the logic program.
-\paragraph{--help}
+The first parameters are gringo specific parameters and consist of a number of options and any arbritary number of filenames of the logic program.
+For the special purpose that Gringo is used with its clasp module a double minus ``\texttt{--}'' is used to seperate the clasp parameters from the Gringo parameters.
+\paragraph{\texttt{--help}}
 Gives a short overview of all parameters.
-\paragraph{--const}
+\paragraph{\texttt{--const}}
 You can set the value of constants defined in your logic program via the command line of gringo.
@@
 
-\paragraph{--convert}
+\paragraph{\texttt{--convert}}
 If this parameter is used, the grounder expects an already grounded program as input.
 For converting an already grounded logic program into an output format supported by gringo this is the right option. This mode saves a lot of memory but does not ground the program.
-\paragraph{-l}
+\paragraph{\texttt{-l}}
 This mode prints smodels conform output. For a definition of the output format see \cite{lparseManual}.
@@
 So a transformation as explaned in \cite{lparseManual} is used to get rid of the negative weights.
-\paragraph{-c}
+\paragraph{\texttt{-c}}
 In this mode, the output is directly parsed to clasp, an ASP solver. For solving logic programs, no intermediate files are produced and the answersets are printed directly.
 With this option enabled, all parameters for clasp are valid and can be declared in the command line.
-Therefore adouble minus ``--'' must be given after all gringo related parameters are set.
+Therefore adouble minus ``\texttt{--}'' must be given after all gringo related parameters are set.
 After them die clasp parameters follow.
@@
 Furthermore all restriction on smodels output hold as clasps internal interface currently reads smodels format.
-\paragraph{-p}
-Parameter ``p'' invokes the plain output method, so the output is printed in human readable format, which is also the input format for gringo.
-\paragraph{-g [1-7]}
+\paragraph{\texttt{-p}}
+Parameter ``\texttt{-p}'' invokes the plain output method, so the output is printed in human readable format, which is also the input format for gringo.
+\paragraph{\texttt{-g [1-7]}}
 This parameter invokes the new output format for the intermediate languages ASPils \cite{atleastthisreferenceishouldgive}.
@@
 \paragraph{default parameters}
-The default parameter is ``-l'' for smodels output.
-\section{Language Constructs}
-The basic elements of a logic program are atoms, constants, variables and rules.
-\paragraph{Constant}
-A constant is a representation of an entity in the problem domain.
-A constant is either a name, starting with a lower case letter or an underscore, or a positive integer.
-It also can be encompassed by double quotation marks.
-When using quotation marks, which is sometimes helpful when using constants with spaces, it has to be remarked that a constant with quotation marks is never equal to a constant without, so ``$a$'' is not $a$.
-\paragraph{Variable}
-A variable is programming construct to simplify writing logic programs.
-It always starts with a capitalized letter and can contain digits and letters.
-In the grounding process a variable is replaced by one or more constants.
-Each variable occurring in a rule must be domain restricted as described in section \ref{domainrestricted}.
-\paragraph{Atom}
+The default parameter is ``\texttt{-l}'' for smodels output.
+\section{Syntax}
+The basic elements of a logic program and its syntax is now described.
+Disjunctive logic programs are based on an alphabet \cite{syntax} consisting of:
+\begin{itemize}
+\item a set P of predicate symbols,
+\item a set F of function symbols,
+\item a set V of variables,
+\item a set C of constants,
+\item a set \{ ``$\mid$'', ``,'' ,$\leftarrow$, not \} of connectives, and
+\item a set \{ ``('', ``)'', ``,'' \} of punctuation symbols.
+\end{itemize}
 
-An atom is a predicate with a name. The name can consist of letters, digits and the underscore.
-It has to start with a lower case letter or an underscore.
-An atom can have several parameters, encapsulated by parenthesis and separated with commas.
-The parameters are either constants, variables or function symbols.
-Function symbols do have the same syntax as atoms. This way they can be finitely nested.
-As an example, an atom with three parameters is given:
-The first parameter being a constant $a$, the others being function symbols $f(b,X)$ and $g(f(X), Y))$.
+TODO: Which greek letters for P, F, L and C
+\paragraph{Predicates and function symbols}
+Predicates and function symbols have a predicate symbol/name and an arity.
+For the function symbol $f$ with arity 3 we write $f/3$.
+The names of the function symbols and predicates have to start with either a lower case letter or an underscore and can contain digits and letters.
+With $P$ we denote to the set of predicates and $F$ denotes the set of function symbols.
+\paragraph{Constants}
+Constants are function symbols with an arity of zero.
+The name of a constant can also be encompassed by double quotation marks, which is sometimes helpful when using constants with spaces.
+Tt has to be remarked that a constant with quotation marks is never equal to a constant without, so ``$a$'' is not equal to $a$.
+$C$ is the set of constants
+\paragraph{Variables}
+Variables have to start with a capitalized letter and can contain digits and letters.
+The set of variables is denoted by $V$.
+The language over $P$, $F$and $C$ is denoted by $L(P,F,C)$.
+\paragraph{Terms}
+The building blocks of the language are terms.
+Each constant $c \in C$ and each variable $f \in F$ is a term.
+Furthermore $f(t_1, \dots, t_n)$ is a term if $f/n \in F$ and $t_i \mid i \in \{1 \dots n\}$ are terms.
+Since a term is defined inductively, it can be nested.
+A term with the first parameter being a constant $a$, the others being terms $f(b,X)$ and $g(f(X), Y))$ would be expressed this way.
 \begin{verbatim}
@@
 \end{verbatim}
+\paragraph{Atoms}
+An atom over $L(P, F, C)$ is of the form $p(t1 , . . . , tn )$ where $p/n$ is a predicate
+symbol with arity $n$, and $t_i \mid i \in \{1 \dots n\}$ are terms.
+\paragraph{Rules}
+A rule over the language $L(P,F,C)$ has the form
+$$
+A_0 \mid \dots \mid A_l \leftarrow A_{l+1}, \dots, A_m, not\ A_{m+1}, \dots, not\ A_n.
+$$
+where $A_0$ to $A_n$ are atoms. \newline
+A normal logic program $\Pi$ over $L(P,F,C)$ is a finite set of rules over $L(P,F,C)$.
 
-\paragraph{Rule}
-A rule normally consists of a head and a body and ends with a dot.
-The body is separated from the head using ``:-'' as in
-\begin{verbatim}
- head :- body.
-\end{verbatim}
-The body is a comma separated conjunction of expressions.
-The supported expressions are introduced in the next section.
-The variables used in the head must be restricted to a domain in the body.
-Check (\ref{domainrestricted}) to see the restrictions on domains.
-\newline
-A rule can occur without a body as a fact, therefore the body is omitted.
-\begin{verbatim}
- head.
-\end{verbatim}
-The same can be done with the body to express an integrity constraint.
-\begin{verbatim}
- :- body.
-\end{verbatim}
+%\paragraph{Function Symbols}
+%A constant is a representation of an entity in the problem domain.
+%A constant is either a name, starting with a lower case letter or an underscore, or a positive integer.
+%It also can be encompassed by double quotation marks.
+When using quotation marks, which is sometimes helpful when using %constants with spaces, it has to be remarked that a constant with quotation marks is never equal to a constant without, so ``$a$'' is not $a$.
+%\paragraph{Variable}
+%A variable is programming construct to simplify writing logic programs.
+%It always starts with a capitalized letter and can contain digits and letters.
+%In the grounding process a variable is replaced by one or more constants.
+%Each variable occurring in a rule must be domain restricted as described in section \ref{domainrestricted}.
+%\paragraph{Atom}
+%An atom is a predicate with a name. The name can consist of letters, digits and the underscore.
+%It has to start with a lower case letter or an underscore.
+%An atom can have several parameters, encapsulated by parenthesis and separated with commas.
+%The parameters are either constants, variables or function symbols.
+%Function symbols do have the same syntax as atoms. This way they can be finitely nested.
+%As an example, an atom with three parameters is given:
+%The first parameter being a constant $a$, the others being function symbols $f(b,X)$ and $g(f(X), Y))$.
+%\begin{verbatim}
+% p(a,f(b,X), g(f(X),Y)).
+%\end{verbatim}
 
-\paragraph{Comments}
-Comments included in the input are line based.
-Every line starting with ``\%'' is a comment. The whole line is omitted and not to be handled by the grounder.
-\paragraph{Warnings}
-Several warnings are produced by gringo to give hints on common mistakes.
-They are all printed to standard error output.
+%\paragraph{Rule}
+%A rule normally consists of a head and a body and ends with a dot.
+%The body is separated from the head using ``:-'' as in
+%\begin{verbatim}
+% head :- body.
+%\end{verbatim}
+%The body is a comma separated conjunction of expressions.
+%The supported expressions are introduced in the next section.
+%The variables used in the head must be restricted to a domain in the body.
+%Check (\ref{domainrestricted}) to see the restrictions on domains.
+%\newline
+%A rule can occur without a body as a fact, therefore the body is omitted.
+%\begin{verbatim}
+% head.
+%\end{verbatim}
+%The same can be done with the body to express an integrity constraint.
+%\begin{verbatim}
+% :- body.
+%\end{verbatim}
 
-\begin{itemize}
- \item 
-\begin{verbatim}
- "Warning: sum with negative bounds in the body of a rule"
-\end{verbatim}
-This warning occurs if a sum aggregate with negative weights has been found and the output format smodels is used or the output is redirected to clasp. In this format negative bounds are not allowed.
- \item 
-\begin{verbatim}
- "Warning: multiple definitions of #const n"
-\end{verbatim}
-This warning is printed if more than one \#const statement is used to set the same constant.
-The last attached value is used for grounding.
-\item
-\begin{verbatim}
- "Warning: p/3 is never defined"
-\end{verbatim}
-This warnings states that an atom is used in the body of a rule that never occurs in the head.
-``p'' is the name of the predicate and ``3'' is the arity.
-\item
-\begin{verbatim}
- "got ground program i hope u have enough memory :)"
-\end{verbatim}
-Use the command line option ``--convert'' to convert a ground program into another format.
-
-\end{itemize}
-Furthermore all warnings from clasp are included.
-
-\paragraph{Errors}
-gringo2.0 stops the execution and prints an error message to standard error output.
-\begin{itemize}
- \item 
-\begin{verbatim}
- "the following rule cannot not be grounded,
-  non domain predicates : { ... }"
-\end{verbatim}
-This error message gives a list of predicates that are not domain restricted.
- \item 
-\begin{verbatim}
- "the following rule cannot not be grounded,
-  weakly restricted variables: { ... }"
-\end{verbatim}
-This error message gives a list of predicates that are not lambda restricted. 
-\item
-\begin{verbatim}
- "syntax error on line Y column Z unexpected token: X"
-\end{verbatim}
-The parser can not read the input format.
-\begin{verbatim}
- "Disjunction not allowed on this normalform."
- "Count aggregate/cardinality constraint not allowed
-  in this normalform."
- "Count aggregate/cardinality constraint not allowed 
-  with non trivial upper bound in this normalform."
- "Count aggregate/cardinality constraint not allowed 
-  with negative lower bound in this normalform."
- "Sum aggregate/weight constraint not allowed
-  in this normalform."
- "Sum aggregate/weight constraint not allowed 
-  with non trivial upper bound in this normalform."
- "Sum aggregate/weight constraint not allowed 
-  with negative lower bound in this normalform."
- "Sum aggregate/weight constraint not allowed
-  with negative weights in this normalform."
- "Sum aggregate/weight constraint not allowed
-  with negative weights in this normalform."
- "Max aggregate not allowed in this normalform."
- "Max aggregate not allowed
-  with negative weights in this normalform."
- "Min aggregate not allowed in this normalform."
- "Min aggregate not allowed 
-  with negative weights in this normalform."
- "Optimize statement not allowed in this normalform."
-\end{verbatim}
-%\begin{verbatim}
-% "Times aggregate not allowed in this normalform."
-% "Times aggregate not allowed
-%  with negative weights in this normalform."
-%\end{verbatim}
-These errors belong to the ASPils output format. Another normal form as to be chosen.
-\end{itemize}
-
-\subsection{Meta Constructs}
+\subsection{Meta-Syntax}
 The ``\#'' sign in front of all meta expressions is optional and can therefore be omitted.
@@
 The compute statement is solver dependent and supported by every solver.
+\paragraph{Comments}
+Comments included in the input are line based.
+Every line starting with ``\%'' is a comment. The whole line is omitted and not to be handled by the grounder.
 
@@
 The extended features of the input syntax are now explained in detail. For a full syntax description in EBNF see appendix (\ref{appendix}).
+\paragraph{Function Symbols}
+Function symbols are fully supported with Gringo2.0.
+They can be finitly nested and also be used as domain predicates.
+While in lParse function symbols are allowed in a very restricted way, no restrictions apply in Gringo2.0.
+The following program can not be grounded with lParse since $X$ is an unrestricted variable. With Gringo2.0 the program is now groundable.
+\begin{verbatim}
+a(X) :- f(1,g(X)).
+f(1,X) :- f(X).
+f(g(2)).
+\end{verbatim}
+From the second rule the domain \texttt{f(1,g(2))} is generated.
+This domain can be used for the first rule and results in the grounded rule
+\begin{verbatim}
+ a(2) :- f(1,g(2)).
+\end{verbatim}
+
 \paragraph{Expanding Terms}
@@
 \paragraph{Arithmetic Expressions}
-The build-in arithmetics can be used for integer arithmetic.
+The built-in arithmetics can be used for integer arithmetic.
 As operators addition (+), subtraction (-), multiplication (*), division (/), modulo (mod) and the absolute value operator (abs) can be used.
@@
 
+\subsection{Warnings}
+Several warnings are produced by gringo to give hints on common mistakes.
+They are all printed to standard error output.
+
+\begin{itemize}
+ \item 
+\begin{verbatim}
+ "Warning: sum with negative bounds in the body of a rule"
+\end{verbatim}
+This warning occurs if a sum aggregate with negative weights has been found and the output format smodels is used or the output is redirected to clasp. In this format negative bounds are not allowed.
+ \item 
+\begin{verbatim}
+ "Warning: multiple definitions of #const n"
+\end{verbatim}
+This warning is printed if more than one \#const statement is used to set the same constant.
+The last attached value is used for grounding.
+\item
+\begin{verbatim}
+ "Warning: p/3 is never defined"
+\end{verbatim}
+This warnings states that an atom is used in the body of a rule that never occurs in the head.
+``p'' is the name of the predicate and ``3'' is the arity.
+\item
+\begin{verbatim}
+ "got ground program i hope u have enough memory :)"
+\end{verbatim}
+Use the command line option ``--convert'' to convert a ground program into another format.
+
+\end{itemize}
+Furthermore all warnings from clasp are included.
+
+\subsection{Errors}
+gringo2.0 stops the execution and prints an error message to standard error output.
+\begin{itemize}
+ \item 
+\begin{verbatim}
+ "the following rule cannot not be grounded,
+  non domain predicates : { ... }"
+\end{verbatim}
+This error message gives a list of predicates that are not domain restricted.
+ \item 
+\begin{verbatim}
+ "the following rule cannot not be grounded,
+  weakly restricted variables: { ... }"
+\end{verbatim}
+This error message gives a list of predicates that are not lambda restricted. 
+\item
+\begin{verbatim}
+ "syntax error on line Y column Z unexpected token: X"
+\end{verbatim}
+The parser can not read the input format.
+\begin{verbatim}
+ "Disjunction not allowed on this normalform."
+ "Count aggregate/cardinality constraint not allowed
+  in this normalform."
+ "Count aggregate/cardinality constraint not allowed 
+  with non trivial upper bound in this normalform."
+ "Count aggregate/cardinality constraint not allowed 
+  with negative lower bound in this normalform."
+ "Sum aggregate/weight constraint not allowed
+  in this normalform."
+ "Sum aggregate/weight constraint not allowed 
+  with non trivial upper bound in this normalform."
+ "Sum aggregate/weight constraint not allowed 
+  with negative lower bound in this normalform."
+ "Sum aggregate/weight constraint not allowed
+  with negative weights in this normalform."
+ "Sum aggregate/weight constraint not allowed
+  with negative weights in this normalform."
+ "Max aggregate not allowed in this normalform."
+ "Max aggregate not allowed
+  with negative weights in this normalform."
+ "Min aggregate not allowed in this normalform."
+ "Min aggregate not allowed 
+  with negative weights in this normalform."
+ "Optimize statement not allowed in this normalform."
+\end{verbatim}
+%\begin{verbatim}
+% "Times aggregate not allowed in this normalform."
+% "Times aggregate not allowed
+%  with negative weights in this normalform."
+%\end{verbatim}
+These errors belong to the ASPils output format. Another normal form as to be chosen.
+\end{itemize}
+
 \section{Examples}
@@
 \input{benchmark2.result}
-\appendixname
+\section{Future Work}
+$\Pi$-restrictedness
+bottom up grounding
+\newpage
+\textbf{\appendixname}
 \begin{appendix}

### 2008-07-30 doc/documentation/paper.tex
Further bugfixes in the documentation, changed some error messages and one typo
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@414 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 This parameter invokes the new output format for the intermediate languages ASPils \cite{atleastthisreferenceishouldgive}.
-The optional number can range from $1$ to $7$ and refers to one of the following normalforms (Simple=1, SModels=2, DisjunctiveLogicProgram=3, CModels=4, CModelsExtended=5, DLV=6, Conglomeration=7).
-If no normalform is specified, conglomeration(7)  is assumed.
+The optional number can range from $1$ to $7$ and refers to one of the following normal forms (Simple=1, SModels=2, DisjunctiveLogicProgram=3, CModels=4, CModelsExtended=5, DLV=6, Conglomeration=7).
+If no normal form is specified, conglomeration(7)  is assumed.
 
@@
 \subsection{Meta-Syntax}
-The ``\#'' sign in front of all meta expressions is optional and can therefore be omitted.
+The ``\#'' sign in front of all meta expressions is optional but is it strongly recommend to use it.
+It separates the meta directives from the logic program more clearly and can prevent severe mistakes,
+ confusing those constructs with atoms as shown in the example.
+\begin{verbatim}
+ hide.
+ hidden :- hide.
+\end{verbatim}
+This program hides all predicates from the output of the ASP solver instead of deriving the facts \texttt{hide} and \texttt{hidden}.
+
 \paragraph{hide}
@@
 \end{verbatim}
-This warning occurs if a sum aggregate with negative weights has been found and the output format smodels is used or the output is redirected to clasp. In this format negative bounds are not allowed.
+This warning occurs if a sum aggregate with negative weights has been found and the output format smodels is used or the output is redirected to clasp.
+In this format negative bounds are compiled away using the transformation described in \cite{lparseManual}.
+This may lead to counterintuitive results.
  \item 
@@
 Use the command line option ``--convert'' to convert a ground program into another format.
+For converting already ground programs to an output language a special module has been implemented that uses far less memory.
+Because trying to ground an already variable free program could need a huge amount of memory.
 
@@
 \end{verbatim}
-This error message gives a list of predicates that are not lambda restricted. 
+This error message gives a list of variables that are not lambda restricted. 
 \item
@@
 The parser can not read the input format.
+\newline
+The next errors all belong to the ASPils output format.
+Some constructs in the input language can not be expressed in certain normal forms.
+Another normal form as to be chosen to support this feature.
 \begin{verbatim}
- "Disjunction not allowed on this normalform."
+ "Disjunction not allowed on this normal form,
+  please choose normal form 2 or 4-7."
  "Count aggregate/cardinality constraint not allowed
-  in this normalform."
+  in this normal form,
+  please choose normal form 3-7."
  "Count aggregate/cardinality constraint not allowed 
-  with non trivial upper bound in this normalform."
+  with non trivial upper bound in this normal form,
+  please choose normal form 5-7."
  "Count aggregate/cardinality constraint not allowed 
-  with negative lower bound in this normalform."
+  with negative lower bound in this normal form,
+  please choose normal form 5-7."
  "Sum aggregate/weight constraint not allowed
-  in this normalform."
+  in this normal form,
+  please choose normal form 3-7."
  "Sum aggregate/weight constraint not allowed 
-  with non trivial upper bound in this normalform."
+  with non trivial upper bound in this normal form,
+  please choose normal form 5-7."
  "Sum aggregate/weight constraint not allowed 
-  with negative lower bound in this normalform."
+  with negative lower bound in this normal form,
+  please choose normal form 5-7"
  "Sum aggregate/weight constraint not allowed
-  with negative weights in this normalform."
- "Sum aggregate/weight constraint not allowed
-  with negative weights in this normalform."
- "Max aggregate not allowed in this normalform."
+  with negative weights in this normal form,
+  please choose normal form 5 or 7."
+ "Max aggregate not allowed in this normal form,
+  please choose normal form 6 or 7."
  "Max aggregate not allowed
-  with negative weights in this normalform."
- "Min aggregate not allowed in this normalform."
+  with negative weights in this normal form,
+  please choose normal form 7."
+ "Min aggregate not allowed in this normal form,
+  please choose normal form 6 or 7."
  "Min aggregate not allowed 
-  with negative weights in this normalform."
- "Optimize statement not allowed in this normalform."
+  with negative weights in this normal form,
+  please choose normal form 7."
+ "Optimize statement not allowed in this normal form,
+  please choose normal form 3-7."
 \end{verbatim}
 %\begin{verbatim}
-% "Times aggregate not allowed in this normalform."
+% "Times aggregate not allowed in this normal form."
 % "Times aggregate not allowed
-%  with negative weights in this normalform."
+%  with negative weights in this normal form."
 %\end{verbatim}
-These errors belong to the ASPils output format. Another normal form as to be chosen.
 \end{itemize}

### 2008-07-30 lib/gringo/src/output.cpp

@@ -339,7 +339,7 @@ void Disjunction::print(Output *o, std::ostream &out)
 	
 	unsigned int normalForm = static_cast<PilsOutput*>(o)->getNormalForm();
 	if (normalForm == 1 || normalForm == 3)
-		throw GrinGoException("Disjunction not allowed on this normalform, please choose another output.");
+		throw GrinGoException("Disjunction not allowed on this normal form, please choose normal form 2 or 4-7.");
 
 	static_cast<PilsOutput*>(o)->addOptimizedID(uid_);
 	for(ObjectVector::iterator it = lits_.begin(); it != lits_.end(); it++)
@@ -478,13 +478,13 @@ void Aggregate::print(Output *o, std::ostream &out)
 		case COUNT:
 			{
 				if (normalForm == 1 || normalForm == 2)
-					throw GrinGoException("Count aggregate/cardinality constraint not allowed in this normalform, please choose another output.");
+					throw GrinGoException("Count aggregate/cardinality constraint not allowed in this normal form, please choose normal form 3-7.");
 				if (normalForm == 3 || normalForm == 4)
 				{
 					if (bounds_ == U || bounds_ == LU)
-						throw GrinGoException("Count aggregate/cardinality constraint not allowed with non trivial upper bound in this normalform, please choose another output.");
+						throw GrinGoException("Count aggregate/cardinality constraint not allowed with non trivial upper bound in this normal form, please choose normal form 5-7.");
 					if (lower_ < 0)
-						throw GrinGoException("Count aggregate/cardinality constraint not allowed with negative lower bound in this normalform, please choose another output.");
+						throw GrinGoException("Count aggregate/cardinality constraint not allowed with negative lower bound in this normal form, please choose normal form 5-7.");
 
 				}
 				out << "e";
@@ -493,45 +493,45 @@ void Aggregate::print(Output *o, std::ostream &out)
 		case SUM:
 			 {
 				if (normalForm == 1 || normalForm == 2)
-					throw GrinGoException("Sum aggregate/weight constraint not allowed in this normalform, please choose another output.");
+					throw GrinGoException("Sum aggregate/weight constraint not allowed in this normal form, please choose another output, please choose normal form 3-7");
 				if (normalForm == 3 || normalForm == 4)
 				{
 					if (bounds_ == U || bounds_ == LU)
-						throw GrinGoException("Sum aggregate/weight constraint not allowed with non trivial upper bound in this normalform, please choose another output.");
[…]

### 2008-07-30 lib/gringo/src/program.cpp

@@ -121,7 +121,7 @@ bool Program::check(Grounder *g)
 			if(it == rules_.end())
 			{
 				std::cerr << "the following rule cannot be grounded, ";
-				std::cerr << "weakly restricted varibles: { ";
+				std::cerr << "weakly restricted variables: { ";
 				bool comma = false;
 				for(VarVector::iterator it = free.begin(); it != free.end(); it++)
 				{

### 2008-07-31 doc/documentation/paper.tex
reworked documentation, fixed bug in 3colorability.lp
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@415 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 \usepackage{multirow}
+\usepackage{color}
 
@@
 The goal of this work is to give an overview of the features supported by gringo2.0.
-First, the general restrictions to the input language are stated.
 A broad set of potentialities for writing encodings for logic programs exists.
@@
 gringo2.0 supports several output formats and even a human readable format for debugging purposes.
-It also has an internal interface to the solver clasp \cite{gekanesc07b}, so the output can directly by parsed to the solver and answersets can be computed.
-Furthermore gringo can be used with clasp incrementally. Detailed information about incremental solving can be found here \cite{IncrementalASP} 
+It also has an internal interface to the solver clasp \cite{gekanesc07b}, so the output can directly by parsed to the solver and answer sets can be computed.
+Furthermore gringo2.0 can be used with clasp incrementally. Detailed information about incremental solving can be found here \cite{IncrementalASP} 
+First, the general restrictions to the input language are stated. Then a brief description of how to use gringo2.0 is given.
+In the next chapter the syntax of the supported language is described in detail and a short introduction of how to write a logic program is given.
+The paper closes with some examples of logic programs and gives an outlook of future work that has to be done in the area of grounding logic programs.
 The syntax of the gringo2.0 input language can be found in the appendix \ref{appendix}.
@@
 \label{domainrestricted}
-There exist two major grounders for ASP besides gringo, lParse \cite{lparseManual} and  the DLV grounding module\cite{DLV}.
+There exist two major grounders for ASP besides gringo1.0, lParse \cite{lparseManual} and  the DLV grounding module\cite{DLV}.
 To ensure the existence of an equivivalent ground instantiation of a program both grounders except a subclass of logic programs.
 For lParse the logic programs must be $\omega$-restricted\cite{omega} while DLV can handle safe programs\cite{safe}.
-Gringo verifies that the input program is $\lambda$-
+gringo2.0 verifies that the input program is $\lambda$-
 restricted \cite{gringopaper}, so that the existence of a finite equivalent ground instantiation is guaranteed. 
@@
 
-\section{Interface}
-Gringo can be started on the command line and has some parameters that can be useful.
-The first parameters are gringo specific parameters and consist of a number of options and any arbritary number of filenames of the logic program.
-For the special purpose that Gringo is used with its clasp module a double minus ``\texttt{--}'' is used to seperate the clasp parameters from the Gringo parameters.
+\section{Using gringo2.0}
+\subsection{Command line Interface}
+gringo2.0 can be started on the command line and has some parameters that can be useful.
+The first parameters are gringo2.0 specific parameters and consist of a number of options and any arbritary number of filenames of the logic program.
+For the special purpose that gringo2.0 is used with its clasp module a double minus ``\texttt{--}'' is used to seperate the clasp parameters from the gringo2.0 parameters.
+If no filename is given, gringo2.0 reads from standard input.
+\begin{verbatim}
+ ./gringo (gringo option|file)* [--[ clasp options]]
+\end{verbatim}
+\subsection{Options}
+
 \paragraph{\texttt{--help}}
@@
 \paragraph{\texttt{--const}}
-You can set the value of constants defined in your logic program via the command line of gringo.
+You can set the value of constants defined in your logic program via the command line of gringo2.0.
 After the parameter ``--const'' you have to give the name of the constant you want to set followed by a ``='' and the value of that constant.
@@
 If this parameter is used, the grounder expects an already grounded program as input.
-For converting an already grounded logic program into an output format supported by gringo this is the right option. This mode saves a lot of memory but does not ground the program.
+For converting an already grounded logic program into an output format supported by gringo2.0 this is the right option. This mode saves a lot of memory but does not ground the program.
 \paragraph{\texttt{-l}}
 This mode prints smodels conform output. For a definition of the output format see \cite{lparseManual}.
-Since some features of gringo are not natively supported for this output language, new atoms are introduced to create an equivalent program.\newline
-This is the current default output format of gringo.
+Since some features of gringo2.0 are not natively supported for this output language, new atoms are introduced to create an equivalent program.\newline
+This is the current default output format of gringo2.0.
 Sum aggregates with negative weights are handled like weight constraint in lparse for this output format.
@@
 \paragraph{\texttt{-c}}
-In this mode, the output is directly parsed to clasp, an ASP solver. For solving logic programs, no intermediate files are produced and the answersets are printed directly.
+In this mode, the output is directly parsed to clasp, an ASP solver. For solving logic programs, no intermediate files are produced and the answer sets are printed directly.
 With this option enabled, all parameters for clasp are valid and can be declared in the command line.
-Therefore adouble minus ``\texttt{--}'' must be given after all gringo related parameters are set.
-After them die clasp parameters follow.
+Therefore a double minus ``\texttt{--}'' must be given after all gringo2.0 related parameters are set.
+After them the clasp parameters follow.
 For example:
@@
 \paragraph{\texttt{-p}}
-Parameter ``\texttt{-p}'' invokes the plain output method, so the output is printed in human readable format, which is also the input format for gringo.
+Parameter ``\texttt{-p}'' invokes the plain output method, so the output is printed in human readable format, which is also the input format for gringo2.0.
+The output is variable free and some simplifications on the program were made.
+To convert the plain format into a format readable by a solver use the \texttt{--convert} option.
 \paragraph{\texttt{-g [1-7]}}
@@
 \paragraph{default parameters}
-The default parameter is ``\texttt{-l}'' for smodels output.
+The default command line is ``\texttt{./gringo -l}'' to read from standard input and for smodels output.
+
+\subsection{Warnings}
+Several warnings are produced by gringo2.0 to give hints on common mistakes.
+They are all printed to standard error output.
+
+\begin{itemize}
+ \item 
+\begin{verbatim}
+ "Warning: sum with negative bounds in the body of a rule"
+\end{verbatim}
+This warning occurs if a sum aggregate with negative weights has been found and the output format smodels is used or the output is redirected to clasp.
+In this format negative bounds are compiled away using the transformation described in \cite{lparseManual}.
+This may lead to counterintuitive results.
+ \item 
+\begin{verbatim}
+ "Warning: multiple definitions of #const n"
+\end{verbatim}
+This warning is printed if more than one \#const statement is used to set the same constant.
+The last attached value is used for grounding.
+\item
+\begin{verbatim}
+ "Warning: p/3 is never defined"
+\end{verbatim}
+This warnings states that an atom is used in the body of a rule that never occurs in the head.
+The predicate literal is given in the warning. The predicate will be removed trhough optimization.
+\item
+\begin{verbatim}
+ "got ground program i hope u have enough memory :)"
+\end{verbatim}
+Use the command line option ``--convert'' to convert a ground program into another format.
+For converting already ground programs to an output language a special module has been implemented that uses far less memory.
+Because trying to ground an already variable free program could need a huge amount of memory.
+
+\end{itemize}
+Furthermore all warnings from clasp are included.
+
+\subsection{Errors}
+gringo2.0 stops the execution and prints an error message to standard error output.
+\begin{itemize}
+ \item 
+\begin{verbatim}
+ "the following rule cannot not be grounded,
+  non domain predicates : { ... }"
+\end{verbatim}
+This error message gives a list of predicates that are not domain restricted.
+You explicitly have to give a domain for those predicates.
+ \item 
+\begin{verbatim}
+ "the following rule cannot not be grounded,
+  weakly restricted variables: { ... }"
+\end{verbatim}
+This error message gives a list of variables that are not lambda restricted. 
+\item
+\begin{verbatim}
+ "syntax error on line Y column Z unexpected token: X"
+\end{verbatim}
+The parser can not read the input format. Some unexpected token occured in the input file.
+\newline
+\newline
+The next errors all belong to the ASPils output format.
+Some constructs in the input language can not be expressed in certain normal forms.
+Another normal form as to be chosen to support this feature.
+\begin{verbatim}
+ "Disjunction not allowed on this normal form,
+  please choose normal form 2 or 4-7."
+ "Count aggregate/cardinality constraint not allowed
+  in this normal form,
+  please choose normal form 3-7."
+ "Count aggregate/cardinality constraint not allowed 
+  with non trivial upper bound in this normal form,
+  please choose normal form 5-7."
+ "Count aggregate/cardinality constraint not allowed 
+  with negative lower bound in this normal form,
+  please choose normal form 5-7."
+ "Sum aggregate/weight constraint not allowed
+  in this normal form,
+  please choose normal form 3-7."
+ "Sum aggregate/weight constraint not allowed 
+  with non trivial upper bound in this normal form,
+  please choose normal form 5-7."
+ "Sum aggregate/weight constraint not allowed 
+  with negative lower bound in this normal form,
+  please choose normal form 5-7"
+ "Sum aggregate/weight constraint not allowed
+  with negative weights in this normal form,
+  please choose normal form 5 or 7."
+ "Max aggregate not allowed in this normal form,
+  please choose normal form 6 or 7."
+ "Max aggregate not allowed
+  with negative weights in this normal form,
+  please choose normal form 7."
+ "Min aggregate not allowed in this normal form,
+  please choose normal form 6 or 7."
+ "Min aggregate not allowed 
+  with negative weights in this normal form,
+  please choose normal form 7."
+ "Optimize statement not allowed in this normal form,
+  please choose normal form 3-7."
+\end{verbatim}
+%\begin{verbatim}
+% "Times aggregate not allowed in this normal form."
+% "Times aggregate not allowed
+%  with negative weights in this normal form."
+%\end{verbatim}
+\end{itemize}
+
+\subsection{Building gringo2.0 from source}
+For compiling gringo2.0, a binary version of ``re2c'' is needed to compile the lexer and the parser.
+TODO: What about lemon.
+As the build environment ``cmake'' is used.
+TODO: How to build gringo2.0, clingo, iclingo etc...
+
 \section{Syntax}
@@
 The name of a constant can also be encompassed by double quotation marks, which is sometimes helpful when using constants with spaces.
-Tt has to be remarked that a constant with quotation marks is never equal to a constant without, so ``$a$'' is not equal to $a$.
+It has to be remarked that a constant with quotation marks is never equal to a constant without, so ``\texttt{a}'' is not equal to \texttt{a}.
 $C$ is the set of constants
@@
 The set of variables is denoted by $V$.
-The language over $P$, $F$and $C$ is denoted by $L(P,F,C)$.
+The language over $P$, $F$ and $C$ is denoted by $L(P,F,C)$.
 \paragraph{Terms}
 The building blocks of the language are terms.
-Each constant $c \in C$ and each variable $f \in F$ is a term.
+Each constant $c \in C$ and each function symbol $f \in F$ is a term.
 Furthermore $f(t_1, \dots, t_n)$ is a term if $f/n \in F$ and $t_i \mid i \in \{1 \dots n\}$ are terms.
-Since a term is defined inductively, it can be nested.
+Since a term is defined recursively, it can be nested.
 A term with the first parameter being a constant $a$, the others being terms $f(b,X)$ and $g(f(X), Y))$ would be expressed this way.
@@
 \paragraph{Atoms}
-An atom over $L(P, F, C)$ is of the form $p(t1 , . . . , tn )$ where $p/n$ is a predicate
+An atom over $L(P, F, C)$ is of the form $p(t_1, \dots, t_n)$ where $p/n$ is a predicate
 symbol with arity $n$, and $t_i \mid i \in \{1 \dots n\}$ are terms.
@@
 
+\subsection{How to write a logic program}
+Writing a logic program to solve a problem is rather describing a problem than describing an algorithm to solve a problem.
+Therefore the problem must be encoded as a set of rules.
+Typically these are divided into a problem description and an instance of the problem.
+The solution to a problem can then be read from the answer sets of the program.
+The procedure of writing a logic program shall now be explained with a little example.
+We want to write a problem encoding for the 3-colorability problem of graph coloring.
+Given an arbitraty graph, no two connected edges may have the same color.
+The most common technique is the ``generate \& test'' approach.
+Therefore, in the generator, rules are created that can derive a superset of the solution to the problem.
+In the tester all solutions that do not fit the constraints are omitted.
+\newline
+Usually you start with the non instance domains, in our example the colors we use.
+\begin{verbatim}
+ color(red). color(green). color(blue).
+\end{verbatim}
+After that, we start with the generator.
+Therefore we have to color all the vertices, but we must give attention that each vertex has only one color.
+We can do this by simply saying that a vertex is colored with one color $C$, and that the same vertex does not have any of the other colors $D$ or $E$.
+\begin{verbatim}
+ colored(V,C) :- not colored(V,D), not colored(V,E),
+                 C != D, C != E, D != E,
+                 color(E), color(C), color(D), vertex(V).
+\end{verbatim}
+As you can see, you have to give the domain of the variables in the rule.
+The variable $V$ is used for vertices and the variables $C$, $D$ and $E$ are used for colors.
+After every vertex now has a color we would get all possible solutions of all vertices colorfully painted.
+We now need to restrict our solutions in the tester by saying that two vertices that are connected by an edge should not have the same color.
+We can do this by using an integrity constraint.
+An integrity constraint is a rule with an empty head, and if an assignment of variables fulfills the body, the assignment is not a solution of our problem.
+\begin{verbatim}
+ :- edge(V,U), colored(V,C), colored(U,C),
+    vertex(V), vertex(U), color(C).
+\end{verbatim}
+Here we have the edge between two vertices $V$ and $U$, and if both are colored with the same color $C$ the assignment is not a solution to our problem.
+\newline
+We have finally written our problem description, all we need now is a problem instance. This step is fairly simple.
+We just need to define several vertices and edges that do connect the vertices.
+For the following simple graph
+\[
+ \xymatrix{1\ar@{-}[dr]\ar@{-}[d]\ar@{-}[r]&2\ar@{-}[dl]\\
+           3&4}
+\]
+
+we could just write
+\begin{verbatim}
+ vertex(1..4).
+ edge(1,2). edge(1,4).
+ edge(3,1). edge(3,2).
+\end{verbatim}
+
+If we give the problem description and instance to the grounder, it will generate a variable free version of the problem.
+This can then be solved by a solver and in the answer sets of the solver the ``\texttt{colored}'' predicate states a valid coloring of all vertices.
+We can do this by simply using the internal interface to clasp, to get a solution.
+\begin{verbatim}
+ ./gringo -c 3colorability.lp instance1.lp
+\end{verbatim}
+In the first answer set we find the ``\texttt{colored}'' predicates
+\begin{verbatim}
+ colored(1,green) colored(2,red) colored(3,blue) colored(4,red)
+\end{verbatim}
+and we can interpret them as the colors of the vertices
+
+\[
+ \xymatrix{\color{green}1\ar@{-}[dr]\ar@{-}[d]\ar@{-}[r]&\color{red}2\ar@{-}[dl]\\
+           \color{blue}3&\color{red}4}
+\]
+
 \subsection{Meta-Syntax}
@@
 The hide construct is used to hide atoms from the output of the ASP solver.
-Two different syntaxes are supported by gringo.
+Two different syntaxes are supported by gringo2.0.
 The first 
@@
 \end{verbatim}
-Both variants hide the predicate ``name''.
+Both variants hide the predicate ``\texttt{name}''.
 The arity of the predicate can be given explicit or implicit with a list of variables.
-The variables are just placeholder and do not unify with other variables. For example hide p(X,X) would also hide the fact p(a,b). The second syntax is implemented for compatibility reasons only.
+The variables are just placeholder and do not unify with other variables.
+For example \texttt{hide p(X,X)} would also hide the fact \texttt{p(a,b)}.
+The second syntax is implemented for compatibility reasons only.
 If the predicate that has to be hidden has arity zero the syntax is
@@
 \end{verbatim}
-Not to give explicitly all predicates that are hidden, only the ones which are visible the statement
-\begin{verbatim}
- #hide.
-\end{verbatim}
-in conjunction with the ``show'' statement is used.
-The ``show'' statement has the same syntax as ``hide'' but uses the keyword ``show''.
+Not to give explicitly all predicates that are hidden, only the ones which are visible the empty hide statement
+in conjunction with the ``\texttt{show}'' statement is used.
+With show all predicates are named that shall not be hidden from the hide statement.
+So to hide all predicates but $p/3$ the following example can be used.
 \begin{verbatim}
@@
 \end{verbatim}
-The example hides all atoms except the predicate $p$ with arity $3$.
+
 \paragraph{const}
-Within the ``const'' statement, a default value for constants can be defined in the logic program.
+Within the ``\texttt{const}'' statement, a default value for constants can be defined in the logic program.
 while they are normally defined via the command line options.
@@
 \end{verbatim}
-makes sure that the variable ``T'' is always bound to the domain of ``time''.
-This has the same effect as adding the domain ``time(T)'' to every rule the variable ``T'' occurs in.
+makes sure that the variable ``$T$'' is always bound to the domain of ``$time$''.
+This has the same effect as adding the domain ``\texttt{time(T)}'' to every rule the variable ``$T$'' occurs in.
 \paragraph{compute}
 The compute statement is a hint for the solver which models to compute.
-Is has the syntax
+It has the syntax
 \begin{verbatim}
@@
 \end{verbatim}
-$number$ is the number of models that shall be computed by the answerset solver and can be omitted. If it is zero, no restriction to the number of models is applied.
+$number$ is the number of models that shall be computed by the answer set solver and can be omitted. If it is zero, no restriction to the number of models is applied.
 All positive atoms in the compute statement must be in the computed models, and all negative atoms do not have to be in the models of the program.
-The compute statement is solver dependent and supported by every solver.
-\paragraph{Comments}
+The compute statement is solver dependent and not supported by every solver.
+\paragraph{comments}
 Comments included in the input are line based.
-Every line starting with ``\%'' is a comment. The whole line is omitted and not to be handled by the grounder.
+Whenever a ``\texttt{\%}'' occurs in the input, the rest of the line is treated as a comment and is therefore omitted.
 
-\subsection{How to write a logic program}
-Writing a logic program to solve a problem is rather describing a problem than describing an algorithm to solve a problem.
-Therefore the problem must be encoded as a set of rules.
-Typically these are divided into a problem description and an instance of the problem.
-The solution to a problem can then be read from the answer sets of the program.
-The procedure of writing a logic program shall now be explained with a little example.
-We want to write a problem encoding for the 3-colorability problem of graph coloring.
-Given an arbitraty graph, no two connected edges may have the same color.
-\newline
-Usually you start with the non instance domains, in our example the colors we use.
-\begin{verbatim}
- color(red). color(green). color(blue).
-\end{verbatim}
-After that, we have to color the vertices, but we must give attention that each vertex has only one color.
-We can do this by simply saying that a vertex is colored with one color X, and that the same vertex does not have another color Y.
-\begin{verbatim}
- colored(V,C) :- not colored(V,D), C != D, color(C;D), vertex(V).
-\end{verbatim}
-As you can see, you have to give the domain of the variables in the rule.
-The variable V is used for vertices and the variables C and D are used for colors.
-After every vertex now has a color we would get all possible solutions of all vertices colorfully painted.
-We now need to restrict our solutions again by saying that two vertices that are connected by an edge should not have the same color.
-We can do this by using an integrity constraint.
-An integrity constraint is a rule with an empty head, and if an assignment of variables fulfills the body, the assignment is not a solution of our problem.
-\begin{verbatim}
- :- edge(V,U), colored(V,C), colored(U,C), vertex(V;U), color(C).
-\end{verbatim}
-Here we have the edge between two vertices V and U, and if both are colored with the same color C the assignment is not a solution to our problem.
-\newline
-We have finally written our problem description, all we need now is a problem instance. This step is fairly simple.
-We just need to define several vertices and edges that do connect the vertices.
-\begin{verbatim}
- vertices(1..4).
- edge(1,2). edge(1,4). edge(2,4).
- edge(3,1). edge(4,1). edge(3,2).
-\end{verbatim}
-
-If we give the problem description and instance to the grounder, it will generate a variable free version of the problem.
-This can then be solved by a solver and in the answer sets of the solver the ``colored'' predicate states a valid coloring of all vertices.
-\newline
-To ease the process writing logic programs some features in the input syntax were introduced and are now explained in detail.
 \subsection{Features}
-The extended features of the input syntax are now explained in detail. For a full syntax description in EBNF see appendix (\ref{appendix}).
+To ease the process writing logic programs some features in the input syntax are introduced and are now explained in detail.
+For the most problems it is possible to write a much shorter version of the program with the extended syntax.
+%Even a speedup in solving the instances of the problem can happen TODO ?
+For a full syntax description in EBNF see appendix (\ref{appendix}).
 \paragraph{Function Symbols}
-Function symbols are fully supported with Gringo2.0.
+Function symbols are fully supported with gringo2.0.
 They can be finitly nested and also be used as domain predicates.
-While in lParse function symbols are allowed in a very restricted way, no restrictions apply in Gringo2.0.
-The following program can not be grounded with lParse since $X$ is an unrestricted variable. With Gringo2.0 the program is now groundable.
+While in lParse function symbols are allowed in a very restricted way, no restrictions apply in gringo2.0.
+The following program can not be grounded with lParse since $X$ is an unrestricted variable. With gringo2.0 the program is now groundable.
 \begin{verbatim}
@@
 
-\paragraph{Expanding Terms}
-\subparagraph{Range Operator}
+\paragraph{Range Operator}
 To define several atoms that stretch on one or more parameters over a certain range, two dots can be used to express an increasing integer range.
@@
 \end{verbatim}
-will result in the rules\footnote{gringo would simplify the grounded rules to facts, since all literals in the body are facts}
+will result in the rules
 \begin{verbatim}
@@
 With the expander ``:'' a domain for an atom can be set.
-On the left side of the operator an atom with variables is expected and on the right side the domain for that variables is given.
+On the left side of the operator an atom with variables is expected and on the right side the domains for the variables are given.
 In the head, the atoms get connected via a disjunction, in the body they get connected via a conjunction, so no new rules are created with this operator.
@@
 \paragraph{Conjunctor ;}
-The conjunctor ``;'' will create a conjunction of the its elements.
-If used in the head of a rule this means that the rule is duplicated.
+The conjunctor ``;'' will create a conjunction of its elements.
+Used in the head of a rule means that the rule is duplicated.
 If used in the body instead, the atoms will be logically combined using a conjunction.
@@
 \end{verbatim}
-will results in the rules
+will results in the rules\footnote{gringo2.0 would simplify the grounded rules since some literals in the body are facts}
 \begin{verbatim}
@@
 The bitwise operators are executed on signed integers using the binary complement for negative numbers.
-Arithmetic expressions can occur everywhere in head and in the body, even in complex constructs.
+Arithmetic expressions can occur everywhere in the head and in the body, even in complex constructs.
 As parameters, integer constants or expressions that evaluate to integers are used.
@@
 They can be used to constrain certain variables.
-For example, the following rule is only generated for all unifications where X is greater than Y-1.
+For example, the following rule is only generated for all unifications where $X$ is greater than $Y-1$.
 \begin{verbatim}
@@
 \end{verbatim}
-This would let $b(4,2)$ a valid substitution, since $4 >= 2-1$ but $b(0,2)$ not, so the grounded program could be:
+This would let $b(4,2)$ be a valid substitution, since $4 >= 2-1$ but $b(0,2)$ not, so the grounded program could be:
 \begin{verbatim}
@@
 \end{verbatim}
-Using disjunctions in logic programs increase expressiveness and complexity of the logic program.
+Using disjunctions in logic programs can increase expressiveness and complexity of the logic program.
 A solver for disjunctive logic programs such as claspD \cite{claspD} or cmodels \cite{cmodels} is needed to solve such programs.
@@
 \begin{verbatim}
- lower sum [atom1=w1, ..., atomN=wN] upper
+ lower [atom1=w1, ..., atomN=wN] upper
 \end{verbatim}
@@
 \subparagraph{Minimize/Maximize}
-An optimize statement is used to find a minimal or maximal answerset of the logic program.
+An optimize statement is used to find a minimal or maximal answer set of the logic program.
 The statement can be weighted or not, therefore weights can be omitted in the following syntax.
@@
 The maximize statement uses the keyword ``maximize'' instead.
-
-\section{Compiling Gringo}
-For compiling gringo, a binary version of ``re2c'' is needed to compile the lexer and the parser.
-TODO: What about lemon.
-As the build environment ``cmake'' is used.
-TODO: How to build gringo, clingo, iclingo etc...
-
-\subsection{Warnings}
-Several warnings are produced by gringo to give hints on common mistakes.
-They are all printed to standard error output.
-
-\begin{itemize}
- \item 
-\begin{verbatim}
- "Warning: sum with negative bounds in the body of a rule"
-\end{verbatim}
-This warning occurs if a sum aggregate with negative weights has been found and the output format smodels is used or the output is redirected to clasp.
-In this format negative bounds are compiled away using the transformation described in \cite{lparseManual}.
-This may lead to counterintuitive results.
- \item 
-\begin{verbatim}
- "Warning: multiple definitions of #const n"
-\end{verbatim}
-This warning is printed if more than one \#const statement is used to set the same constant.
-The last attached value is used for grounding.
-\item
-\begin{verbatim}
- "Warning: p/3 is never defined"
-\end{verbatim}
-This warnings states that an atom is used in the body of a rule that never occurs in the head.
-``p'' is the name of the predicate and ``3'' is the arity.
-\item
-\begin{verbatim}
- "got ground program i hope u have enough memory :)"
-\end{verbatim}
-Use the command line option ``--convert'' to convert a ground program into another format.
-For converting already ground programs to an output language a special module has been implemented that uses far less memory.
-Because trying to ground an already variable free program could need a huge amount of memory.
-
-\end{itemize}
-Furthermore all warnings from clasp are included.
-
-\subsection{Errors}
-gringo2.0 stops the execution and prints an error message to standard error output.
-\begin{itemize}
- \item 
-\begin{verbatim}
- "the following rule cannot not be grounded,
-  non domain predicates : { ... }"
-\end{verbatim}
-This error message gives a list of predicates that are not domain restricted.
- \item 
-\begin{verbatim}
- "the following rule cannot not be grounded,
-  weakly restricted variables: { ... }"
-\end{verbatim}
-This error message gives a list of variables that are not lambda restricted. 
-\item
-\begin{verbatim}
- "syntax error on line Y column Z unexpected token: X"
-\end{verbatim}
-The parser can not read the input format.
-\newline
-The next errors all belong to the ASPils output format.
-Some constructs in the input language can not be expressed in certain normal forms.
-Another normal form as to be chosen to support this feature.
-\begin{verbatim}
- "Disjunction not allowed on this normal form,
-  please choose normal form 2 or 4-7."
- "Count aggregate/cardinality constraint not allowed
-  in this normal form,
-  please choose normal form 3-7."
- "Count aggregate/cardinality constraint not allowed 
-  with non trivial upper bound in this normal form,
-  please choose normal form 5-7."
- "Count aggregate/cardinality constraint not allowed 
-  with negative lower bound in this normal form,
-  please choose normal form 5-7."
- "Sum aggregate/weight constraint not allowed
-  in this normal form,
-  please choose normal form 3-7."
- "Sum aggregate/weight constraint not allowed 
-  with non trivial upper bound in this normal form,
-  please choose normal form 5-7."
- "Sum aggregate/weight constraint not allowed 
-  with negative lower bound in this normal form,
-  please choose normal form 5-7"
- "Sum aggregate/weight constraint not allowed
-  with negative weights in this normal form,
-  please choose normal form 5 or 7."
- "Max aggregate not allowed in this normal form,
-  please choose normal form 6 or 7."
- "Max aggregate not allowed
-  with negative weights in this normal form,
-  please choose normal form 7."
- "Min aggregate not allowed in this normal form,
-  please choose normal form 6 or 7."
- "Min aggregate not allowed 
-  with negative weights in this normal form,
-  please choose normal form 7."
- "Optimize statement not allowed in this normal form,
-  please choose normal form 3-7."
-\end{verbatim}
-%\begin{verbatim}
-% "Times aggregate not allowed in this normal form."
-% "Times aggregate not allowed
-%  with negative weights in this normal form."
-%\end{verbatim}
-\end{itemize}
+If there were several minimize/maximize statements in the logic program the latter will be preferenced.
+Only after the n'th statement results in the same value for two different answer set, the n-1'th statement is considered.
 
@@
 bottom up grounding
+get rid of backward compatibility
+more aggregates
+
 \newpage

### 2008-07-31 lib/gringo/src/grounder.cpp
- fixed a bug initializing the output twice
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@416 3e02f596-3131-0410-8352-e04eeaec37cc

@@ -175,7 +175,6 @@ void Grounder::start()
 	reset();
 	std::cerr << "done" << std::endl;
 	std::cerr << "grounding ... " << std::endl;
-	output_->initialize(getPred());
 	substitution_.resize(varMap_.size() + 2);
 	binder_.resize(varMap_.size() + 2, -1);
 	ground();

### 2008-08-04 doc/documentation/paper.tex
further improvements, spellcheck again, removed Benchmarks, etc... all doc
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@417 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 It was developed by Roland Kaminski and is orientated on the ideas of gringo1.0.
+Therefore it supports Bindersplitting, Backjumping and all other features already implemented in gringo1.0.\cite{gringopaper}
 The goal of this work is to give an overview of the features supported by gringo2.0.
@@
 There exist two major grounders for ASP besides gringo1.0, lParse \cite{lparseManual} and  the DLV grounding module\cite{DLV}.
-To ensure the existence of an equivivalent ground instantiation of a program both grounders except a subclass of logic programs.
+To ensure the existence of an equivalent ground instantiation of a program both grounders except a subclass of logic programs.
 For lParse the logic programs must be $\omega$-restricted\cite{omega} while DLV can handle safe programs\cite{safe}.
@@
 gringo2.0 can be started on the command line and has some parameters that can be useful.
-The first parameters are gringo2.0 specific parameters and consist of a number of options and any arbritary number of filenames of the logic program.
-For the special purpose that gringo2.0 is used with its clasp module a double minus ``\texttt{--}'' is used to seperate the clasp parameters from the gringo2.0 parameters.
+The first parameters are gringo2.0 specific parameters and consist of a number of options and any arbitrary number of filenames of the logic program.
+For the special purpose that gringo2.0 is used with its clasp module a double minus ``\texttt{--}'' is used to separate the clasp parameters from the gringo2.0 parameters.
 If no filename is given, gringo2.0 reads from standard input.
@@
 Sum aggregates with negative weights are handled like weight constraint in lparse for this output format.
-So a transformation as explaned in \cite{lparseManual} is used to get rid of the negative weights.
+So a transformation as explained in \cite{lparseManual} is used to get rid of the negative weights.
 \paragraph{\texttt{-c}}
@@
 This warnings states that an atom is used in the body of a rule that never occurs in the head.
-The predicate literal is given in the warning. The predicate will be removed trhough optimization.
+The predicate literal is given in the warning. The predicate will be removed through optimization.
 \item
@@
 \end{verbatim}
-The parser can not read the input format. Some unexpected token occured in the input file.
+The parser can not read the input format. Some unexpected token occurred in the input file.
 \newline
@@
 \begin{itemize}
-\item a set P of predicate symbols,
-\item a set F of function symbols,
-\item a set V of variables,
-\item a set C of constants,
+\item a set $\mathcal{P}$ of predicate symbols,
+\item a set $\mathcal{F}$ of function symbols,
+\item a set $\mathcal{V}$ of variables,
+\item a set $\mathcal{C}$ of constants,
 \item a set \{ ``$\mid$'', ``,'' ,$\leftarrow$, not \} of connectives, and
@@
 
-TODO: Which greek letters for P, F, L and C
 \paragraph{Predicates and function symbols}
@@
 The names of the function symbols and predicates have to start with either a lower case letter or an underscore and can contain digits and letters.
-With $P$ we denote to the set of predicates and $F$ denotes the set of function symbols.
+With $\mathcal{P}$ we denote to the set of predicates and $\mathcal{F}$ denotes the set of function symbols.
 \paragraph{Constants}
@@
 It has to be remarked that a constant with quotation marks is never equal to a constant without, so ``\texttt{a}'' is not equal to \texttt{a}.
-$C$ is the set of constants
+$\mathcal{C}$ is the set of constants
 \paragraph{Variables}
 Variables have to start with a capitalized letter and can contain digits and letters.
-The set of variables is denoted by $V$.
-The language over $P$, $F$ and $C$ is denoted by $L(P,F,C)$.
+The set of variables is denoted by $\mathcal{V}$.
+The language over $\mathcal{P}$, $\mathcal{F}$ and $\mathcal{C}$ is denoted by $\mathcal{L(P,F,C)}$.
 \paragraph{Terms}
 The building blocks of the language are terms.
-Each constant $c \in C$ and each function symbol $f \in F$ is a term.
-Furthermore $f(t_1, \dots, t_n)$ is a term if $f/n \in F$ and $t_i \mid i \in \{1 \dots n\}$ are terms.
+Each constant $c \in \mathcal{C}$ and each function symbol $f \in \mathcal{F}$ is a term.
+Furthermore $f(t_1, \dots, t_n)$ is a term if $f/n \in \mathcal{F}$ and $t_i \mid i \in \{1 \dots n\}$ are terms.
 Since a term is defined recursively, it can be nested.
@@
 \paragraph{Atoms}
-An atom over $L(P, F, C)$ is of the form $p(t_1, \dots, t_n)$ where $p/n$ is a predicate
+An atom over $\mathcal{L(P, F, C)}$ is of the form $p(t_1, \dots, t_n)$ where $p/n$ is a predicate
 symbol with arity $n$, and $t_i \mid i \in \{1 \dots n\}$ are terms.
 \paragraph{Rules}
-A rule over the language $L(P,F,C)$ has the form
+A rule over the language $\mathcal{L(P,F,C)}$ has the form
 $$
@@
 where $A_0$ to $A_n$ are atoms. \newline
-A normal logic program $\Pi$ over $L(P,F,C)$ is a finite set of rules over $L(P,F,C)$.
+A normal logic program $\Pi$ over $\mathcal{L(P,F,C)}$ is a finite set of rules over $\mathcal{L(P,F,C)}$.
 
@@
 We want to write a problem encoding for the 3-colorability problem of graph coloring.
-Given an arbitraty graph, no two connected edges may have the same color.
+Given an arbitrary graph, no two connected edges may have the same color.
 The most common technique is the ``generate \& test'' approach.
@@
 Function symbols are fully supported with gringo2.0.
-They can be finitly nested and also be used as domain predicates.
+They can be finitely nested and also be used as domain predicates.
 While in lParse function symbols are allowed in a very restricted way, no restrictions apply in gringo2.0.
@@
 \end{verbatim}
-If squared brackets are used, the atom list is interpretated as a set, otherwise, for curly brackets, as a multiset of atoms.
+If squared brackets are used, the atom list is interpreted as a set, otherwise, for curly brackets, as a multiset of atoms.
 The lower and upper bound is optional, and can be omitted.
@@
 \subparagraph{Max}
-The ``max'' aggregate is similiar to the minimum aggregate but considers the maximum of the weights.
+The ``max'' aggregate is similar to the minimum aggregate but considers the maximum of the weights.
 
@@
 
-% fail if we havnt found a better solution
+% fail if we have not found a better solution
  :- costs(C), C [ altserves(R,Dist) : dist(Dist) : restaurant(R,RK) = Dist ].
@@
 
-\section{Benchmarks}
-The benchmarks were run on an Athlon XP 1700+ with 1GB of RAM.
-TODO: Official (non internal wiki) link to the instances.
+%\section{Benchmarks}
+%The benchmarks were run on an Athlon XP 1700+ with 1GB of RAM.
+%TODO: Official (non internal wiki) link to the instances.
 
-\input{benchmark2.result}
+%\input{benchmark2.result}
 \section{Future Work}
-$\Pi$-restrictedness
-bottom up grounding
-get rid of backward compatibility
-more aggregates
-
+Further research has to be done on features in the input language.
+Additionally the support of incremental grounding will be improved.
+To speed up the grounding process research on a bottom up grounding procedure is done.
+To support an even less restricted input language the support for $\Pi$-restricted programs is planned in a future release of gringo.
+All these features will lead to more acceptance of ASP-systems as the programs become easier to write and the bottleneck of ASP, the grounding part, will be faster.
 \newpage

### 2008-08-04 doc/documentation/paper.tex
additional bitex references, thx to martin, small changes
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@418 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 It also has an internal interface to the solver clasp \cite{gekanesc07b}, so the output can directly by parsed to the solver and answer sets can be computed.
-Furthermore gringo2.0 can be used with clasp incrementally. Detailed information about incremental solving can be found here \cite{IncrementalASP} 
+Furthermore gringo2.0 can be used with clasp incrementally. Detailed information about incremental solving can be found here \cite{gekaosscth08a} 
 First, the general restrictions to the input language are stated. Then a brief description of how to use gringo2.0 is given.
@@
 \paragraph{\texttt{-g [1-7]}}
-This parameter invokes the new output format for the intermediate languages ASPils \cite{atleastthisreferenceishouldgive}.
+This parameter invokes the new output format for the intermediate languages ASPils \cite{gejaosscth08b}.
 The optional number can range from $1$ to $7$ and refers to one of the following normal forms (Simple=1, SModels=2, DisjunctiveLogicProgram=3, CModels=4, CModelsExtended=5, DLV=6, Conglomeration=7).
@@
 \subsection{Meta-Syntax}
+This section is about constructs that do not originally belong to the logic program but can be used to give additional information for example to the solver.
 The ``\#'' sign in front of all meta expressions is optional but is it strongly recommend to use it.
@@
 
+% this can be set via the command line
 #const queens = 200.
+
 #const diags  = queens * 2 - 1.
-
 d(1..queens).

### 2008-08-04 lib/gringo/src/aggregateliteral.cpp
- weakened the check for aggregates   (one less dependency is added now)
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@419 3e02f596-3131-0410-8352-e04eeaec37cc

@@ -120,8 +120,8 @@ void AggregateLiteral::getVars(VarSet &vars) const
 SDGNode *AggregateLiteral::createNode(SDG *dg, SDGNode *prev, DependencyAdd todo)
 {
 	// TODO: this is only needed as long as the truth value of aggregates is not determined
-	if(todo == ADD_BODY_DEP)
-		prev->addDependency(prev, true);
+	//if(todo == ADD_BODY_DEP)
+	//	prev->addDependency(prev, true);
 	for(ConditionalLiteralVector::iterator it = literals_->begin(); it != literals_->end(); it++)
 	{
 		// aggregate literals always depend negativly on its literals 

### 2008-08-05 doc/documentation/paper.tex
DOKU: small changes, fixed errors, added some sentences, nothing of interest
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@420 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 It was developed by Roland Kaminski and is orientated on the ideas of gringo1.0.
-Therefore it supports Bindersplitting, Backjumping and all other features already implemented in gringo1.0.\cite{gringopaper}
+Therefore it supports binder splitting, backjumping and all other features already implemented in gringo1.0.\cite{gringopaper}
 The goal of this work is to give an overview of the features supported by gringo2.0.
@@
 gringo2.0 supports several output formats and even a human readable format for debugging purposes.
-It also has an internal interface to the solver clasp \cite{gekanesc07b}, so the output can directly by parsed to the solver and answer sets can be computed.
+It also has an internal interface to the solver clasp \cite{gekanesc07b}, so the output can directly be given to the solver and answer sets can be computed.
 Furthermore gringo2.0 can be used with clasp incrementally. Detailed information about incremental solving can be found here \cite{gekaosscth08a} 
-First, the general restrictions to the input language are stated. Then a brief description of how to use gringo2.0 is given.
-In the next chapter the syntax of the supported language is described in detail and a short introduction of how to write a logic program is given.
+In this paper, at first, the general restrictions to the input language are stated. Then a brief description of how to use gringo2.0 is given.
+In the next chapter the supported input language is described in detail and a short introduction of how to write a logic program is given.
 The paper closes with some examples of logic programs and gives an outlook of future work that has to be done in the area of grounding logic programs.
@@
 
-\section{Using gringo2.0}
-\subsection{Command line Interface}
-gringo2.0 can be started on the command line and has some parameters that can be useful.
+\section{Using Gringo2.0}
+\subsection{Command Line Interface}
+gringo2.0 can be started from the command line and has some parameters that can be useful.
 The first parameters are gringo2.0 specific parameters and consist of a number of options and any arbitrary number of filenames of the logic program.
@@
 Gives a short overview of all parameters.
-\paragraph{\texttt{--const}}
+\paragraph{\texttt{--const c=n}}
 You can set the value of constants defined in your logic program via the command line of gringo2.0.
-After the parameter ``--const'' you have to give the name of the constant you want to set followed by a ``='' and the value of that constant.
-The integer value can even be an arithmetic expression.
+After the parameter ``--const'' you have to give the name of the constant $c$ you want to set followed by a ``$=$'' and the value of that constant.
+The integer value $n$ can even be an arithmetic expression.
 For example:
@@
 \end{verbatim}
-the atom $n$ is replaced by $10$ in the logic program.
+the constant $c$ is replaced by $10$ in the logic program.
 
@@
 This mode prints smodels conform output. For a definition of the output format see \cite{lparseManual}.
-Since some features of gringo2.0 are not natively supported for this output language, new atoms are introduced to create an equivalent program.\newline
+Since some features of gringo2.0 are not natively supported for this output language, new atoms may be introduced to create an equivalent program.\newline
 This is the current default output format of gringo2.0.
@@
 \end{verbatim}
-Furthermore all restriction on smodels output hold as clasps internal interface currently reads smodels format.
+Furthermore all restriction on smodels output hold, as clasps internal interface currently reads smodels format.
 \paragraph{\texttt{-p}}
@@
 \paragraph{\texttt{-g [1-7]}}
-This parameter invokes the new output format for the intermediate languages ASPils \cite{gejaosscth08b}.
+This parameter invokes the new output format for the intermediate language ASPils \cite{gejaosscth08b}.
 The optional number can range from $1$ to $7$ and refers to one of the following normal forms (Simple=1, SModels=2, DisjunctiveLogicProgram=3, CModels=4, CModelsExtended=5, DLV=6, Conglomeration=7).
@@
 
-\paragraph{default parameters}
+\paragraph{Default Parameters}
 The default command line is ``\texttt{./gringo -l}'' to read from standard input and for smodels output.
@@
 
-\subsection{Building gringo2.0 from source}
+\subsection{Building Gringo2.0 From Source}
 For compiling gringo2.0, a binary version of ``re2c'' is needed to compile the lexer and the parser.
@@
 
-\paragraph{Predicates and function symbols}
+\paragraph{Predicates and Function Symbols}
 Predicates and function symbols have a predicate symbol/name and an arity.
@@
 It has to be remarked that a constant with quotation marks is never equal to a constant without, so ``\texttt{a}'' is not equal to \texttt{a}.
-$\mathcal{C}$ is the set of constants
+$\mathcal{C}$ is the set of constants.
 \paragraph{Variables}
@@
 A term with the first parameter being a constant $a$, the others being terms $f(b,X)$ and $g(f(X), Y))$ would be expressed this way.
+\newline
+TODO: This is not a term following the definition, it contains variables!!! Is a variable a Term ?
 \begin{verbatim}
@@
 where $A_0$ to $A_n$ are atoms. \newline
-A normal logic program $\Pi$ over $\mathcal{L(P,F,C)}$ is a finite set of rules over $\mathcal{L(P,F,C)}$.
+A logic program $\Pi$ over $\mathcal{L(P,F,C)}$ is a finite set of rules over $\mathcal{L(P,F,C)}$.
 
@@
 
-\subsection{How to write a logic program}
+\subsection{How to Write a Logic Program}
 Writing a logic program to solve a problem is rather describing a problem than describing an algorithm to solve a problem.
@@
 \newline
-Usually you start with the non instance domains, in our example the colors we use.
+Usually you start with the with the generator.
+At first, we define the non instance domains, in our example the colors we use.
 \begin{verbatim}
@@
 \end{verbatim}
-After that, we start with the generator.
-Therefore we have to color all the vertices, but we must give attention that each vertex has only one color.
+Now we have to color all the vertices, but we must give attention that each vertex has only one color.
 We can do this by simply saying that a vertex is colored with one color $C$, and that the same vertex does not have any of the other colors $D$ or $E$.
@@
 
-If we give the problem description and instance to the grounder, it will generate a variable free version of the problem.
+If we give the problem description and the instance to the grounder, it will generate a variable free version of the problem.
 This can then be solved by a solver and in the answer sets of the solver the ``\texttt{colored}'' predicate states a valid coloring of all vertices.
@@
 \end{verbatim}
-Not to give explicitly all predicates that are hidden, only the ones which are visible the empty hide statement
-in conjunction with the ``\texttt{show}'' statement is used.
-With show all predicates are named that shall not be hidden from the hide statement.
+If you want to express that all but a few predicates shall be hidden the empty hide statement
+in conjunction with the ``\texttt{show}'' statement can be is used.
+With ``\texttt{show}'' all predicates are named that shall be visible.
 So to hide all predicates but $p/3$ the following example can be used.
@@
 \end{verbatim}
-makes sure that the variable ``$T$'' is always bound to the domain of ``$time$''.
-This has the same effect as adding the domain ``\texttt{time(T)}'' to every rule the variable ``$T$'' occurs in.
+makes sure that the variable ``\texttt{T}'' is always bound to the domain of ``\texttt{time}''.
+This has the same effect as adding the domain ``\texttt{time(T)}'' to every rule the variable ``\texttt{T}'' occurs in.
 \paragraph{compute}
@@
 $number$ is the number of models that shall be computed by the answer set solver and can be omitted. If it is zero, no restriction to the number of models is applied.
-All positive atoms in the compute statement must be in the computed models, and all negative atoms do not have to be in the models of the program.
+All positive atoms in the compute statement must be in the computed models, and all negated atoms do not have to be in the models of the program.
 The compute statement is solver dependent and not supported by every solver.
@@
 \subsection{Features}
-To ease the process writing logic programs some features in the input syntax are introduced and are now explained in detail.
+To ease the process of writing logic programs some features in the input syntax are introduced and are now explained in detail.
 For the most problems it is possible to write a much shorter version of the program with the extended syntax.
@@
 \end{verbatim}
-If the range operator is used in an aggregate, it creates a a comma separated list of atoms, so
+If the range operator is used in an aggregate, it creates a comma separated list of atoms, so
 \begin{verbatim}
@@
 \paragraph{Expander :}
-With the expander ``:'' a domain for an atom can be set.
+With the expander ``:'' a domain can be set.
 On the left side of the operator an atom with variables is expected and on the right side the domains for the variables are given.
@@
 They can be used to constrain certain variables.
-For example, the following rule is only generated for all unifications where $X$ is greater than $Y-1$.
+For example, the following rule is only generated for all unifications where $X$ is greater or equal $Y-1$.
 \begin{verbatim}
@@
 \end{verbatim}
-This would let $b(4,2)$ be a valid substitution, since $4 >= 2-1$ but $b(0,2)$ not, so the grounded program could be:
+This would let $\texttt{b(4,2)}$ be a valid substitution, since $4 >= 2-1$ but $\texttt{b(0,2)}$ not, so the grounded program could be:
 \begin{verbatim}
@@
 \paragraph{Classical Negation}
-True negation can be used. Therefore the atom is preceded by a minus sign.
+For classical negation the atom is preceded by a minus sign.
 \begin{verbatim}
@@
 \subparagraph{Sum}
-The ``sum'' aggregate is a multiset of atoms with weight and has an optional lower and upper bound.
+The ``sum'' aggregate is a multiset of weighted atoms and has an optional lower and upper bound.
 It evaluates to true if the sum of the weights of all true atoms in the multiset is between the lower an upper bound.
@@
 \section{Examples}
-In this section come encodings for common problems are presented and explained.
+In this section some encodings for common problems are presented and explained.
 \paragraph{Fastfood}
@@
 All these features will lead to more acceptance of ASP-systems as the programs become easier to write and the bottleneck of ASP, the grounding part, will be faster.
+Furthermore the acceptance of a standardized intermediate format for ASP is necessary.
+gringo2.0 supports the ASPils output format, but many more tools for converting, shuffling and merging must be developed.
 \newpage

### 2008-08-05 app/clasp_main.h
- updated to clasp version 1.1.0   - there is a bug in the answerset enumerating code - added debug code to clasp-/iclaspoutput - every incremental grounding step gets its own false atom now   (this seems not to be needed)
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@421 3e02f596-3131-0410-8352-e04eeaec37cc

@@ -18,11 +18,10 @@
 // Foundation, Inc., 51 Franklin St, Fifth Floor, Boston, MA  02110-1301  USA
 //
 #include <clasp/include/lparse_reader.h>
-#include <clasp/include/unfounded_check.h>
-#include <clasp/include/smodels_constraints.h>
 #include <clasp/include/satelite.h>
-#include <clasp/include/util/misc_types.h>
+#include <clasp/include/unfounded_check.h>
 #include "options.h"
+#include "enumerator.h"
 #include "timer.h"
 #include <iostream>
 #include <iomanip>
@@ -80,11 +79,12 @@ private:
 		}
 		return std::cin;
 	}
-	bool solveAsp();
+	bool solve();
 	bool parseLparse();
-	bool solveSat();
-	std::auto_ptr<LparseStats> lpStats_;
-	std::auto_ptr<PreproStats> preStats_;
+	bool readSat();
+	std::auto_ptr<LparseStats>	lpStats_;
+	std::auto_ptr<PreproStats>	preStats_;
+	std::auto_ptr<Enumerator>		enum_;
 	std::ifstream	inFile_;
 } clasp_g;
 
@@ -93,6 +93,7 @@ const int S_UNKNOWN				=	0;
 const int S_SATISFIABLE		= 10;
 const int S_UNSATISFIABLE = 20;
 const int S_ERROR					= EXIT_FAILURE;
+const int S_MEMORY				= 107;
 
 static void sigHandler(int) {
 	// ignore further signals
@@ -108,36 +109,6 @@ static void sigHandler(int) {
 	printf("%sRestarts  : %u\n", (satM?"c ":""), (uint32)clasp_g.solver.stats.restarts);
 	exit(S_UNKNOWN);
 }
-
-struct StdOutPrinter : ModelPrinter {
-	StdOutPrinter() : index(0) {}
-	const AtomIndex*	index;
-	void printModel(Solver& s) {
-		if (index && clasp_g.mode == ClaspApp::ASP_MODE) {
-			cout << "Answer: " << s.stats.models << "\n";
-			for (AtomIndex::size_type i = 0; i != index->size(); ++i) {
-				if (s.value((*index)[i].lit.var()) == trueValue((*index)[i].lit) && !(*index)[i].name.empty()) {
-					cout << (*index)[i].name << " ";
-				}
-			}
-			cout << "\n";
[…]

### 2008-08-05 app/options.cpp

@@ -136,11 +136,12 @@ void Options::setDefaults() {
 	satPreParams.assign(3, 0);
 	file							= "";
 	heuristic					= "berkmin";
+	cons							= "";
 	numModels					= 1;
 	seed							= -1;
 	loopRep						= DefaultUnfoundedCheck::common_reason;
 	lookahead					= -1;
-	eqIters						= -1;
+	eqIters						= 5;
 	transExt					= LparseReader::transform_no;
 	help							= false;	
 	version						= false;
@@ -149,7 +150,7 @@ void Options::setDefaults() {
 	suppModels				= false;
 	stats							= false;
 	dimacs						= false;
-	optimizeAll				= false;
+	optimize					= 0;
 	ccmExp						= false;
 }
 
@@ -172,15 +173,18 @@ void Options::initOptions(ProgramOptions::OptionGroup& allOpts, ProgramOptions::
 			"\t  no    : Don't transform extended rules; handle them natively")
 		("eq", value<int>(&eqIters), 
 			"Run Eq-Preprocessor for at most <n> passes.\n"
-			"\tDefault: -1 (run to fixpoint). 0 = disable Eq-Preprocessor\n", "<n>")
+			"\tDefault: -1 (run to fixpoint); 0 = disable Eq-Preprocessor\n", "<n>")
 		("sat-prepro", value<vector<int> >(&satPreParams)->parser(mapSatElite),
 			"Enable SatElite-like preprocessor\n"
 			"\tDefault: no (yes if --dimacs), Valid: yes/no or <n1,[n2,n3]>\n"
 			"\t  n1 : Max iterations      (-1 means run to fixpoint)\n"
 			"\t  n2 : Heuristic cutoff    (-1 means no cutoff) \n"
 			"\t  n3 : Max time in seconds (-1 means no time limit)\n")
-		("optimize-all", bool_switch(&optimizeAll), "When optimizing compute all optimal solutions")
-		("optimize-value", value<std::vector<int> >(&optVals), "Initialize the optimization function")
+		("opt-all", bool_switch(), "When optimizing compute all optimal solutions")
+		("opt-rand",bool_switch(), "When computing one optimal solution randomize search")
+		("opt-value", value<std::vector<int> >(&optVals),"Initialize the optimization function")
+		("brave", bool_switch(), "compute the set of brave consequences\n")
[…]

### 2008-08-05 app/options.h

@@ -32,7 +32,7 @@
 #include <clasp/include/solver.h>
 #include <clasp/include/solve_algorithms.h>
 
-const char* const VERSION = "1.0.5-SR08";
+const char* const VERSION = "1.1.0";
 
 namespace Clasp {
 
@@ -53,18 +53,19 @@ public:
 	std::vector<int> satPreParams;	// Params for the SatElite-preprocessor
 	std::string		file;							// Default: "" -> read from stdin
 	std::string		heuristic;				// Default: berkmin
+	std::string		cons;							// Default: ""
 	int						seed;							// Default: -1 -> use default seed
 	int						transExt;					// Default: 0 -> do not transform extended rules
 	int						eqIters;					// Default: -1 -> run eq-preprocessing to fixpoint
 	int						numModels;				// Default: 1
 	int						lookahead;				// Default: lookahead_no
 	int						loopRep;					// Default: common
+	int						optimize;					// Default: 0
 	bool					help;							// Default: false
 	bool					version;					// Defailt: false
 	bool					quiet;						// Default: false
 	bool					stats;						// Default: false
 	bool					dimacs;						// Default: false
-	bool					optimizeAll;			// Default: false
 	bool					suppModels;				// Default: false
 	bool					initialLookahead;	// Default: false
 	bool					ccmExp;						// Default: false

### 2008-08-05 app/program_opts/program_options.cpp

@@ -36,7 +36,7 @@ namespace ProgramOptions {
 				temp << ",-" << o.shortName();
 			}
 			if (!o.getValue()->isImplicit())
-				temp << '=' << o.argDescription();
+				temp << (o.shortName().empty()?'=':' ') << o.argDescription();
 
 			os << temp.str();
 			if (!o.description().empty())
@@ -248,7 +248,7 @@ namespace ProgramOptions {
 	{
 		string::size_type n = name.find(',');
 		string shortName, longName, desc, arg;
-		arg = "arg";
+		arg = "<arg>";
 		if (description)
 			desc = description;
 		if (argDesc) {

### 2008-08-05 lib/clasp/include/heuristics.h

@@ -334,18 +334,27 @@ public:
 	void undoUntil(const Solver&, LitVec::size_type);
 	void simplify(Solver&, LitVec::size_type);
 	void resurrect(Var v) {
-		queue_.update(v);	
+		vars_.update(v);	
 	}
 private:
 	Literal doSelect(Solver& s);
-	void updateVarActivity(Var v);
-	void decayActivity();
+	void updateVarActivity(Var v) {
+		if ( (score_[v].first += inc_) > 1e100 ) {
+			for (LitVec::size_type i = 0; i != score_.size(); ++i) {
+				score_[i].first *= 1e-100;
+			}
+			inc_ *= 1e-100;
+		}
+		if (vars_.is_in_queue(v)) {
+			vars_.increase(v);
+		}
+	}
 	// HScore.first: activity of a variable
 	// HScore.second: occurrence counter for a variable's literal
 	//	> 0: positive literal occurred more often
 	//	< 0: negative literal occurred more often
 	//	= 0: literals occurred equally often
-	typedef std::pair<uint32, int32>	HScore;
+	typedef std::pair<double, int32>	HScore;
 	typedef PodVector<HScore>::type		Scores;
 	struct GreaterActivity {
 		explicit GreaterActivity(const Scores& s) : sc_(s) {}
@@ -356,9 +365,11 @@ private:
 		GreaterActivity& operator=(const GreaterActivity&);
 		const Scores& sc_;
 	};
-	Scores score_;
-	bk_lib::indexed_priority_queue<GreaterActivity> queue_;
-	bool considerLoops_;
+	typedef bk_lib::indexed_priority_queue<GreaterActivity> VarOrder;
+	Scores		score_;
+	VarOrder	vars_;
+	double		inc_;
+	bool			scoreLoops_;
 };
 
 }

### 2008-08-05 lib/clasp/include/program_builder.h

@@ -36,13 +36,13 @@ namespace Clasp {
 
 class Solver;
 class ClauseCreator;
-class MinimizeConstraint;
 class PrgNode;
 class PrgBodyNode;
 class PrgAtomNode;
 class ProgramBuilder;
 class Preprocessor;
 class DefaultUnfoundedCheck;
+class MinimizeConstraint;
 typedef PodVector<PrgAtomNode*>::type AtomList;
 typedef PodVector<PrgBodyNode*>::type BodyList;
 typedef PodVector<PrgNode*>::type NodeList;
@@ -306,7 +306,8 @@ public:
 	 * Call this method to load the program (increment) into the solver
 	 *
 	 * \param solver The solver object in which the optimized representation of the lp should be stored.
-	 * \param initialLookahead if true, the program is simplified using a one-step lookahead operation.
+	 * \param initialLookahead if true and finalizeSolver is true, the program is simplified using a one-step lookahead operation.
+	 * \param finalizeSolver if true, endProgram() calls solver.endAddConstraints()
 	 * \return false if the program is initially conflicting, true otherwise.
 	 *
 	 * \note If endProgram returned false, the state of the ProgramBuilder object is undefined. In that case,
@@ -317,7 +318,17 @@ public:
 	 * \note To load the same program into different solvers, call endProgram for each solver.
 	 *
 	 */
-	bool endProgram(Solver& solver, bool initialLookahead = false);
+	bool endProgram(Solver& solver, bool initialLookahead = false, bool finalizeSolver = true);
+
+	//! returns true if the program contains at least one minimize statement
+	bool hasMinimize() const { return minimize_ != 0; }
+	
+	//! Creates a minimize constraints from the minimize statements contained in the program
+	/*!
+	 * \pre The program is frozen, ie. endProgram was already called.
+	 * \return 0 if the program does not contain minimize statements
+	 */
+	MinimizeConstraint* createMinimize(Solver& s);
 
 	/*!
 	 * Writes the (possibly simplified) program in lparse-format to the given stream.
@@ -361,6 +372,7 @@ private:
 	void cloneVars(Solver& s);
 	void minimize(Solver& s);
 	bool addConstraints(Solver& s, CycleChecker& c, uint32 startAtom);
+	void freezeMinimize(Solver&);
 	// ------------------------------------------------------------------------
 	void writeRule(PrgBodyNode*, uint32 h, std::ostream&);
 

### 2008-08-05 lib/clasp/include/smodels_constraints.h

@@ -162,7 +162,7 @@ public:
 	uint32 numRules() const { return (uint32)minRules_.size(); }
 
 	//! Sets the comparison mode used in this constraint
-	void setMode(Mode m) { mode_ = m; }
+	void setMode(Mode m, bool restartAfterModel = false) { mode_ = m; restart_ = (mode_ == compare_less && restartAfterModel); }
 
 	//! Sets the optimum of a minimize-statement 
 	/*!
@@ -204,6 +204,7 @@ public:
 	//! returns the number of optimal models found so far
 	uint64 models() const;
 	void printOptimum(std::ostream& os) const;
+	bool allowRestart() const { return restart_; }
 	
 	// constraint interface
 	PropResult propagate(const Literal& p, uint32& data, Solver& s);
@@ -263,6 +264,7 @@ private:
 	uint32		activePL_;		// priority level under consideration
 	uint32		activeIdx_;		// index into rule of current priority level
 	Mode			mode_;				// how to compare assignments?
+	bool			restart_;			// Restart after each optimal model?
 };
 
 }

### 2008-08-05 lib/clasp/include/solve_algorithms.h

@@ -33,7 +33,7 @@
  */
 namespace Clasp { 
 
-class Solver;
+class		Solver;
 
 //! Interface for printing models
 /*!
@@ -42,7 +42,30 @@ class Solver;
 class ModelPrinter {
 public:
 	virtual ~ModelPrinter();
-	virtual void printModel(Solver&);
+	virtual void printModel(const Solver&) = 0;
+};
+
+//! Interface for enumerating models
+class Enumerator {
+public:
+	Enumerator();
+	virtual ~Enumerator();
+	void	setPrinter(ModelPrinter* p) { printer_ = p; }
+	//! Defaults to a noop
+	virtual void		updateModel(Solver& s);
+	//! Defaults to return s.backtrack()
+	virtual bool		backtrackFromModel(Solver& s);
+	//! Defaults to return false
+	virtual bool		allowRestarts()							const;
+	//! Defaults to return s.stats.models
+	virtual uint64	numModels(const Solver& s)	const;
+	//! Defaults to a noop
+	virtual void		report(const Solver& s)			const;
+protected:
+	ModelPrinter* printer_;
+private:
+	Enumerator(const Enumerator&);
+	Enumerator& operator=(const Enumerator&);
 };
 
 ///////////////////////////////////////////////////////////////////////////////
@@ -64,12 +87,11 @@ struct SolveParams {
 	 * deletion			: initial size: vars()/3, grow factor: 1.1, max factor: 3.0, do not reduce on restart
 	 * randomization: disabled
 	 * randomProp		: 0.0 (disabled)
-	 * model printer: not used (i.e. models are ignored)
+	 * enumerator		: default
 	 */
 	SolveParams();
 	
-	//! calls p.printModel(solver) whenever a model is found.
-	void setModelPrinter(ModelPrinter& p) { printer_ = &p; }
+	void setEnumerator(Enumerator& e)			{ enumerator_ = &e; }
 	
 	//! sets the restart-parameters to use during search.
 	/*!
@@ -158,15 +180,14 @@ struct SolveParams {
 	
 	uint32	shuffleBase() const { return shuffleFirst_; }
 	uint32	shuffleNext() const { return shuffleNext_; }
-	
-	ModelPrinter* printer() const { return printer_;}
+	Enumerator*		enumerator()	const { return enumerator_; }
 private:
 	double				reduceBase_;
 	double				reduceInc_;
 	double				reduceMaxF_;
[…]

### 2008-08-05 lib/clasp/include/solver.h

@@ -35,7 +35,6 @@ namespace Clasp {
 class DecisionHeuristic;
 class PostPropagator;
 class ClauseCreator;
-class MinimizeConstraint;
 
 //! Base class for preprocessors working on clauses only
 class SatPreprocessor {
@@ -93,7 +92,6 @@ struct SolverStrategies {
 	 *  - search					: use_learning
 	 *  - cflMin					: beame_minimization
 	 *	- cflMinAntes			: all_antes
-	 *	- minimizer				: none
 	 *  - compression			: 60
 	 *	- randomWatches		: false
 	 *	- saveProgress		: false
@@ -105,7 +103,6 @@ struct SolverStrategies {
 	Heuristic						heuristic;
 	PostProp						postProp;
 	SearchStrategy			search;
-	MinimizeConstraint*	minimizer;
 	CflMinStrategy			cflMin;
 	CflMinAntes					cflMinAntes;
 	bool								randomWatches;
@@ -121,7 +118,6 @@ struct SolverStrategies {
 		postProp.reset( o.postProp.release() );
 		search				= o.search;
 		cflMin				= o.cflMin;
-		minimizer			= o.minimizer;
 		compress_			= o.compress_;
 		randomWatches = o.randomWatches;
 		saveProgress	= o.saveProgress;

### 2008-08-05 lib/clasp/include/unfounded_check.h

@@ -206,23 +206,18 @@ public:
 		return 0;
 	}
 	// helpers
-	void enqueueTodo(UfsAtomNode* head) {
-		if (head->pickedOrTodo == 0) {
-			todo_.push_back(head);
-			head->pickedOrTodo = 1;
-		}
-	}
-	void enqueueUnfounded(UfsAtomNode* head) {
-		if (head->typeOrUnfounded == 0) {
-			unfounded_.push_back(head);
-			head->typeOrUnfounded = 1;
-		}
-	}
-	void enqueueUnsourced(UfsAtomNode* a) {
-		sourceQueue_.push_back(a);
-	}
+	// The body b can no longer be used as source pointer.
+	// Mark it for later source pointer removal
 	void enqueueInvalid(UfsBodyNode* b) {
 		invalid_.push_back(b);
+	}
+
+	// Mark atom a for later source pointer removal
+	// Pre: a->hasSource() == true
+	void enqueuePropagateUnsourced(UfsAtomNode* a) {
+		assert(a->hasSource() == true);
+		a->markSourceInvalid();
+		sourceQueue_.push_back(a);
 	}
 	void addWatch(Literal p, UfsBodyNode* b, uint32 data);
 	void addReasonLit(Literal p, const UfsBodyNode* reason);
@@ -269,6 +264,18 @@ private:
 	void createLoopFormula();
 // -------------------------------------------------------------------------------------------	
 // more helpers
+	void enqueueTodo(UfsAtomNode* head) {
+		if (head->pickedOrTodo == 0) {
+			todo_.push_back(head);
+			head->pickedOrTodo = 1;
+		}
+	}
+	void enqueueUnfounded(UfsAtomNode* head) {
+		if (head->typeOrUnfounded == 0) {
+			unfounded_.push_back(head);
+			head->typeOrUnfounded = 1;
+		}
+	}
 	UfsAtomNode* dequeueTodo() {
 		UfsAtomNode* head = todo_.front();
 		todo_.pop_front();

### 2008-08-05 lib/clasp/src/heuristics.cpp

@@ -502,28 +502,39 @@ Literal ClaspVmtf::doSelect(Solver& s) {
 /////////////////////////////////////////////////////////////////////////////////////////
 // ClaspVsids selection strategy
 /////////////////////////////////////////////////////////////////////////////////////////
+const double vsidsVarDecay_g = 1 / 0.95;
+
 ClaspVsids::ClaspVsids(bool loops) 
-	: queue_(GreaterActivity(score_)) 
-	, considerLoops_(loops) {}
+	: vars_(GreaterActivity(score_)) 
+	, inc_(1.0)
+	, scoreLoops_(loops) {}
 
 void ClaspVsids::startInit(const Solver& s) {
 	score_.clear();
 	score_.resize(s.numVars()+1);
 }
 void ClaspVsids::endInit(const Solver& s) {
-	queue_.clear();
+	vars_.clear();
+	inc_ = 1.0;
+	double maxS = 0.0;
 	for (Var v = 1; v <= s.numVars(); ++v) {
 		if (s.value(v) == value_free) {
 			// initialize activity to moms-score
 			score_[v].first = momsScore(s, v);
-			queue_.push(v);
+			if (score_[v].first > maxS) {
+				maxS = score_[v].first;
+			}
+			vars_.push(v);
 		}
+	}
+	for (VarVec::size_type i = 0; i != score_.size(); ++i) {
+		score_[i].first /= maxS;
 	}
 }
 
 void ClaspVsids::simplify(Solver& s, LitVec::size_type i) {
 	for (; i < s.numAssignedVars(); ++i) {
-		queue_.remove(s.assignment()[i].var());
+		vars_.remove(s.assignment()[i].var());
 	}
 }
 
@@ -531,50 +542,35 @@ void ClaspVsids::newConstraint(const Solver&, const Literal* first, LitVec::size
 	if (t != Constraint_t::native_constraint) {
 		for (LitVec::size_type i = 0; i < size; ++i, ++first) {
 			score_[first->var()].second += 1 - (first->sign() + first->sign());
-			if (t == Constraint_t::learnt_conflict || considerLoops_) {
+			if (t == Constraint_t::learnt_conflict || scoreLoops_) {
 				updateVarActivity(first->var());
 			}
+		}
+		if (t == Constraint_t::learnt_conflict) {
+			inc_ *= vsidsVarDecay_g;
 		}
 	}
 }
 
[…]

### 2008-08-05 lib/clasp/src/preprocessor.cpp

@@ -113,7 +113,7 @@ bool Preprocessor::preprocessSimple(bool bodyEq) {
 
 // If the program is defined incrementally, mark atoms from previous steps as supported
 void Preprocessor::updatePreviouslyDefinedAtoms(Var startAtom, bool strong) {
-	for (VarVec::size_type i = 0; i != startAtom; ++i) {
+	for (Var i = 0; i != startAtom; ++i) {
 		PrgAtomNode* a = prg_->atoms_[i];
 		if (!strong && a->visited() == 0) {
 			// a is an atom about to be unfrozen
@@ -122,7 +122,7 @@ void Preprocessor::updatePreviouslyDefinedAtoms(Var startAtom, bool strong) {
 		PrgAtomNode* aEq = 0;
 		if (strong) {
 			nodes_[i].aSeen = 1;
-			Var eq = prg_->getEqAtom(i);
+			uint32 eq = prg_->getEqAtom(i);
 			if (eq != i) {
 				aEq = prg_->atoms_[eq];
 			}
@@ -152,10 +152,10 @@ void Preprocessor::updatePreviouslyDefinedAtoms(Var startAtom, bool strong) {
 //
 /////////////////////////////////////////////////////////////////////////////////////////
 bool Preprocessor::preprocessEq(uint32 maxIters) {
-	nodes_.resize( std::max(prg_->bodies_.size(), prg_->atoms_.size()) );
+	nodes_.resize( prg_->bodies_.size() >= prg_->atoms_.size() ? prg_->bodies_.size() : prg_->atoms_.size() );
 	Var startAtom		= prg_->incData_?prg_->incData_->startAtom_ : 0;
 	Var stopAtom		= startAtom;
-	Var		startVar	= prg_->incData_?prg_->incData_->startVar_	: 1;
+	Var	startVar		= prg_->incData_?prg_->incData_->startVar_	: 1;
 	conflict_ = false;
 	pass_			= 0;
 	maxPass_	= maxIters;

### 2008-08-05 lib/clasp/src/program_builder.cpp

@@ -790,7 +790,7 @@ void ProgramBuilder::clearRuleState(const PrgRule& r) {
 	}	
 }
 
-bool ProgramBuilder::endProgram(Solver& solver, bool initialLookahead) {
+bool ProgramBuilder::endProgram(Solver& solver, bool initialLookahead, bool finalizeSolver) {
 	if (frozen_ == false) {
 		stats.bodies += (uint32)bodies_.size();
 		updateFrozenAtoms();
@@ -827,7 +827,7 @@ bool ProgramBuilder::endProgram(Solver& solver, bool initialLookahead) {
 		stats.ufsNodes = (uint32)ufs_->nodes();
 	}
 	if (incData_) atoms_.resize( atoms_.size() - incData_->unfreeze_.size() );
-	return ret && solver.endAddConstraints(initialLookahead);
+	return ret && (!finalizeSolver || solver.endAddConstraints(initialLookahead));
 }
 
 void ProgramBuilder::updateFrozenAtoms() {
@@ -910,15 +910,13 @@ bool ProgramBuilder::addConstraints(Solver& s, CycleChecker& c, uint32 startAtom
 		if ( !(*it)->toConstraint(s, gc, *this) ) { return false; }
 		c.visit(*it);
 	}
-	minimize(s);
+	freezeMinimize(s);
 	return true;
 }
-// adds minimize constraint
-void ProgramBuilder::minimize(Solver& solver) {
-	if (!minimize_) { return; }
+
+MinimizeConstraint* ProgramBuilder::createMinimize(Solver& solver) {
+	if (!minimize_) { return 0; }
 	MinimizeConstraint* m = new MinimizeConstraint();
-	solver.add(m);
-	solver.strategies().minimizer = m;
 	WeightLitVec lits;
 	for (MinimizeRule* r = minimize_; r; r = r->next_) {
 		for (WeightLitVec::iterator it = r->lits_.begin(); it != r->lits_.end(); ++it) {
@@ -929,6 +927,21 @@ void ProgramBuilder::minimize(Solver& solver) {
 		}
 		m->minimize(solver, lits);
 		lits.clear();
+	}
+	m->simplify(solver,false);
+	return m;
+}
+
+// exclude vars contained in minimize statements from var elimination
+void ProgramBuilder::freezeMinimize(Solver& solver) {
+	if (!minimize_) return;
+	for (MinimizeRule* r = minimize_; r; r = r->next_) {
[…]

### 2008-08-05 lib/clasp/src/satelite.cpp

@@ -214,6 +214,10 @@ bool SatElite::preprocess() {
 			Literal unit = c->size() == 1 ? (*c)[0] : negLit(0);
 			c->destroy(); clauses_[i] = 0;
 			if (!(solver_->force(unit, 0)&&solver_->propagate()) || !propagateFacts()) {
+				for (++i; i != clauses_.size(); ++i) {
+					clauses_[i]->destroy();
+				}
+				clauses_.erase(clauses_.begin()+j, clauses_.end());
 				return false;
 			}
 		}

### 2008-08-05 lib/clasp/src/smodels_constraints.cpp

@@ -227,7 +227,8 @@ MinimizeConstraint::MinimizeConstraint()
 	: models_(0)
 	, activePL_(0)
 	, activeIdx_(0)
-	, mode_(compare_less) {
+	, mode_(compare_less) 
+	, restart_(false) {
 	undoList_.push_back( UndoLit(posLit(0), 0, true) );	// sentinel
 }
 
@@ -407,11 +408,19 @@ bool MinimizeConstraint::backtrackFromModel(Solver& s) {
 	if (dl < s.rootLevel() || dl == static_cast<uint32>(-1)) {
 		return false;
 	}
-	while (s.backtrack() && dl < s.decisionLevel());
-	assert(dl == s.decisionLevel());
-	activePL_		= 0;
-	activeIdx_	= 0;
-	return backpropagate(s, rule(activePL_));
+	if (!restart_) {
+		while (s.backtrack() && dl < s.decisionLevel());
+		assert(dl == s.decisionLevel());
+		activePL_		= 0;
+		activeIdx_	= 0;
+		return backpropagate(s, rule(activePL_));
+	}
+	else {
+		s.undoUntil( 0 );
+		activePL_		= 0;
+		activeIdx_	= 0;
+		return select(s) && backpropagate(s, rule(activePL_));
+	}
 }
 
 void MinimizeConstraint::addUndo(Solver& s, Literal p, uint32 key, bool forced) {
@@ -457,7 +466,7 @@ bool MinimizeConstraint::select(Solver& s) {
 			}
 		}
 	}
-	return false;
+	return s.strategies().heuristic->select(s);
 }
 
 

### 2008-08-05 lib/clasp/src/solve_algorithms.cpp

@@ -25,8 +25,14 @@ using std::log;
 namespace Clasp { 
 
 ModelPrinter::~ModelPrinter() {}
-void ModelPrinter::printModel(Clasp::Solver&) {}
-static ModelPrinter noPrinter_s;
+Enumerator::Enumerator() : printer_(0) {}
+Enumerator::~Enumerator() {}
+void Enumerator::updateModel(Solver& s)							{ if (printer_) printer_->printModel(s); }
+bool Enumerator::backtrackFromModel(Solver& s)			{	return s.backtrack();		}
+bool Enumerator::allowRestarts() const							{ return false;						}
+uint64 Enumerator::numModels(const Solver& s)	const	{ return s.stats.models;	}
+void Enumerator::report(const Solver&)				const { }
+static Enumerator defaultEnum_s;
 
 /////////////////////////////////////////////////////////////////////////////////////////
 // SolveParams
@@ -35,7 +41,7 @@ SolveParams::SolveParams()
 	: reduceBase_(3.0), reduceInc_(1.1), reduceMaxF_(3.0)
 	, restartInc_(1.5)
 	, randProp_(0.0)
-	, printer_(&noPrinter_s)
+	, enumerator_(&defaultEnum_s)
 	, restartBase_(100)
 	, restartOuter_(0)
 	, randRuns_(0), randConflicts_(0)
@@ -98,12 +104,11 @@ private:
 /////////////////////////////////////////////////////////////////////////////////////////
 namespace {
 bool backtrackFromModel(Solver& s, const SolveParams& p) {
-	p.printer()->printModel(s);
+	p.enumerator()->updateModel(s);
 	if (s.strategies().satPrePro.get() && s.strategies().satPrePro->hasSymModel()) {
 		return true;
 	}
-	MinimizeConstraint* c = s.strategies().minimizer;
-	return !c ? s.backtrack() : c->backtrackFromModel(s);
+	return p.enumerator()->backtrackFromModel(s);
 }
 }
 
@@ -131,6 +136,7 @@ bool solve(Solver& s, uint32 maxAs, const SolveParams& p) {
 	double randProp	= randRuns == 0 ? p.randomPropability() : 1.0;
 	uint64 maxCfl		= randRuns == 0 ? rs.next() : p.randConflicts();
 	uint32 shuffle	= p.shuffleBase();
+	bool	noRestartAfterModel = !p.enumerator()->allowRestarts() && !p.restartBounded();
 	while (result == value_free) {
[…]

### 2008-08-05 lib/clasp/src/solver.cpp

@@ -101,7 +101,6 @@ SolverStrategies::SolverStrategies()
 	, heuristic(new SelectFirst) 
 	, postProp(new NoPostPropagator)
 	, search(use_learning)
-	, minimizer(0)
 	, cflMin(beame_minimization)
 	, cflMinAntes(all_antes)
 	, randomWatches(false)
@@ -450,24 +449,26 @@ bool Solver::unitPropagate() {
 			}
 		}
 		// and finally do general BCP
-		GWL& gWL = gwl(wl);
-		Constraint::PropResult r;
-		LitVec::size_type j;
-		for (j = 0, i = 0, end = gWL.size(); i != gWL.size(); ) {
-			Watch& w = gWL[i++];
-			r = w.propagate(*this, p);
-			if (r.second) {	// keep watch
-				gWL[j++] = w;
-			}
-			if (!r.first) {
-				while (i != end) {
-					gWL[j++] = gWL[i++];
+		if (!wl.gens.empty()) {
+			GWL& gWL = gwl(wl);
+			Constraint::PropResult r;
+			LitVec::size_type j;
+			for (j = 0, i = 0, end = gWL.size(); i != gWL.size(); ) {
+				Watch& w = gWL[i++];
+				r = w.propagate(*this, p);
+				if (r.second) {	// keep watch
+					gWL[j++] = w;
 				}
-				gWL.erase(gWL.begin()+j, gWL.end());
-				return false;
+				if (!r.first) {
+					while (i != end) {
+						gWL[j++] = gWL[i++];
+					}
+					gWL.erase(gWL.begin()+j, gWL.end());
+					return false;
+				}
 			}
+			gWL.erase(gWL.begin()+j, gWL.end());
 		}
-		gWL.erase(gWL.begin()+j, gWL.end());
 	}
 	return true;
 }

### 2008-08-05 lib/clasp/src/unfounded_check.cpp

@@ -54,8 +54,7 @@ public:
 		assert(choice_ == 1);
 		if (succs[data]->hasSource() && succs[data]->watch() == this) {
 			assert(ufs.solver().isFalse(succs[data]->lit));
-			succs[data]->markSourceInvalid();
-			ufs.enqueueUnsourced(succs[data]);
+			ufs.enqueuePropagateUnsourced(succs[data]);
 		}
 	}
 	bool canSource(const UfsAtomNode&)		{ return circWeight_ == 0; }
@@ -233,9 +232,9 @@ private:
 /////////////////////////////////////////////////////////////////////////////////////////
 DefaultUnfoundedCheck::DefaultUnfoundedCheck(ReasonStrategy r)
 	: solver_(0) 
+	, reasons_(0)
 	, activeClause_(0)
-	, strategy_(r)
-	, reasons_(0) {
+	, strategy_(r) {
 }
 DefaultUnfoundedCheck::~DefaultUnfoundedCheck() { clear(); }
 
@@ -436,8 +435,10 @@ DefaultUnfoundedCheck::UfsBodyNode* DefaultUnfoundedCheck::addBody(PrgBodyNode* 
 /////////////////////////////////////////////////////////////////////////////////////////
 bool DefaultUnfoundedCheck::findUnfoundedSet() {
 	// first: remove all sources that were recently falsified
-	if (!sourceQueue_.empty()) propagateSource();
-	for (VarVec::size_type i = 0; i < invalid_.size(); ++i) { 
+	if (!sourceQueue_.empty()) {
+		propagateSource();
+	}
+	for (VarVec::size_type i = 0; i != invalid_.size(); ++i) { 
 		removeSource(invalid_[i]); 
 	}
 	invalid_.clear();
@@ -512,7 +513,11 @@ bool DefaultUnfoundedCheck::findSource(UfsAtomNode* head) {
 // POST: source(head) != 0
 void DefaultUnfoundedCheck::setSource(UfsBodyNode* body, UfsAtomNode* head) {
 	assert(!solver_->isFalse(body->lit));
-	if (!head->hasSource()) {
+	// For normal rules from not false B follows not false head, but
+	// for choice rules this is not the case. Therefore, the 
+	// check for isFalse(head) is needed so that we do not inadvertantly
+	// source a head that is currently false.
+	if (!head->hasSource() && !solver_->isFalse(head->lit)) {
 		head->updateSource(body);
[…]

### 2008-08-05 lib/gringo/include/smodelsconverter.h

@@ -57,9 +57,10 @@ namespace NS_GRINGO
 			void handleSum(bool body, Aggregate *a, int &l, int &u);
 			void handleMin(Aggregate *a, int &l, int &u);
 			void handleMax(Aggregate *a, int &l, int &u);
+		protected:
+			int false_;
 		private:
 			bool negBoundsWarning_;
-			int false_;
 			IntVector compute_;
 			IntVector head_;
 			IntVector pos_, neg_, wPos_, wNeg_;

### 2008-08-05 lib/gringo/src/claspoutput.cpp

@@ -17,7 +17,15 @@
 
 #include "claspoutput.h"
 
+//#define DEBUG_ICLASP
+
+#ifdef DEBUG_ICLASP
+#include <fstream>
+std::fstream g_out("iclasp.h", std::ios_base::out | std::ios_base::trunc);
+#endif
+
 #ifdef WITH_CLASP
+
 
 #include "gringoexception.h"
 #include "grounder.h"
@@ -46,6 +54,15 @@ void ClaspOutput::printBasicRule(int head, const IntVector &pos, const IntVector
 	for(IntVector::const_iterator it = pos.begin(); it != pos.end(); it++)
 		b_->addToBody(*it, true);
 	b_->endRule();
+#ifdef DEBUG_ICLASP
+	g_out << "api.startRule().addHead(t" << head << ")";
+	for(IntVector::const_iterator it = neg.begin(); it != neg.end(); it++)
+		g_out << ".addToBody(t" << *it << ", false)";
+	for(IntVector::const_iterator it = pos.begin(); it != pos.end(); it++)
+		g_out << ".addToBody(t" << *it << ", true)";
+	g_out << ".endRule();" << NL;
+#endif
+
 }
 
 void ClaspOutput::printConstraintRule(int head, int bound, const IntVector &pos, const IntVector &neg)
@@ -62,6 +79,14 @@ void ClaspOutput::printConstraintRule(int head, int bound, const IntVector &pos,
 		b_->addAsNormalRules(r);
 	else
 		b_->addRule(r);
+#ifdef DEBUG_ICLASP
+	g_out << "api.startRule(Clasp::CONSTRAINTRULE, " << bound << ").addHead(t" << head << ")";
+	for(IntVector::const_iterator it = neg.begin(); it != neg.end(); it++)
+		g_out << ".addToBody(t" << *it << ", false)";
+	for(IntVector::const_iterator it = pos.begin(); it != pos.end(); it++)
+		g_out << ".addToBody(t" << *it << ", true)";
+	g_out << ".endRule();" << NL;
+#endif
 }
 
 void ClaspOutput::printChoiceRule(const IntVector &head, const IntVector &pos, const IntVector &neg)
@@ -78,6 +103,16 @@ void ClaspOutput::printChoiceRule(const IntVector &head, const IntVector &pos, c
 		b_->addAsNormalRules(r);
 	else
 		b_->addRule(r);
[…]

### 2008-08-28 doc/documentation/paper.tex
changed some command line options, corrected some mistakes, DOKU: added clingo and iclingo section, added incremental meta predicates.
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@466 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 Therefore it supports binder splitting, backjumping and all other features already implemented in gringo1.0.\cite{gringopaper}
-The goal of this work is to give an overview of the features supported by gringo2.0.
-A broad set of potentialities for writing encodings for logic programs exists.
-Currently, no other grounder supports such a rich featured language.
-gringo2.0 supports several output formats and even a human readable format for debugging purposes.
-It also has an internal interface to the solver clasp \cite{gekanesc07b}, so the output can directly be given to the solver and answer sets can be computed.
-Furthermore gringo2.0 can be used with clasp incrementally. Detailed information about incremental solving can be found here \cite{gekaosscth08a} 
-In this paper, at first, the general restrictions to the input language are stated. Then a brief description of how to use gringo2.0 is given.
+A broad set of potentialities for writing encodings for logic programs exists. Currently, no other grounder supports such a rich featured language.
+The goal of this work is to give an overview of the features supported by gringo2.0 and its derivates.
+Three different packages of gringo exist.
+The first version is gringo2.0, it acts as a grounder and supports several output formats and even a human readable format for debugging purposes.
+This version can ground logic programs and output them to files or pipe them to the solver.
+The second version is called clingo and is a grounder with a built-in solver clasp \cite{gekanesc07b}.
+Output can directly be given to the solver and answer sets can be computed.
+The third package is called iclingo, which stands for ``incremental clingo''.
+In this version the grounder gringo2.0 can be used with clasp incrementally.
+Detailed information about incremental solving can be found here \cite{gekaosscth08a}
+In this paper, at first, the general restrictions to the input language are stated. Then a brief description of how to use gringo2.0, clingo and iclingo is given.
 In the next chapter the supported input language is described in detail and a short introduction of how to write a logic program is given.
@@
 
-\section{Using Gringo2.0}
+\section{Using Gringo}
 \subsection{Command Line Interface}
-gringo2.0 can be started from the command line and has some parameters that can be useful.
-The first parameters are gringo2.0 specific parameters and consist of a number of options and any arbitrary number of filenames of the logic program.
-For the special purpose that gringo2.0 is used with its clasp module a double minus ``\texttt{--}'' is used to separate the clasp parameters from the gringo2.0 parameters.
-If no filename is given, gringo2.0 reads from standard input.
+The gringo tools can be started from the command line and have some parameters that can be useful.
+All programs are fully backward compatible meaning, all options of gringo2.0 are valid in clingo and iclingo, and all options of clingo are valid in iclingo.
+\subsection{Gringo2.0 Options}
+Gringo2.0 can be called from the command line via
 \begin{verbatim}
- ./gringo (gringo option|file)* [--[ clasp options]]
+ gringo (gringo option|file)*
 \end{verbatim}
-\subsection{Options}
+If no filename is given, the standard input is used.
 
-\paragraph{\texttt{--help}}
+\paragraph{\texttt{--help,-h}}
 Gives a short overview of all parameters.
-\paragraph{\texttt{--const c=n}}
+\paragraph{\texttt{--version,-v}}
+Prints the version information and exists.
+\paragraph{\texttt{--verbose,-V}}
+Prints extra information during processing.
+\paragraph{\texttt{--stats}}
+Prints extended statistic information.
+
+\paragraph{\texttt{--text,-t}}
+Parameter ``\texttt{--text}'' invokes the text output method, so the output is printed in human readable format, which is also the input format for gringo2.0.
+The output is variable free and some simplifications on the program were made.
+To convert a variable free program in plain format into a format readable by a solver use the \texttt{--ground} option.
+\paragraph{\texttt{--lparse,-l}}
+This mode prints lparse conform output. For a definition of the output format see \cite{lparseManual}.
+Since some features of gringo2.0 are not natively supported for this output language, new atoms may be introduced to create an equivalent program.\newline
+Sum aggregates with negative weights are handled like weight constraint in lparse for this output format.
+So a transformation as explained in \cite{lparseManual} is used to get rid of the negative weights.
+\paragraph{\texttt{--aspils,-a 1-7}}
+This parameter invokes the new output format for the intermediate language ASPils \cite{gejaosscth08b}.
+The optional number can range from $1$ to $7$ and refers to one of the following normal forms (Simple=1, SModels=2, DisjunctiveLogicProgram=3, CModels=4, CModelsExtended=5, DLV=6, Conglomeration=7).
+% If no normal form is specified, conglomeration(7)  is assumed.
+
+\paragraph{\texttt{--ground,-g}}
+If this parameter is used, the grounder expects an already grounded program as input.
+For converting an already grounded logic program into an output format supported by gringo2.0 this is the right option. This mode saves a lot of memory but does not ground the program.
+\paragraph{\texttt{--const,-c C=v}}
 You can set the value of constants defined in your logic program via the command line of gringo2.0.
-After the parameter ``--const'' you have to give the name of the constant $c$ you want to set followed by a ``$=$'' and the value of that constant.
-The integer value $n$ can even be an arithmetic expression.
+After the parameter ``--const'' you have to give the name of the constant $C$ you want to set followed by a ``$=$'' and the value of that constant.
+The integer value $v$ can even be an arithmetic expression.
 For example:
@@
 \end{verbatim}
-the constant $c$ is replaced by $10$ in the logic program.
+the constant $n$ is replaced by $10$ in the logic program.
+\paragraph{\texttt{--bindersplitting=<arg>}}
+With this option you can enable or disable the bindersplitting used by gringo2.0
+%TODO	Default: yes
+\begin{itemize}
+ \item yes : Enable bindersplitting
+ \item no  : Disable bindersplitting
+\end{itemize}
+\paragraph{\texttt{--ifixed=<num>}}
+An incremental program is required as input. Gringo2.0 grounds the fixed number of steps of the incremental program.
 
-\paragraph{\texttt{--convert}}
-If this parameter is used, the grounder expects an already grounded program as input.
-For converting an already grounded logic program into an output format supported by gringo2.0 this is the right option. This mode saves a lot of memory but does not ground the program.
-\paragraph{\texttt{-l}}
-This mode prints smodels conform output. For a definition of the output format see \cite{lparseManual}.
-Since some features of gringo2.0 are not natively supported for this output language, new atoms may be introduced to create an equivalent program.\newline
-This is the current default output format of gringo2.0.
-Sum aggregates with negative weights are handled like weight constraint in lparse for this output format.
-So a transformation as explained in \cite{lparseManual} is used to get rid of the negative weights.
-\paragraph{\texttt{-c}}
-In this mode, the output is directly parsed to clasp, an ASP solver. For solving logic programs, no intermediate files are produced and the answer sets are printed directly.
+\paragraph{Default Parameters}
+The default command line is ``\texttt{./gringo --lparse}'' to read from standard input and for smodels output.
+
+\subsection{Clingo Options}
+
+Clingo can be called from the command line via
+\begin{verbatim}
+ clingo (clingo option|file)*
+\end{verbatim}
+If no filename is given, the standard input is used.
+Since clingo uses gringo2.0 as a grounder in combination with the solver clasp, all options from gringo2.0 and clasp are valid.
+For a detailed list of the clasp options see the clasp help (also available with \texttt{clingo --help}).
+\paragraph{\texttt{--clasp,-C}}
+In this mode, the output is directly parsed to clasp. For solving logic programs, no intermediate files are produced and the answer sets are printed directly.
 With this option enabled, all parameters for clasp are valid and can be declared in the command line.
-Therefore a double minus ``\texttt{--}'' must be given after all gringo2.0 related parameters are set.
-After them the clasp parameters follow.
 For example:
 \begin{verbatim}
- gringo -c --const n=5 -- --supp-models 3
+ clingo --clasp --const n=5 --supp-models 3
 \end{verbatim}
 Furthermore all restriction on smodels output hold, as clasps internal interface currently reads smodels format.
-\paragraph{\texttt{-p}}
-Parameter ``\texttt{-p}'' invokes the plain output method, so the output is printed in human readable format, which is also the input format for gringo2.0.
-The output is variable free and some simplifications on the program were made.
-To convert the plain format into a format readable by a solver use the \texttt{--convert} option.
-\paragraph{\texttt{-g [1-7]}}
-This parameter invokes the new output format for the intermediate language ASPils \cite{gejaosscth08b}.
-The optional number can range from $1$ to $7$ and refers to one of the following normal forms (Simple=1, SModels=2, DisjunctiveLogicProgram=3, CModels=4, CModelsExtended=5, DLV=6, Conglomeration=7).
-If no normal form is specified, conglomeration(7)  is assumed.
+This is default for clingo.
+\paragraph{\texttt{--solveonly=<arg>}}
+This option can be used to set the working mode of clingo to use it as a solver only.
+\begin{itemize}
+\item yes : Pass input directly to clasp
+\item no : Standard mode
+\end{itemize}
+The default is \texttt{--solveonly=no} to ground the logic program first.
+If the input shall directly be passed to clasped, it has to be in the valid smodels input format.
 
 \paragraph{Default Parameters}
-The default command line is ``\texttt{./gringo -l}'' to read from standard input and for smodels output.
+The default is to read from standard input and search for one model.
+\begin{verbatim}
+ clingo 1 --clasp --solveonly=No
+           --trans-ext=no --eq=5 
+           --lookahead=no --lookback=yes --heuristic=Berkmin
+           --rand-prop=0.0 --randomize=no --rand-watches=yes
+           --restarts=100,1.5 --deletion=3,1.1,3.0
+           --minimize=all --contraction=250 --loops=common
+\end{verbatim}
+
+\subsection{Iclingo Options}
+Iclingo can be called from the command line via
+\begin{verbatim}
+ iclingo (iclingo option|file)*
+\end{verbatim}
+Iclingo is used for incremental solving. For a detailed description see \cite{gekaosscth08a}.
+It uses the grounder gringo2.0 and the solver clasp, and therefore all options if gringo2.0, clingo and clasp are valid.
+Additionally some options for controlling the incremental behavior are implemented.
+\paragraph{\texttt{--incremental,-i}}
+Activates the incremental solving. This is the default for iclingo.
+\paragraph{\texttt{--istats}}
+Prints additional statistics for each incremental step.
+\paragraph{\texttt{--imin=<num>}}
+At least \texttt{num} incremental steps are solved, wether an answer set has been found or not.
+\paragraph{\texttt{--imax=<num>}}
+At most \texttt{num} incremental steps are done. After them iclingo stops the search for answer sets.
+\paragraph{\texttt{--ibase}}
+With this option enabled, iclingo solves only the base program and ignores the incremental and query part of the program.
+This option can be helpful when debugging.
+\paragraph{\texttt{--istop=<arg>}}
+The stop condition during incremental solving be set to ``\texttt{SAT}'' or ``\texttt{UNSAT}'' to stop the search process either if a solution was found (\texttt{SAT} and default) or if no solution was found (\texttt{UNSAT}).
+\paragraph{\texttt{--ilearnt=<arg>}}
+This option influences the behavior of clasp when used incrementally.
+\begin{itemize}
+ \item keep   : Learnt nogoods are kept for the next step while solving.
+ \item forget : Clasp forgets the learnt nogoods after each incremental step.
+\end{itemize}
+As default, clasp forgets the learnt nogoods.
+\paragraph{\texttt{--iheuristic=<arg>}}
+This option sets how to handle heuristic information during incremental solving.
+\begin{itemize}
+ \item keep   : Heuristic information is kept for the next step while solving.
+ \item forget : Clasp forgets the heuristic information after each incremental step.
+\end{itemize}
+As default, clasp keeps the heuristic information gathered during the previous steps.
+\paragraph{Default Parameters}
+The default is to read from standard input and search for one model incrementally.
+\begin{verbatim}
+  iclingo 1 --incremental --istop=SAT --ilearnt=forget --iheuristic=keep
+            --trans-ext=no --eq=5 
+            --lookahead=no --lookback=yes --heuristic=Berkmin
+            --rand-prop=0.0 --randomize=no --rand-watches=yes
+            --restarts=100,1.5 --deletion=3,1.1,3.0
+            --minimize=all --contraction=250 --loops=common
+
+\end{verbatim}
 
@@
 They are all printed to standard error output.
+TODO: Some new warnings and Errors were added lately by Roland
 
@@
 \end{verbatim}
-Use the command line option ``--convert'' to convert a ground program into another format.
+Use the command line option ``--ground'' to convert a ground program into another format.
 For converting already ground programs to an output language a special module has been implemented that uses far less memory.
@@
 
-\subsection{Building Gringo2.0 From Source}
-For compiling gringo2.0, a binary version of ``re2c'' is needed to compile the lexer and the parser.
-TODO: What about lemon.
-As the build environment ``cmake'' is used.
-TODO: How to build gringo2.0, clingo, iclingo etc...
+%\subsection{Building Gringo2.0 From Source}
+%For compiling gringo2.0, a binary version of ``re2c'' is needed to compile the lexer and the parser.
+%TODO: What about lemon.
+%As the build environment ``cmake'' is used.
+%TODO: How to build gringo2.0, clingo, iclingo etc...
 
@@
 Whenever a ``\texttt{\%}'' occurs in the input, the rest of the line is treated as a comment and is therefore omitted.
+\subsubsection{Meta-Syntax for Incremental Programs}
+Incremental programs must be splitted into 3 parts.
+The base part, for the information that hold in all steps.
+The incremental part, where the incremental rules must be set, and the volatile part.
+In this part the queries to the program are given.
+To seperate the part, three statement exists. Everything beyond such a statement is considered to belong to the specific part.
+To start the base part
+\begin{verbatim}
+ #base.
+\end{verbatim}
+must be written. Now all rules of the base part can follow.
+For the incremental part an additional variable has to be passed to the statement.
+\begin{verbatim}
+ #incremental(X).
+\end{verbatim}
+Now all incremental rules follow. The incremental variable used for grounding is $X$.
+The same holds for the volatile part.
+\begin{verbatim}
+ #volatile(X).
+\end{verbatim}
+The ``\#''-sign can not be omitted for these meta statements.
+These statements can also be mixed, only the last statement is important.
+If no statement is given at the beginning of the file, base is assumed.
 
@@
 \end{verbatim}
-If squared brackets are used, the atom list is interpreted as a set, otherwise, for curly brackets, as a multiset of atoms.
+If squared brackets are used, the atom list is interpreted as a multiset, otherwise, for curly brackets, as a set of atoms.
 The lower and upper bound is optional, and can be omitted.

### 2008-08-28 app/options.cpp

@@ -153,8 +153,15 @@ bool mapASPils(const std::string& s, int &out, int*) {
 
 bool mapMode(const std::string& s, bool &out, bool*) {
 	std::string temp = toLower(s);
-	if (temp == "grounder") { out = true;  return true; }
-	if (temp == "solver")   { out = false; return true; }
+	if (temp == "no") { out = true;  return true; }
+	if (temp == "yes")   { out = false; return true; }
+	return false;
+}
+
+bool mapStop(const std::string& s, bool &out, bool*) {
+	std::string temp = toLower(s);
+	if (temp == "unsat") { out = true;  return true; }
+	if (temp == "sat")   { out = false; return true; }
 	return false;
 }
 
@@ -241,10 +248,10 @@ void Options::initOptions(ProgramOptions::OptionGroup& allOpts, ProgramOptions::
 		("verbose,V", bool_switch(&verbose), "Print extra information")
 		("stats"    , bool_switch(&stats),  "Print extended statistics")
 #ifdef WITH_CLASP
-		("mode"     , value<bool>(&grounder)->parser(mapMode), "Set the working mode\n"
-			"\tDefault: grounder\n"
-			"\t  grounder : Standard grounding mode\n"
-			"\t  solver   : Pass input directly to clasp")
+		("solveonly"     , value<bool>(&grounder)->parser(mapMode), "Set the working mode\n"
+			"\tDefault: No\n"
+			"\t  No  : Standard mode\n"
+			"\t  Yes : Pass input directly to clasp (lparse format required)")
 #endif
 	;
 	allOpts.addOptions(common);
@@ -273,14 +280,14 @@ void Options::initOptions(ProgramOptions::OptionGroup& allOpts, ProgramOptions::
 		("imax"        , value<int>(&imax)    , "Maximum number of incremental steps", "<num>")
 		("incremental,i"   , bool_switch(&iclaspOut), "Use incremental clasp interface")
 		("ibase"       , bool_switch(&ibase)  , "(not implemented yet)")
-		("iunsat"      , value<bool>(&iunsat)->defaultValue(false), "Stop condition during incremental solving\n"
-			"\tDefault: no\n"
-			"\t  yes : Stop if no solution found\n"
-			"\t  no  : Stop if solution found")
[…]

### 2008-08-28 lib/gringo/include/statementdependencygraph.h
- made the tarjan iterative
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@468 3e02f596-3131-0410-8352-e04eeaec37cc

@@ -51,7 +51,7 @@ namespace NS_GRINGO
 		~SDG();
 	private:
 		void calcSCCDep(SDGNode *v1, SCC *scc, bool &root);
-		void tarjan(SDGNode *v1, std::stack<SDGNode*> &stack, int &index);
+		void tarjan(SDGNode *v, int &index, int &back, std::vector<SDGNode*> &stack);
 	private:
 		int uids_;
 		SDGNodeVector ruleNodes_;
@@ -78,10 +78,8 @@ namespace NS_GRINGO
 		void addDependency(SDGNode *n, bool neg = false);
 		~SDGNode();
 	private:
-		int lowlink_;
 		int index_;
 		unsigned int type_ : 1;
-		unsigned int stacked_ : 1;
 		unsigned int done_ : 1;
 		SDG::SCC *scc_;
 		union

### 2008-08-28 lib/gringo/src/grounder.cpp

@@ -235,7 +235,7 @@ void Grounder::ground_()
 	for(ProgramVector::iterator it = sccs_.begin(); it != sccs_.end(); it++)
 	{
 		Program *scc = *it;
-		//std::cerr << scc << std::endl;
+		//std::cerr << pp(this, scc) << std::endl;
 		eval_ = scc->getEvaluator();
 		eval_->initialize(this);
 		StatementVector *rules = scc->getStatements();

### 2008-08-28 lib/gringo/src/statementdependencygraph.cpp

@@ -57,55 +57,80 @@ SDGNode *SDG::createPredicateNode(PredicateLiteral *pred)
 	return n;
 }
 
-void SDG::tarjan(SDGNode *v1, std::stack<SDGNode*> &s, int &index)
+namespace
 {
-	// do a depth first search to find the sccs
-	v1->index_ = index;
-	v1->lowlink_ = index;
-	index = index + 1;
-	s.push(v1);
-	v1->stacked_ = true;
-	SDGNodeVector *dep = v1->getDependency();
-	for(SDGNodeVector::iterator it = dep->begin(); it != dep->end(); it++)
+	struct Call
 	{
-		SDGNode *v2 = *it;
-		if(v2->done_)
-			continue;
-		if(v2->index_ == -1)
+		Call(SDGNode *v, SDGNodeVector::iterator w) : v(v), root(true), w(w)
 		{
-			tarjan(v2, s, index);
-			v1->lowlink_ = std::min(v1->lowlink_, v2->lowlink_);
 		}
-		else if (v2->stacked_)
-		{
-			v1->lowlink_ = std::min(v1->lowlink_, v2->index_);
-		}
-	}
-	if (v1->lowlink_ == v1->index_)
+
+		SDGNode *v;
+		bool root;
+		SDGNodeVector::iterator w;
+	};
+}
+
+void SDG::tarjan(SDGNode *v, int &index, int &back, std::vector<SDGNode*> &stack)
+{
+	std::vector<Call> callstack;
+	callstack.push_back(Call(v, v->getDependency()->begin()));
+	v->index_ = index++;
+	while(!callstack.empty())
 	{
-		SCC *scc = new SCC();
-		int nodes = 0;
-		SDGNode *v2;
-		do
+Start:
+		Call &c = callstack.back();
+		v = c.v;
+		for(; c.w != c.v->getDependency()->end(); c.w++)
 		{
-			v2 = s.top();
-			v2->scc_ = scc;
-			if(v2->getStatement())
-				scc->rules_.push_back(v2->getStatement());
-			s.pop();
-			v2->stacked_ = false;
-			nodes++;
+			SDGNode *w = *c.w;
+			if(w->index_ == 0)
+			{
+				callstack.push_back(Call(w, w->getDependency()->begin()));
+				w->index_ = index++;
+				goto Start;
+			}
+			if(w->index_ < v->index_)
+			{
+				v->index_ = (*c.w)->index_;
+				c.root = false;
+			}
 		}
-		while(v1 != v2);
[…]

### 2008-08-28 app/main.cpp
- made gringo more quiet - totally banned recursion from statementdependencygraph   (dependencygraph building)
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@469 3e02f596-3131-0410-8352-e04eeaec37cc

@@ -209,7 +209,8 @@ int MainApp::run(int argc, char **argv)
 	}
 
 #ifdef WITH_CLASP
-	cerr << (mode == ASP_MODE ? "" : "c ") << EXECUTABLE << " version " << GRINGO_VERSION << " (clasp " << CLASP_VERSION << ")" << endl;
+	if(options.verbose)
+		cerr << (mode == ASP_MODE ? "" : "c ") << EXECUTABLE << " version " << GRINGO_VERSION << " (clasp " << CLASP_VERSION << ")" << endl;
 	getStream();
 	if(options.files.size() > 0)
 	{
@@ -227,7 +228,8 @@ int MainApp::run(int argc, char **argv)
 	else if(options.verbose)
 		cerr << (mode == ASP_MODE ? "" : "c ") << "Reading from stdin" << endl;
 #else
-	cerr << EXECUTABLE << " version " << GRINGO_VERSION << endl;
+	if(options.verbose)
+		cerr << EXECUTABLE << " version " << GRINGO_VERSION << endl;
 	if(options.files.size() > 0)
 	{
 		vector<string>::iterator i = options.files.begin();

### 2008-08-28 lib/gringo/include/statementdependencygraph.h

@@ -50,7 +50,7 @@ namespace NS_GRINGO
 		void calcSCCs(Grounder *g);
 		~SDG();
 	private:
-		void calcSCCDep(SDGNode *v1, SCC *scc, bool &root);
+		bool calcSCCDep(SDGNode *v);
 		void tarjan(SDGNode *v, int &index, int &back, std::vector<SDGNode*> &stack);
 	private:
 		int uids_;

### 2008-08-28 lib/gringo/src/statementdependencygraph.cpp

@@ -98,18 +98,17 @@ Start:
 		}
 		if(c.root)
 		{
-			SCC *scc = new SCC();
 			int nodes = 1;
-			v->scc_ = scc;
+			v->scc_ = new SCC();
 			if(v->getStatement())
-				scc->rules_.push_back(v->getStatement());
+				v->scc_->rules_.push_back(v->getStatement());
 			index--;
 			while(!stack.empty() && v->index_ <= stack.back()->index_)
 			{
 				stack.back()->index_ = back;
-				stack.back()->scc_ = scc;
+				stack.back()->scc_ = v->scc_;
 				if(stack.back()->getStatement())
-					scc->rules_.push_back(stack.back()->getStatement());
+					v->scc_->rules_.push_back(stack.back()->getStatement());
 				stack.pop_back();
 				index--;
 				nodes++;
@@ -117,14 +116,12 @@ Start:
 			v->index_ = back;
 			back--;
 			// initialize with fact or basic program
-			scc->type_ = nodes == 1 ? SCC::FACT : SCC::BASIC;
+			v->scc_->type_ = nodes == 1 ? SCC::FACT : SCC::BASIC;
 			//scc->type_ = nodes == 1 ? SCC::FACT : SCC::NORMAL;
-			sccs_.push_back(scc);
-			bool root = true;
+			sccs_.push_back(v->scc_);
 			// calc type and dependency of program
-			calcSCCDep(v, scc, root);
-			if(root)
-				sccRoots_.insert(scc);
+			if(calcSCCDep(v))
+				sccRoots_.insert(v->scc_);
 		}
 		else
 		{
@@ -134,41 +131,51 @@ Start:
 	}
 }
 
-void SDG::calcSCCDep(SDGNode *v1, SCC *scc, bool &root)
+bool SDG::calcSCCDep(SDGNode *v)
 {
+	bool root = true;
+	std::vector<SDGNode*> bfs;
+	bfs.push_back(v);
+
 	// do a depth first search limited to the scc to build a tree of sccs
-	v1->done_ = true;
-	SDGNodeVector *dep = v1->getDependency();
-	// build tree of sccs
-	for(SDGNodeVector::iterator it = dep->begin(); it != dep->end(); it++)
+	while(!bfs.empty())
 	{
-		SDGNode *v2 = *it;
-		assert(v2->scc_);
[…]

### 2008-09-01 lib/gringo/include/conjunctionaggregate.h
- improved the sokoban encoding   its more readable now - fixed a bug ignoring not's in conjunctive aggregates
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@477 3e02f596-3131-0410-8352-e04eeaec37cc

@@ -28,6 +28,7 @@ namespace NS_GRINGO
 	public:
 		ConjunctionAggregate(ConditionalLiteralVector *literals);
 		ConjunctionAggregate(const ConjunctionAggregate &a);
+		virtual void setNeg(bool neg);
 		virtual Literal *clone() const;
 		virtual bool match(Grounder *g);
 		virtual void match(Grounder *g, int &lower, int &upper, int &fixed);

### 2008-09-01 lib/gringo/include/disjunctionaggregate.h

@@ -28,6 +28,7 @@ namespace NS_GRINGO
 	public:
 		DisjunctionAggregate(ConditionalLiteralVector *literals);
 		DisjunctionAggregate(const DisjunctionAggregate &a);
+		virtual void setNeg(bool neg);
 		virtual Literal *clone() const;
 		virtual void preprocess(Grounder *g, Expandable *e);
 		virtual bool match(Grounder *g);

### 2008-09-01 lib/gringo/include/literal.h

@@ -44,12 +44,12 @@ namespace NS_GRINGO
 		 * \brief Sets the sign of the literal
 		 * \param neg The new sign
 		 */
-		void setNeg(bool neg);
+		virtual void setNeg(bool neg);
 		/**
 		 * \brief Retrieve the sign ofthe literal.
 		 * \return Returns the sign
 		 */
-		bool getNeg() const;
+		virtual bool getNeg() const;
 		/**
 		 * \brief Retrieve the vars of the literal
 		 * \param vars Reference to a set used to store the result

### 2008-09-01 lib/gringo/src/conjunctionaggregate.cpp

@@ -29,6 +29,13 @@ ConjunctionAggregate::ConjunctionAggregate(ConditionalLiteralVector *literals) :
 {
 }
 
+void ConjunctionAggregate::setNeg(bool neg)
+{
+	// the literal itself will never be negative
+	for(ConditionalLiteralVector::iterator it = literals_->begin(); it != literals_->end(); it++)
+		(*it)->setNeg(neg);
+}
+
 bool ConjunctionAggregate::match(Grounder *g)
 {
 	fact_ = true;

### 2008-09-01 lib/gringo/src/disjunctionaggregate.cpp

@@ -28,6 +28,13 @@ DisjunctionAggregate::DisjunctionAggregate(ConditionalLiteralVector *literals) :
 {
 }
 
+void DisjunctionAggregate::setNeg(bool neg)
+{
+	// the literal itself will never be negative
+	for(ConditionalLiteralVector::iterator it = literals_->begin(); it != literals_->end(); it++)
+		(*it)->setNeg(neg);
+}
+
 bool DisjunctionAggregate::match(Grounder *g)
 {
 	fact_ = false;

### 2008-09-01 app/main.cpp
- fixed the timer - reset the solve statistics after each step
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@478 3e02f596-3131-0410-8352-e04eeaec37cc

@@ -615,6 +615,14 @@ bool MainApp::solveIncremental()
 	solver.strategies().heuristic->reinit(options.keepHeuristic);
 	grounder.prepare(true);
 	setState(end_read);
+	
+	struct SStats
+	{
+		uint64 restarts;
+		uint64 choices;
+		uint64 conflicts;
+		SStats() : restarts(0), choices(0), conflicts(0) {}
+	} sstats;
 
 	struct IStats
 	{
@@ -624,9 +632,7 @@ bool MainApp::solveIncremental()
 		double tpre;
 		double tsolve;
 		uint32 rules;
-		uint64 choices;
-		uint64 conflicts;
-		IStats() : models(0), ttotal(0), tground(0), tpre(0), tsolve(0), rules(0), choices(0), conflicts(0) {}
+		IStats() : models(0), ttotal(0), tground(0), tpre(0), tsolve(0), rules(0) {}
 		void print(uint32 models, double ttotal, double tground, double tpre, double tsolve, uint32 rules, uint64 choices, uint64 conflicts)
 		{
 			cerr << endl;
@@ -636,23 +642,27 @@ bool MainApp::solveIncremental()
 				<< ", p: " << (tpre - this->tpre) 
 				<< ", s: " << (tsolve - this->tsolve) << ")" << endl;
 			cerr << "Rules    : " << (rules - this->rules) << endl;
-			cerr << "Choices  : " << (choices - this->choices) << endl;
-			cerr << "Conflicts: " << (conflicts - this->conflicts) << endl;
+			cerr << "Choices  : " << choices << endl;
+			cerr << "Conflicts: " << conflicts << endl;
 			this->models    = models;
 			this->ttotal    = ttotal;
 			this->tground   = tground;
 			this->tpre      = tpre;
 			this->tsolve    = tsolve;
 			this->rules     = rules;
-			this->choices   = choices;
-			this->conflicts = conflicts;
 		}
 	} istats;
+
+	StdOutPrinter printer;
+	if(!options.quiet)
+	{
+		printer.index = &api.stats.index;
+		options.solveParams.enumerator()->setPrinter(&printer);
+	}
 
 	CTimer all;
 	do 
 	{
-		//solver.stats = SolverStatistics();
 		all.Start();
 		setState(start_ground);
 		if(options.verbose || options.istats)
@@ -675,12 +685,6 @@ bool MainApp::solveIncremental()
 			if(options.verbose)
 				cerr << "Solving..." << endl;
[…]

### 2008-09-01 app/timer.cpp

@@ -53,12 +53,12 @@ void CTimer::Stop () {
   _sec+= stop.tv_sec - _start.tv_sec;
   if(stop.tv_usec >= _start.tv_usec)
   {
-    _usec+= stop.tv_usec - _start.tv_usec;
+  	_usec+= stop.tv_usec - _start.tv_usec;
   }
   else
   {
-    _usec+= _start.tv_usec - stop.tv_usec;
-    _sec--;
+  	_usec+= 1000000 - _start.tv_usec + stop.tv_usec;
+	_sec--;
   }
   if(_usec >= 1000000)
   {

### 2008-09-01 app/main.cpp
- using a more aggressive exit in the sigHandler now :)
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@479 3e02f596-3131-0410-8352-e04eeaec37cc

@@ -153,7 +153,7 @@ static void sigHandler(int)
 #else
 	printf("\n*** INTERRUPTED! ***\n");
 #endif
-	exit(S_UNKNOWN);
+	_exit(S_UNKNOWN);
 } 
 
 void MainApp::setState(State s)

### 2008-09-01 app/main.cpp
- fixed a bug resetting the wrong values
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@480 3e02f596-3131-0410-8352-e04eeaec37cc

@@ -703,9 +703,9 @@ bool MainApp::solveIncremental()
 		sstats.restarts += solver.stats.restarts;
 		sstats.choices  += solver.stats.choices;
 		sstats.conflicts+= solver.stats.conflicts;
-		sstats.restarts  = sstats.choices = sstats.conflicts = 0;
+		solver.stats.restarts = solver.stats.choices = solver.stats.conflicts = 0;
 	}
-	while(options.imax-- > 1 &&(options.imin-- > 1 || ret == options.iunsat));
+	while(options.imax-- > 1 && (options.imin-- > 1 || ret == options.iunsat));
 	// for the summary
 	*lpStats_ = output.getStats();
 	api.stats.moveTo(*preStats_);

### 2008-09-01 doc/documentation/paper.tex
DOKU: added missing aggregate
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@467 3e02f596-3131-0410-8352-e04eeaec37cc
added a description of how to write an incremental program to the docu, many things still have to be done
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@481 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 
+\subsection{How to Write an Incremental Program}
+As in writing a non incremental logic problem, we describe a problem rather than an algorithm.
+The difference between a non incremental and an incremental program is that we can use an incremental constant.
+This constant is set to zero at the beginning and the solver tries to find a solution of the problem.
+If it can not find a solution it starts to increment the constant by one.
+At each step the solver searches for a solution to the problem and increments the constant until he finds a valid answer set for the logic program.
+The incremental constant can therefore be used among other things to emulate time steps.
+For example, if the solution for a planning problem can not be found in 3 steps, it is tried to find one in more steps.
+To make it concrete, an incremental encoding for the blocksworld problem is presented.
+This is a very simple problem, where $n$ blocks are stacked on top of each other and you can move only one block at the same time. The goal is to stack the blocks in the inverse order.
+Since an incremental logic program is divided into three parts, we start with the base part.
+All rules that hold in all timesteps that are independent of the incremental constant are put here.
+At first we declare $n$ blocks and that each block is a location where a block can be put on.
+Furthermore we define an extra location, the table. That is the location where the bottom most block is put.
+\begin{verbatim}
+ #base.
+
+ block(1..n).
+ location(X) :- block(X).
+ location(table).
+\end{verbatim}
+Now we stack all blocks on top of each other, so block 1 is on block 2, block 2 is on block 3 etc.
+The last block we put on the table.
+This all holds in the initial state, so a $0$ is appended to the ``\texttt{on}'' predicate.
+\begin{verbatim}
+ on(X,X+1,0) :- block(X), X < n.
+ on(n,table,0).
+\end{verbatim}
+After declaring the base part and describing the initial situation we have to describe valid moves for the blocks.
+This is done in the cumulative part, since moves can be done in each step.
+For the incremental constant we choose the name ``\texttt{k}''.
+At first we want to execute at least one move every step, but not more than one. This can be done with a cardinality constraint.
+\begin{verbatim}
+ #cumulative k.
+
+ 1{move(X,Y,k) : block(X) : location(Y) : X != Y}1
+\end{verbatim}
+A move is of the form ``\texttt{move(X,Y,k)}'' where \texttt{X} is the block that is moved, \texttt{Y} the new position of the block and \texttt{k} the step when the moved is executed.
+\newline
+Now that we have exactly one move per step we need to define what happens if we move a block.
+Understandably if we move block \texttt{X} to position \texttt{Y} in step \texttt{k}, the new location of the block shall be \texttt{Y}.
+\begin{verbatim}
+ on(X,Y,k) :- move(X,Y,k), block(X), location(Y).
+\end{verbatim}
+For all other blocks that have not moved, the position shall be the same.
+In ASP a little trick is used to describe such a condition.
+We first define a little helper predicate that states that a block is not on a certain position.
+\begin{verbatim}
+ not_on(X,Y,k) :- on(X,Z,k), Y != Z, block(X), location(Y;Z).
+\end{verbatim}
+Now block \texttt{X} is ``\texttt{not\_on}'' a position, where it is not ``\texttt{on}''.
+With this helping predicate we can define that a block is on the same position in step \texttt{k}, if it was not ``\texttt{not\_on}'' that position in the step before.
+\begin{verbatim}
+ on(X,Y,k) :- on(X,Y,k-1), block(X), location(Y), not not_on(X,Y,k).
+\end{verbatim}
+This trick is often used to describe the state of something that has not changed by an action.
+\newline
+Now we have to restrict our moves, because only one block can be moved at the same time and it can only be moved if no other block is on top of it.
+To restrict the moves, an integrity constraint is used.
+\begin{verbatim}
+ :- move(X,Y,k), on(Z,X,k-1), block(X;Z), location(Y).
+\end{verbatim}
+So it is not allowed to move block \texttt{X} in step k, if there is another block on it in the state before.
+Also you can not move a block to a position where another block is on, so only one block on top of another is allowed.
+\begin{verbatim}
+ :- move(X,Y,k), on(Z,Y,k-1), block(X;Y;Z).
+\end{verbatim}
+The description of the problem is nearly finished. All that has to be done is to formulate a query.
+As a constraint to our solution we want to stop if all blocks are on top of each other in reversed order.
+This query shall only hold in the last step, therefore we write it in the volatile part.
+As always for constraints, we use an integrity constraint to express that all solutions are false where there is not block $n+1$ on the block $n$.
+\begin{verbatim}
+ #volatile k.
+
+ :- not on(X+1,X,k), block(X), X < n.
+\end{verbatim}
+To run this example, iclingo has to be used and the constant $n$ has to be set via command line.
+A potential command line call could look like
+\begin{verbatim}
+ iclingo --const n=10 blocksworld.lp
+\end{verbatim}
+and the answer to the problem can be read from the answer set.
+
+%  1 #hide.
+%  2 #show mv(X,Y,Z).
+%  3
+%  4 #base.
+%  5
+%  6 bl(1..n).
+%% Es gibt "n" Blöcke
+%  7 loc(t).
+%  8 loc(X) :- bl(X).
+%% und die Positionen t=table, 1 ... n
+%  9 
+%% Block 1 ist auf Block 2, 2 auf 3 ... und n auf dem Tisch
+% 11
+% 12 #cumulative k.
+% 13
+% 14 1{mv(X,Y,k) : bl(X) : loc(Y) : X != Y}1.
+%% Wähle jeden Zug einen move von Block X nach Y aus
+% 15 on(X,Y,k) :- mv(X,Y,k), bl(X), loc(Y).
+%% Ein Block X ist auf der Position Y (im Schritt k)falls block X nach Y bewegt wird(in Schritt k).
+% 16 on(X,Y,k) :- on(X,Y,k-1), bl(X), loc(Y), not non(X,Y,k).
+% 17 non(X,Y,k) :- on(X,Z,k), Y != Z, bl(X), loc(Y;Z).
+%% Ein Bock X ist auf Position Y (in Schritt k) wenn er nicht, nicht dort lag.
+%% non(X,Y) gilt im Schritt k, falls X NICHT auf Y liegt.
+% 18 :- mv(X,Y,k), on(Z,X,k-1), bl(X;Z), loc(Y).
+%% move von X nach Y (im Schritt k) ist falsch, falls im vorherigen Schritt ein Block auf X lag. ???
+% 19 :- mv(X,Y,k), on(Z,Y,k-1), bl(X;Y;Z).
+%% move von X nach Y (im Schritt k) ist falsch, falls im vorherigen Schritt ein Block auf Y lag
+% 20
+% 21 #volatile k.
+% 22
+% 23 :- not on(X+1,X,k), bl(X), X < n.
+%% solange keine Antwort, bis n auf n-1 ... 2 auf 1 liegt.
+% 24
+% 25 #base.
+
 \subsection{Meta-Syntax}
@@
 Incremental programs must be splitted into 3 parts.
-The base part, for the information that hold in all steps.
-The incremental part, where the incremental rules must be set, and the volatile part.
-In this part the queries to the program are given.
-To seperate the part, three statement exists. Everything beyond such a statement is considered to belong to the specific part.
+The base part, for the information that hold in all steps ans is independent of the incremental constant.
+The cumulative part, where the cumulative rules must be set that hold in all steps, but depend on the incremental parameter.
+And finally the volatile part. In this part the queries to the program are given, that must hold in the last state.
+To seperate the parts, three statement exists. Everything beyond such a statement is considered to belong to the specific part.
 To start the base part
@@
 must be written. Now all rules of the base part can follow.
-For the incremental part an additional variable has to be passed to the statement.
+The use of the incremental constant is not allowed in this part.
+For the cumulative part an additional constant has to be passed to the statement.
 \begin{verbatim}
- #incremental(X).
+ #cumulative k.
 \end{verbatim}
-Now all incremental rules follow. The incremental variable used for grounding is $X$.
+Now all incremental rules follow. The incremental constant used for grounding is $k$.
 The same holds for the volatile part.
 \begin{verbatim}
- #volatile(X).
+ #volatile k.
 \end{verbatim}
@@
 This operator can be used in aggregates to create a comma separated list of atoms.
+For example
+\begin{verbatim}
+ n(1..9).
+ 1{a(X) : n(X) : X > 3}1.
+\end{verbatim}
+would become
+\begin{verbatim}
+ 1 {a(4), a(5), a(6), a(7), a(8), a(9)} 1.
+\end{verbatim}
 
@@
 \end{verbatim}
+Furthermore the assignment operator can be used with aggregates, for example
+\begin{verbatim}
+ price(apple,5). prices(banana, 6). price(computer, 100).
+ fruit(apple).
+ fruit(banana).
+ max_cost(1..200).
+ costs(C) :- C = [price(X,Y) : fruit(X) : max_cost(Y) = Y].
+\end{verbatim}
+would lead to the fact \texttt{costs(11)} as the price of all fruits.
+$C$ in this example becomes all values that have to be generated for the aggregate.
 

### 2008-09-02 doc/documentation/paper.tex
Doku: changed blocksworld example to rolands new version
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@483 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 
- 1{move(X,Y,k) : block(X) : location(Y) : X != Y}1
+ 1{move(X,Y,k) : block(X) : location(Y) : X != Y}1.
 \end{verbatim}
@@
 \begin{verbatim}
- on(X,Y,k) :- move(X,Y,k), block(X), location(Y).
+ on(X,Y,k) :- move(X,Y,k).
 \end{verbatim}
@@
 In ASP a little trick is used to describe such a condition.
-We first define a little helper predicate that states that a block is not on a certain position.
+We first define a little helper predicate that states that a block has moved.
 \begin{verbatim}
- not_on(X,Y,k) :- on(X,Z,k), Y != Z, block(X), location(Y;Z).
+ moved(X, k) :- move(X, Y, k).
 \end{verbatim}
-Now block \texttt{X} is ``\texttt{not\_on}'' a position, where it is not ``\texttt{on}''.
-With this helping predicate we can define that a block is on the same position in step \texttt{k}, if it was not ``\texttt{not\_on}'' that position in the step before.
+With this helping predicate we can define that a block is on the same position in step \texttt{k}, if it was not moved.
 \begin{verbatim}
- on(X,Y,k) :- on(X,Y,k-1), block(X), location(Y), not not_on(X,Y,k).
+ on(X,Y,k) :- on(X,Y,k-1), block(X), location(Y), not moved(X, k).
 \end{verbatim}
-This trick is often used to describe the state of something that has not changed by an action.
-\newline
+
 Now we have to restrict our moves, because only one block can be moved at the same time and it can only be moved if no other block is on top of it.
-To restrict the moves, an integrity constraint is used.
+First we introduce a predicate called ``\texttt{blocked}'' for each block with another block on top of it.
 \begin{verbatim}
- :- move(X,Y,k), on(Z,X,k-1), block(X;Z), location(Y).
+ blocked(Y, k) :- block(Y), on(X, Y, k - 1).
 \end{verbatim}
-So it is not allowed to move block \texttt{X} in step k, if there is another block on it in the state before.
+Now we can restrict the moves with an integrity constraint.
+\begin{verbatim}
+ :- move(X, Y, k), blocked(X, k).
+\end{verbatim}
+So it is not allowed to move block \texttt{X} in step k, if it is blocked.
 Also you can not move a block to a position where another block is on, so only one block on top of another is allowed.
 \begin{verbatim}
- :- move(X,Y,k), on(Z,Y,k-1), block(X;Y;Z).
+ :- move(X, Y, k), blocked(Y, k).
 \end{verbatim}
@@
 and the answer to the problem can be read from the answer set.
-
-%  1 #hide.
-%  2 #show mv(X,Y,Z).
-%  3
-%  4 #base.
-%  5
-%  6 bl(1..n).
-%% Es gibt "n" Blöcke
-%  7 loc(t).
-%  8 loc(X) :- bl(X).
-%% und die Positionen t=table, 1 ... n
-%  9 
-%% Block 1 ist auf Block 2, 2 auf 3 ... und n auf dem Tisch
-% 11
-% 12 #cumulative k.
-% 13
-% 14 1{mv(X,Y,k) : bl(X) : loc(Y) : X != Y}1.
-%% Wähle jeden Zug einen move von Block X nach Y aus
-% 15 on(X,Y,k) :- mv(X,Y,k), bl(X), loc(Y).
-%% Ein Block X ist auf der Position Y (im Schritt k)falls block X nach Y bewegt wird(in Schritt k).
-% 16 on(X,Y,k) :- on(X,Y,k-1), bl(X), loc(Y), not non(X,Y,k).
-% 17 non(X,Y,k) :- on(X,Z,k), Y != Z, bl(X), loc(Y;Z).
-%% Ein Bock X ist auf Position Y (in Schritt k) wenn er nicht, nicht dort lag.
-%% non(X,Y) gilt im Schritt k, falls X NICHT auf Y liegt.
-% 18 :- mv(X,Y,k), on(Z,X,k-1), bl(X;Z), loc(Y).
-%% move von X nach Y (im Schritt k) ist falsch, falls im vorherigen Schritt ein Block auf X lag. ???
-% 19 :- mv(X,Y,k), on(Z,Y,k-1), bl(X;Y;Z).
-%% move von X nach Y (im Schritt k) ist falsch, falls im vorherigen Schritt ein Block auf Y lag
-% 20
-% 21 #volatile k.
-% 22
-% 23 :- not on(X+1,X,k), bl(X), X < n.
-%% solange keine Antwort, bis n auf n-1 ... 2 auf 1 liegt.
-% 24
-% 25 #base.
+To clean up the answer the ``\texttt{show} and \texttt{hide}'' feature of gringo can be used.
+If the following lines were added to the program, the answer set only consists of \texttt{move/3} predicates.
+\begin{verbatim}
+ #hide.
+ #show move/3.
+\end{verbatim}
 

### 2008-09-03 doc/documentation/paper.tex
Doku: Added intuitive description of lamdba-restrictedness, removed examples, restructured whole document
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@485 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 The syntax of the gringo2.0 input language can be found in the appendix \ref{appendix}.
-
-\section{$\lambda$-Restriction}
-\label{domainrestricted}
-There exist two major grounders for ASP besides gringo1.0, lParse \cite{lparseManual} and  the DLV grounding module\cite{DLV}.
-To ensure the existence of an equivalent ground instantiation of a program both grounders except a subclass of logic programs.
-For lParse the logic programs must be $\omega$-restricted\cite{omega} while DLV can handle safe programs\cite{safe}.
-gringo2.0 verifies that the input program is $\lambda$-
-restricted \cite{gringopaper}, so that the existence of a finite equivalent ground instantiation is guaranteed. 
-$\lambda$-restrictedness extends the capabilities of $\omega$-restricted programs.
-The definition of $\lambda$-restrictedness can be found in \cite{gringopaper} and was expanded for aggregates and disjunctions.
-$\lambda$-restrictedness can not be compared with the safe programs of DLV since there exist programs that are safe and not $\lambda$-restricted and vice versa.
 
@@
 
+\subsection{$\lambda$-Restriction}
+\label{domainrestricted}
+There exist two major grounders for ASP besides gringo1.0, lParse \cite{lparseManual} and  the DLV grounding module\cite{DLV}.
+To ensure the existence of an equivalent ground instantiation of a program both grounders except a subclass of logic programs.
+For lParse the logic programs must be $\omega$-restricted\cite{omega} while DLV can handle safe programs\cite{safe}.
+gringo2.0 verifies that the input program is $\lambda$-
+restricted \cite{gringopaper}, so that the existence of a finite equivalent ground instantiation is guaranteed. 
+$\lambda$-restrictedness extends the capabilities of $\omega$-restricted programs.
+The definition of $\lambda$-restrictedness can be found in \cite{gringopaper} and was expanded for aggregates and disjunctions.
+$\lambda$-restrictedness can not be compared with the safe programs of DLV since there exist programs that are safe and not $\lambda$-restricted and vice versa.
+\newline
+To give a gross impression of what $\lambda$-restriction means for writing a logic program an informal description is given.
+In gringo2.0 every variable used in a rule must have a complete domain.
+This means that this variable has to occur in at least one predicate whichs domain is completed.
+Therefore the predicate must be in the head of rules that are already domain restricted.
+So the grounder has to find a valid order of the rules so that every predicate in the head is completed before it is used as a domain predicate for the next rules.
+To make things more clear, an example is given
+\begin{verbatim}
+ zig(0) :- not zag(0).   zag(0) :- not zig(0).
+ zig(1) :- not zag(1).   zag(1) :- not zig(1).
+ zigzag(X,Y) :- zig(X), zag(Y).
+\end{verbatim}
+The rules of this example can be ordered in the following way.
+At first, the rules with \texttt{zig/1} and \texttt{zag/1} are considered.
+Since these are the only rules \texttt{zig/1} and \texttt{zag/1} occurs in the head, the domains of the predicates are closed.
+The domains \{\texttt{zig(0)}, \texttt{zig(1)}\} and \{ \texttt{zag(0)}, \texttt{zag(1)}\} are complete, since no other predicates of the form \texttt{zig/1} or \texttt{zag/1} can be derived from the program.
+Now the third rule with \texttt{zigzag/2} in the head can be grounded.
+The variables \texttt{X} and \texttt{Y} both occur in a domain predicate, and the values $0$ respectively $1$ can substitute the variables.
+\newline
+To give a comparison to the $\omega$-restricted programs of lparse, an extra domain predicate has to be introduced to the example.
+\begin{verbatim}
+ domain(0..1).
+ zig(0) :- not zag(0).   zag(0) :- not zig(0).
+ zig(1) :- not zag(1).   zag(1) :- not zig(1).
+ zigzag(X,Y) :- zig(X), zag(Y), domain(X), domain(Y).
+\end{verbatim}
+\texttt{zig/1} and \texttt{zag/1} can not be used as a domain predicate, since their truth value can not be fully evaluated.
+
 %\paragraph{Function Symbols}
@@
 %\end{verbatim}
-
-\subsection{How to Write a Logic Program}
-Writing a logic program to solve a problem is rather describing a problem than describing an algorithm to solve a problem.
-Therefore the problem must be encoded as a set of rules.
-Typically these are divided into a problem description and an instance of the problem.
-The solution to a problem can then be read from the answer sets of the program.
-The procedure of writing a logic program shall now be explained with a little example.
-We want to write a problem encoding for the 3-colorability problem of graph coloring.
-Given an arbitrary graph, no two connected edges may have the same color.
-The most common technique is the ``generate \& test'' approach.
-Therefore, in the generator, rules are created that can derive a superset of the solution to the problem.
-In the tester all solutions that do not fit the constraints are omitted.
-\newline
-Usually you start with the with the generator.
-At first, we define the non instance domains, in our example the colors we use.
-\begin{verbatim}
- color(red). color(green). color(blue).
-\end{verbatim}
-Now we have to color all the vertices, but we must give attention that each vertex has only one color.
-We can do this by simply saying that a vertex is colored with one color $C$, and that the same vertex does not have any of the other colors $D$ or $E$.
-\begin{verbatim}
- colored(V,C) :- not colored(V,D), not colored(V,E),
-                 C != D, C != E, D != E,
-                 color(E), color(C), color(D), vertex(V).
-\end{verbatim}
-As you can see, you have to give the domain of the variables in the rule.
-The variable $V$ is used for vertices and the variables $C$, $D$ and $E$ are used for colors.
-After every vertex now has a color we would get all possible solutions of all vertices colorfully painted.
-We now need to restrict our solutions in the tester by saying that two vertices that are connected by an edge should not have the same color.
-We can do this by using an integrity constraint.
-An integrity constraint is a rule with an empty head, and if an assignment of variables fulfills the body, the assignment is not a solution of our problem.
-\begin{verbatim}
- :- edge(V,U), colored(V,C), colored(U,C),
-    vertex(V), vertex(U), color(C).
-\end{verbatim}
-Here we have the edge between two vertices $V$ and $U$, and if both are colored with the same color $C$ the assignment is not a solution to our problem.
-\newline
-We have finally written our problem description, all we need now is a problem instance. This step is fairly simple.
-We just need to define several vertices and edges that do connect the vertices.
-For the following simple graph
-\[
- \xymatrix{1\ar@{-}[dr]\ar@{-}[d]\ar@{-}[r]&2\ar@{-}[dl]\\
-           3&4}
-\]
-
-we could just write
-\begin{verbatim}
- vertex(1..4).
- edge(1,2). edge(1,4).
- edge(3,1). edge(3,2).
-\end{verbatim}
-
-If we give the problem description and the instance to the grounder, it will generate a variable free version of the problem.
-This can then be solved by a solver and in the answer sets of the solver the ``\texttt{colored}'' predicate states a valid coloring of all vertices.
-We can do this by simply using the internal interface to clasp, to get a solution.
-\begin{verbatim}
- ./gringo -c 3colorability.lp instance1.lp
-\end{verbatim}
-In the first answer set we find the ``\texttt{colored}'' predicates
-\begin{verbatim}
- colored(1,green) colored(2,red) colored(3,blue) colored(4,red)
-\end{verbatim}
-and we can interpret them as the colors of the vertices
-
-\[
- \xymatrix{\color{green}1\ar@{-}[dr]\ar@{-}[d]\ar@{-}[r]&\color{red}2\ar@{-}[dl]\\
-           \color{blue}3&\color{red}4}
-\]
-
-\subsection{How to Write an Incremental Program}
-As in writing a non incremental logic problem, we describe a problem rather than an algorithm.
-The difference between a non incremental and an incremental program is that we can use an incremental constant.
-This constant is set to zero at the beginning and the solver tries to find a solution of the problem.
-If it can not find a solution it starts to increment the constant by one.
-At each step the solver searches for a solution to the problem and increments the constant until he finds a valid answer set for the logic program.
-The incremental constant can therefore be used among other things to emulate time steps.
-For example, if the solution for a planning problem can not be found in 3 steps, it is tried to find one in more steps.
-To make it concrete, an incremental encoding for the blocksworld problem is presented.
-This is a very simple problem, where $n$ blocks are stacked on top of each other and you can move only one block at the same time. The goal is to stack the blocks in the inverse order.
-Since an incremental logic program is divided into three parts, we start with the base part.
-All rules that hold in all timesteps that are independent of the incremental constant are put here.
-At first we declare $n$ blocks and that each block is a location where a block can be put on.
-Furthermore we define an extra location, the table. That is the location where the bottom most block is put.
-\begin{verbatim}
- #base.
-
- block(1..n).
- location(X) :- block(X).
- location(table).
-\end{verbatim}
-Now we stack all blocks on top of each other, so block 1 is on block 2, block 2 is on block 3 etc.
-The last block we put on the table.
-This all holds in the initial state, so a $0$ is appended to the ``\texttt{on}'' predicate.
-\begin{verbatim}
- on(X,X+1,0) :- block(X), X < n.
- on(n,table,0).
-\end{verbatim}
-After declaring the base part and describing the initial situation we have to describe valid moves for the blocks.
-This is done in the cumulative part, since moves can be done in each step.
-For the incremental constant we choose the name ``\texttt{k}''.
-At first we want to execute at least one move every step, but not more than one. This can be done with a cardinality constraint.
-\begin{verbatim}
- #cumulative k.
-
- 1{move(X,Y,k) : block(X) : location(Y) : X != Y}1.
-\end{verbatim}
-A move is of the form ``\texttt{move(X,Y,k)}'' where \texttt{X} is the block that is moved, \texttt{Y} the new position of the block and \texttt{k} the step when the moved is executed.
-\newline
-Now that we have exactly one move per step we need to define what happens if we move a block.
-Understandably if we move block \texttt{X} to position \texttt{Y} in step \texttt{k}, the new location of the block shall be \texttt{Y}.
-\begin{verbatim}
- on(X,Y,k) :- move(X,Y,k).
-\end{verbatim}
-For all other blocks that have not moved, the position shall be the same.
-In ASP a little trick is used to describe such a condition.
-We first define a little helper predicate that states that a block has moved.
-\begin{verbatim}
- moved(X, k) :- move(X, Y, k).
-\end{verbatim}
-With this helping predicate we can define that a block is on the same position in step \texttt{k}, if it was not moved.
-\begin{verbatim}
- on(X,Y,k) :- on(X,Y,k-1), block(X), location(Y), not moved(X, k).
-\end{verbatim}
-
-Now we have to restrict our moves, because only one block can be moved at the same time and it can only be moved if no other block is on top of it.
-First we introduce a predicate called ``\texttt{blocked}'' for each block with another block on top of it.
-\begin{verbatim}
- blocked(Y, k) :- block(Y), on(X, Y, k - 1).
-\end{verbatim}
-Now we can restrict the moves with an integrity constraint.
-\begin{verbatim}
- :- move(X, Y, k), blocked(X, k).
-\end{verbatim}
-So it is not allowed to move block \texttt{X} in step k, if it is blocked.
-Also you can not move a block to a position where another block is on, so only one block on top of another is allowed.
-\begin{verbatim}
- :- move(X, Y, k), blocked(Y, k).
-\end{verbatim}
-The description of the problem is nearly finished. All that has to be done is to formulate a query.
-As a constraint to our solution we want to stop if all blocks are on top of each other in reversed order.
-This query shall only hold in the last step, therefore we write it in the volatile part.
-As always for constraints, we use an integrity constraint to express that all solutions are false where there is not block $n+1$ on the block $n$.
-\begin{verbatim}
- #volatile k.
-
- :- not on(X+1,X,k), block(X), X < n.
-\end{verbatim}
-To run this example, iclingo has to be used and the constant $n$ has to be set via command line.
-A potential command line call could look like
-\begin{verbatim}
- iclingo --const n=10 blocksworld.lp
-\end{verbatim}
-and the answer to the problem can be read from the answer set.
-To clean up the answer the ``\texttt{show} and \texttt{hide}'' feature of gringo can be used.
-If the following lines were added to the program, the answer set only consists of \texttt{move/3} predicates.
-\begin{verbatim}
- #hide.
- #show move/3.
-\end{verbatim}
-
-\subsection{Meta-Syntax}
-This section is about constructs that do not originally belong to the logic program but can be used to give additional information for example to the solver.
-The ``\#'' sign in front of all meta expressions is optional but is it strongly recommend to use it.
-It separates the meta directives from the logic program more clearly and can prevent severe mistakes,
- confusing those constructs with atoms as shown in the example.
-\begin{verbatim}
- hide.
- hidden :- hide.
-\end{verbatim}
-This program hides all predicates from the output of the ASP solver instead of deriving the facts \texttt{hide} and \texttt{hidden}.
-
-\paragraph{hide}
-The hide construct is used to hide atoms from the output of the ASP solver.
-Two different syntaxes are supported by gringo2.0.
-The first 
-\begin{verbatim}
- #hide name/arity.
-\end{verbatim}
-and the second
-\begin{verbatim}
- #hide name(X1, ... , Xn).
-\end{verbatim}
-Both variants hide the predicate ``\texttt{name}''.
-The arity of the predicate can be given explicit or implicit with a list of variables.
-The variables are just placeholder and do not unify with other variables.
-For example \texttt{hide p(X,X)} would also hide the fact \texttt{p(a,b)}.
-The second syntax is implemented for compatibility reasons only.
-If the predicate that has to be hidden has arity zero the syntax is
-\begin{verbatim}
- #hide name.
-\end{verbatim}
-If you want to express that all but a few predicates shall be hidden the empty hide statement
-in conjunction with the ``\texttt{show}'' statement can be is used.
-With ``\texttt{show}'' all predicates are named that shall be visible.
-So to hide all predicates but $p/3$ the following example can be used.
-\begin{verbatim}
- #hide.
- #show p/3.
-\end{verbatim}
-
-\paragraph{const}
-Within the ``\texttt{const}'' statement, a default value for constants can be defined in the logic program.
-while they are normally defined via the command line options.
-A default value for a constant $n$ can be easily be set.
-\begin{verbatim}
- #const n=5.
-\end{verbatim}
-\paragraph{domain}
-A domain statement states that a specific variable has a global domain restriction.
-The statement
-\begin{verbatim}
- #domain time(T).
-\end{verbatim}
-makes sure that the variable ``\texttt{T}'' is always bound to the domain of ``\texttt{time}''.
-This has the same effect as adding the domain ``\texttt{time(T)}'' to every rule the variable ``\texttt{T}'' occurs in.
-\paragraph{compute}
-The compute statement is a hint for the solver which models to compute.
-It has the syntax
-\begin{verbatim}
- #compute number {a1, ..., aN, not aN+1, ..., not aM}.
-\end{verbatim}
-$number$ is the number of models that shall be computed by the answer set solver and can be omitted. If it is zero, no restriction to the number of models is applied.
-All positive atoms in the compute statement must be in the computed models, and all negated atoms do not have to be in the models of the program.
-The compute statement is solver dependent and not supported by every solver.
-\paragraph{comments}
-Comments included in the input are line based.
-Whenever a ``\texttt{\%}'' occurs in the input, the rest of the line is treated as a comment and is therefore omitted.
-\subsubsection{Meta-Syntax for Incremental Programs}
-Incremental programs must be splitted into 3 parts.
-The base part, for the information that hold in all steps ans is independent of the incremental constant.
-The cumulative part, where the cumulative rules must be set that hold in all steps, but depend on the incremental parameter.
-And finally the volatile part. In this part the queries to the program are given, that must hold in the last state.
-To seperate the parts, three statement exists. Everything beyond such a statement is considered to belong to the specific part.
-To start the base part
-\begin{verbatim}
- #base.
-\end{verbatim}
-must be written. Now all rules of the base part can follow.
-The use of the incremental constant is not allowed in this part.
-For the cumulative part an additional constant has to be passed to the statement.
-\begin{verbatim}
- #cumulative k.
-\end{verbatim}
-Now all incremental rules follow. The incremental constant used for grounding is $k$.
-The same holds for the volatile part.
-\begin{verbatim}
- #volatile k.
-\end{verbatim}
-The ``\#''-sign can not be omitted for these meta statements.
-These statements can also be mixed, only the last statement is important.
-If no statement is given at the beginning of the file, base is assumed.
 
@@
 
+\subsection{Meta-Syntax}
+This section is about constructs that do not originally belong to the logic program but can be used to give additional information for example to the solver.
+The ``\#'' sign in front of all meta expressions is optional but is it strongly recommend to use it.
+It separates the meta directives from the logic program more clearly and can prevent severe mistakes,
+ confusing those constructs with atoms as shown in the example.
+\begin{verbatim}
+ hide.
+ hidden :- hide.
+\end{verbatim}
+This program hides all predicates from the output of the ASP solver instead of deriving the facts \texttt{hide} and \texttt{hidden}.
+
+\paragraph{hide}
+The hide construct is used to hide atoms from the output of the ASP solver.
+Two different syntaxes are supported by gringo2.0.
+The first 
+\begin{verbatim}
+ #hide name/arity.
+\end{verbatim}
+and the second
+\begin{verbatim}
+ #hide name(X1, ... , Xn).
+\end{verbatim}
+Both variants hide the predicate ``\texttt{name}''.
+The arity of the predicate can be given explicit or implicit with a list of variables.
+The variables are just placeholder and do not unify with other variables.
+For example \texttt{hide p(X,X)} would also hide the fact \texttt{p(a,b)}.
+The second syntax is implemented for compatibility reasons only.
+If the predicate that has to be hidden has arity zero the syntax is
+\begin{verbatim}
+ #hide name.
+\end{verbatim}
+If you want to express that all but a few predicates shall be hidden the empty hide statement
+in conjunction with the ``\texttt{show}'' statement can be is used.
+With ``\texttt{show}'' all predicates are named that shall be visible.
+So to hide all predicates but $p/3$ the following example can be used.
+\begin{verbatim}
+ #hide.
+ #show p/3.
+\end{verbatim}
+
+\paragraph{const}
+Within the ``\texttt{const}'' statement, a default value for constants can be defined in the logic program.
+while they are normally defined via the command line options.
+A default value for a constant $n$ can be easily be set.
+\begin{verbatim}
+ #const n=5.
+\end{verbatim}
+\paragraph{domain}
+A domain statement states that a specific variable has a global domain restriction.
+The statement
+\begin{verbatim}
+ #domain time(T).
+\end{verbatim}
+makes sure that the variable ``\texttt{T}'' is always bound to the domain of ``\texttt{time}''.
+This has the same effect as adding the domain ``\texttt{time(T)}'' to every rule the variable ``\texttt{T}'' occurs in.
+\paragraph{compute}
+The compute statement is a hint for the solver which models to compute.
+It has the syntax
+\begin{verbatim}
+ #compute number {a1, ..., aN, not aN+1, ..., not aM}.
+\end{verbatim}
+$number$ is the number of models that shall be computed by the answer set solver and can be omitted. If it is zero, no restriction to the number of models is applied.
+All positive atoms in the compute statement must be in the computed models, and all negated atoms do not have to be in the models of the program.
+The compute statement is solver dependent and not supported by every solver.
+\paragraph{comments}
+Comments included in the input are line based.
+Whenever a ``\texttt{\%}'' occurs in the input, the rest of the line is treated as a comment and is therefore omitted.
+TODO: Block comments
+\subsection{Meta-Syntax for Incremental Programs}
+Incremental programs must be splitted into 3 parts.
+The base part, for the information that hold in all steps ans is independent of the incremental constant.
+The cumulative part, where the cumulative rules must be set that hold in all steps, but depend on the incremental parameter.
+And finally the volatile part. In this part the queries to the program are given, that must hold in the last state.
+To seperate the parts, three statement exists. Everything beyond such a statement is considered to belong to the specific part.
+To start the base part
+\begin{verbatim}
+ #base.
+\end{verbatim}
+must be written. Now all rules of the base part can follow.
+The use of the incremental constant is not allowed in this part.
+For the cumulative part an additional constant has to be passed to the statement.
+\begin{verbatim}
+ #cumulative k.
+\end{verbatim}
+Now all incremental rules follow. The incremental constant used for grounding is $k$.
+The same holds for the volatile part.
+\begin{verbatim}
+ #volatile k.
+\end{verbatim}
+The ``\#''-sign can not be omitted for these meta statements.
+These statements can also be mixed, only the last statement is important.
+If no statement is given at the beginning of the file, base is assumed.
+
 \section{Examples}
 In this section some encodings for common problems are presented and explained.
-\paragraph{Fastfood}
-This example needs a proper instance to work.
+
+\subsection{How to Write a Logic Program}
+Writing a logic program to solve a problem is rather describing a problem than describing an algorithm to solve a problem.
+Therefore the problem must be encoded as a set of rules.
+Typically these are divided into a problem description and an instance of the problem.
+The solution to a problem can then be read from the answer sets of the program.
+The procedure of writing a logic program shall now be explained with a little example.
+We want to write a problem encoding for the 3-colorability problem of graph coloring.
+Given an arbitrary graph, no two connected edges may have the same color.
+The most common technique is the ``generate \& test'' approach.
+Therefore, in the generator, rules are created that can derive a superset of the solution to the problem.
+In the tester all solutions that do not fit the constraints are omitted.
+\newline
+Usually you start with the with the generator.
+At first, we define the non instance domains, in our example the colors we use.
+\begin{verbatim}
+ color(red). color(green). color(blue).
+\end{verbatim}
+Now we have to color all the vertices, but we must give attention that each vertex has only one color.
+We can do this by simply saying that a vertex is colored with one color $C$, and that the same vertex does not have any of the other colors $D$ or $E$.
+\begin{verbatim}
+ colored(V,C) :- not colored(V,D), not colored(V,E),
+                 C != D, C != E, D != E,
+                 color(E), color(C), color(D), vertex(V).
+\end{verbatim}
+As you can see, you have to give the domain of the variables in the rule.
+The variable $V$ is used for vertices and the variables $C$, $D$ and $E$ are used for colors.
+After every vertex now has a color we would get all possible solutions of all vertices colorfully painted.
+We now need to restrict our solutions in the tester by saying that two vertices that are connected by an edge should not have the same color.
+We can do this by using an integrity constraint.
+An integrity constraint is a rule with an empty head, and if an assignment of variables fulfills the body, the assignment is not a solution of our problem.
+\begin{verbatim}
+ :- edge(V,U), colored(V,C), colored(U,C),
+    vertex(V), vertex(U), color(C).
+\end{verbatim}
+Here we have the edge between two vertices $V$ and $U$, and if both are colored with the same color $C$ the assignment is not a solution to our problem.
+\newline
+We have finally written our problem description, all we need now is a problem instance. This step is fairly simple.
+We just need to define several vertices and edges that do connect the vertices.
+For the following simple graph
+\[
+ \xymatrix{1\ar@{-}[dr]\ar@{-}[d]\ar@{-}[r]&2\ar@{-}[dl]\\
+           3&4}
+\]
+
+we could just write
+\begin{verbatim}
+ vertex(1..4).
+ edge(1,2). edge(1,4).
+ edge(3,1). edge(3,2).
+\end{verbatim}
+
+If we give the problem description and the instance to the grounder, it will generate a variable free version of the problem.
+This can then be solved by a solver and in the answer sets of the solver the ``\texttt{colored}'' predicate states a valid coloring of all vertices.
+We can do this by simply using the internal interface to clasp, to get a solution.
+\begin{verbatim}
+ ./gringo -c 3colorability.lp instance1.lp
+\end{verbatim}
+In the first answer set we find the ``\texttt{colored}'' predicates
+\begin{verbatim}
+ colored(1,green) colored(2,red) colored(3,blue) colored(4,red)
+\end{verbatim}
+and we can interpret them as the colors of the vertices
+
+\[
+ \xymatrix{\color{green}1\ar@{-}[dr]\ar@{-}[d]\ar@{-}[r]&\color{red}2\ar@{-}[dl]\\
+           \color{blue}3&\color{red}4}
+\]
+
+\subsection{How to Write an Incremental Program}
+As in writing a non incremental logic problem, we describe a problem rather than an algorithm.
+The difference between a non incremental and an incremental program is that we can use an incremental constant.
+This constant is set to zero at the beginning and the solver tries to find a solution of the problem.
+If it can not find a solution it starts to increment the constant by one.
+At each step the solver searches for a solution to the problem and increments the constant until he finds a valid answer set for the logic program.
+The incremental constant can therefore be used among other things to emulate time steps.
+For example, if the solution for a planning problem can not be found in 3 steps, it is tried to find one in more steps.
+To make it concrete, an incremental encoding for the blocksworld problem is presented.
+This is a very simple problem, where $n$ blocks are stacked on top of each other and you can move only one block at the same time. The goal is to stack the blocks in the inverse order.
+Since an incremental logic program is divided into three parts, we start with the base part.
+All rules that hold in all timesteps that are independent of the incremental constant are put here.
+At first we declare $n$ blocks and that each block is a location where a block can be put on.
+Furthermore we define an extra location, the table. That is the location where the bottom most block is put.
+\begin{verbatim}
+ #base.
+
+ block(1..n).
+ location(X) :- block(X).
+ location(table).
+\end{verbatim}
+Now we stack all blocks on top of each other, so block 1 is on block 2, block 2 is on block 3 etc.
+The last block we put on the table.
+This all holds in the initial state, so a $0$ is appended to the ``\texttt{on}'' predicate.
+\begin{verbatim}
+ on(X,X+1,0) :- block(X), X < n.
+ on(n,table,0).
+\end{verbatim}
+After declaring the base part and describing the initial situation we have to describe valid moves for the blocks.
+This is done in the cumulative part, since moves can be done in each step.
+For the incremental constant we choose the name ``\texttt{k}''.
+At first we want to execute at least one move every step, but not more than one. This can be done with a cardinality constraint.
+\begin{verbatim}
+ #cumulative k.
+
+ 1{move(X,Y,k) : block(X) : location(Y) : X != Y}1.
+\end{verbatim}
+A move is of the form ``\texttt{move(X,Y,k)}'' where \texttt{X} is the block that is moved, \texttt{Y} the new position of the block and \texttt{k} the step when the moved is executed.
+\newline
+Now that we have exactly one move per step we need to define what happens if we move a block.
+Understandably if we move block \texttt{X} to position \texttt{Y} in step \texttt{k}, the new location of the block shall be \texttt{Y}.
+\begin{verbatim}
+ on(X,Y,k) :- move(X,Y,k).
+\end{verbatim}
+For all other blocks that have not moved, the position shall be the same.
+In ASP a little trick is used to describe such a condition.
+We first define a little helper predicate that states that a block has moved.
+\begin{verbatim}
+ moved(X, k) :- move(X, Y, k).
+\end{verbatim}
+With this helping predicate we can define that a block is on the same position in step \texttt{k}, if it was not moved.
+\begin{verbatim}
+ on(X,Y,k) :- on(X,Y,k-1), block(X), location(Y), not moved(X, k).
+\end{verbatim}
+
+Now we have to restrict our moves, because only one block can be moved at the same time and it can only be moved if no other block is on top of it.
+First we introduce a predicate called ``\texttt{blocked}'' for each block with another block on top of it.
+\begin{verbatim}
+ blocked(Y, k) :- block(Y), on(X, Y, k - 1).
+\end{verbatim}
+Now we can restrict the moves with an integrity constraint.
+\begin{verbatim}
+ :- move(X, Y, k), blocked(X, k).
+\end{verbatim}
+So it is not allowed to move block \texttt{X} in step k, if it is blocked.
+Also you can not move a block to a position where another block is on, so only one block on top of another is allowed.
+\begin{verbatim}
+ :- move(X, Y, k), blocked(Y, k).
+\end{verbatim}
+The description of the problem is nearly finished. All that has to be done is to formulate a query.
+As a constraint to our solution we want to stop if all blocks are on top of each other in reversed order.
+This query shall only hold in the last step, therefore we write it in the volatile part.
+As always for constraints, we use an integrity constraint to express that all solutions are false where there is not block $n+1$ on the block $n$.
+\begin{verbatim}
+ #volatile k.
+
+ :- not on(X+1,X,k), block(X), X < n.
+\end{verbatim}
+To run this example, iclingo has to be used and the constant $n$ has to be set via command line.
+A potential command line call could look like
+\begin{verbatim}
+ iclingo --const n=10 blocksworld.lp
+\end{verbatim}
+and the answer to the problem can be read from the answer set.
+To clean up the answer the ``\texttt{show} and \texttt{hide}'' feature of gringo can be used.
+If the following lines were added to the program, the answer set only consists of \texttt{move/3} predicates.
 \begin{verbatim}
  #hide.
-#show altdepot/2.
-#show costs/1.
-
-% calculate the number of depots
-depots(X) :- X = { depot(DN, DK) :  depot(DN, DK) }.
-
-% calculate the costs for the initial depots
-% (the grounder can calculate this on its own)
-costs(C) :- C = [ serves(RN, Dist) : restaurant(RN, RK) : dist(Dist) = Dist ].
-
-% choose the closest depot for a restaurant
-% (the grounder can calculate this on its own)
-serves(Rname,Dist) :- restaurant(Rname,RK),
-   Dist = min [ depot(DN1, DK1) : depot(DN1, DK1) = abs(DK1-RK) ].
-
-dist(abs(X-Y)) :- restaurant(A, X), restaurant(B, Y).
-
-% Each restaurant may be an alternative depot or not.
-X { altdepot(R,K) : restaurant(R,K) } X :- depots(X).
-
-% choose the closest depot for a restaurant
-altserves(Rname,Dist) :- restaurant(Rname,RK),
-   Dist = min [ altdepot(DN1, DK1) : restaurant(DN1, DK1) = abs(DK1-RK) ].
-
-% fail if we have not found a better solution
-:- costs(C), C [ altserves(R,Dist) : dist(Dist) : restaurant(R,RK) = Dist ].
-
-% thats the easiest way to show the calculated costs
-minimize [ altserves(R,Dist) : dist(Dist) : restaurant(R,RK) = Dist ].
+ #show move/3.
 \end{verbatim}
-
-\paragraph{Queens}
-\begin{verbatim}
-#hide.
-#show q(X,Y).
-
-% this can be set via the command line
-#const queens = 200.
-
-#const diags  = queens * 2 - 1.
-d(1..queens).
-diag(1..diags).
-
-% enumerate the diagonals
-diagX(X,Y,X - Y + queens) :- d(X), d(Y).
-diagY(X,Y,X + Y - 1) :- d(X), d(Y).
-
-% exactly 1 queen in each row/col
-1 { q(X,Y) : d(Y) } 1 :- d(X).
-1 { q(X,Y) : d(X) } 1 :- d(Y).
-
-% at most 1 queen on each diagonal
-{ q(X,Y) : diagX(X,Y,D) } 1 :- diag(D).
-{ q(X,Y) : diagY(X,Y,D) } 1 :- diag(D).
-
-\end{verbatim}
+%\paragraph{Fastfood}
+%This example needs a proper instance to work.
+%\begin{verbatim}
+%#hide.
+%#show altdepot/2.
+%#show costs/1.
+%
+%% calculate the number of depots
+%depots(X) :- X = { depot(DN, DK) :  depot(DN, DK) }.
+%
+%% calculate the costs for the initial depots
+%% (the grounder can calculate this on its own)
+%costs(C) :- C = [ serves(RN, Dist) : restaurant(RN, RK) : dist(Dist) = Dist ].
+%
+%% choose the closest depot for a restaurant
+%% (the grounder can calculate this on its own)
+%serves(Rname,Dist) :- restaurant(Rname,RK),
+%   Dist = min [ depot(DN1, DK1) : depot(DN1, DK1) = abs(DK1-RK) ].
+%
+%dist(abs(X-Y)) :- restaurant(A, X), restaurant(B, Y).
+%
+%% Each restaurant may be an alternative depot or not.
+%X { altdepot(R,K) : restaurant(R,K) } X :- depots(X).
+%
+%% choose the closest depot for a restaurant
+%altserves(Rname,Dist) :- restaurant(Rname,RK),
+%   Dist = min [ altdepot(DN1, DK1) : restaurant(DN1, DK1) = abs(DK1-RK) ].
+%
+%% fail if we have not found a better solution
+%:- costs(C), C [ altserves(R,Dist) : dist(Dist) : restaurant(R,RK) = Dist ].
+%
+%% thats the easiest way to show the calculated costs
+%minimize [ altserves(R,Dist) : dist(Dist) : restaurant(R,RK) = Dist ].
+%\end{verbatim}
+%
+%\paragraph{Queens}
+%\begin{verbatim}
+%#hide.
+%#show q(X,Y).
+%
+%% this can be set via the command line
+%#const queens = 200.
+%
+%#const diags  = queens * 2 - 1.
+%d(1..queens).
+%diag(1..diags).
+%
+%% enumerate the diagonals
+%diagX(X,Y,X - Y + queens) :- d(X), d(Y).
+%diagY(X,Y,X + Y - 1) :- d(X), d(Y).
+%
+%% exactly 1 queen in each row/col
+%1 { q(X,Y) : d(Y) } 1 :- d(X).
+%1 { q(X,Y) : d(X) } 1 :- d(Y).
+%
+%% at most 1 queen on each diagonal
+%{ q(X,Y) : diagX(X,Y,D) } 1 :- diag(D).
+%{ q(X,Y) : diagY(X,Y,D) } 1 :- diag(D).
+%
+%\end{verbatim}
 

### 2008-09-03 doc/documentation/paper.tex
Doku: Small changes, nearly every thing done
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@487 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 \author{Max Ostrowski}
-\title{Gringo 2.0}
+\title{Gringo 2.0\\--Draft--}
 \date{\today}
@@
 gringo2.0 is a state of the art grounder and has many features that makes writing logic programs easy.
-It was developed by Roland Kaminski and is orientated on the ideas of gringo1.0.
-Therefore it supports binder splitting, backjumping and all other features already implemented in gringo1.0.\cite{gringopaper}
+It was developed by Roland Kaminski and is orientated on the ideas of gringo1.0\cite{gringopaper}.
+Therefore it supports binder splitting, backjumping and all other features already implemented in gringo1.0.
 A broad set of potentialities for writing encodings for logic programs exists. Currently, no other grounder supports such a rich featured language.
@@
 The first version is gringo2.0, it acts as a grounder and supports several output formats and even a human readable format for debugging purposes.
-This version can ground logic programs and output them to files or pipe them to the solver.
+This version can ground logic programs and output them to files or pipe them to a solver.
 The second version is called clingo and is a grounder with a built-in solver clasp \cite{gekanesc07b}.
@@
 In this version the grounder gringo2.0 can be used with clasp incrementally.
-Detailed information about incremental solving can be found here \cite{gekaosscth08a}
-In this paper, at first, the general restrictions to the input language are stated. Then a brief description of how to use gringo2.0, clingo and iclingo is given.
-In the next chapter the supported input language is described in detail and a short introduction of how to write a logic program is given.
-The paper closes with some examples of logic programs and gives an outlook of future work that has to be done in the area of grounding logic programs.
+Detailed information about incremental solving can be found in \cite{gekaosscth08a}.
+In this paper, at first, a brief description of how to use the three programs gringo2.0, clingo and iclingo and their options for the commandline is given. 
+In the next chapter the general restrictions to the input language are stated and the supported input language is described in detail.
+After that, a short introduction of how to write a logic program is given.
+The paper closes with an outlook of future work that has to be done in the area of grounding logic programs.
 The syntax of the gringo2.0 input language can be found in the appendix \ref{appendix}.
@@
 \subsection{Command Line Interface}
+All gringo tools can be downloaded as binary from \cite{download}. To build them from source, an installation guide ``INSTALL'' is included in the tar file.
 The gringo tools can be started from the command line and have some parameters that can be useful.
@@
 \paragraph{\texttt{--verbose,-V}}
-Prints extra information during processing.
+Prints additional information during processing.
 \paragraph{\texttt{--stats}}
@@
 With this option you can enable or disable the bindersplitting used by gringo2.0
-%TODO	Default: yes
 \begin{itemize}
@@
 \paragraph{Default Parameters}
-The default command line is ``\texttt{./gringo --lparse}'' to read from standard input and for smodels output.
+The default command line is ``\texttt{./gringo --lparse --bindersplitting=yes}'' to read from standard input and for smodels output.
 
@@
 \begin{verbatim}
- clingo 1 --clasp --solveonly=No
+ clingo 1 --clasp --solveonly=no
            --trans-ext=no --eq=5 
@@
 Iclingo is used for incremental solving. For a detailed description see \cite{gekaosscth08a}.
-It uses the grounder gringo2.0 and the solver clasp, and therefore all options if gringo2.0, clingo and clasp are valid.
+It uses the grounder gringo2.0 and the solver clasp, and therefore all options of gringo2.0, clingo and clasp are valid.
 Additionally some options for controlling the incremental behavior are implemented.
@@
 \paragraph{\texttt{--ibase}}
-With this option enabled, iclingo solves only the base program and ignores the incremental and query part of the program.
+With this option enabled, iclingo solves only the base program and ignores the cumulative and volatile part of the program.
 This option can be helpful when debugging.
 \paragraph{\texttt{--istop=<arg>}}
-The stop condition during incremental solving be set to ``\texttt{SAT}'' or ``\texttt{UNSAT}'' to stop the search process either if a solution was found (\texttt{SAT} and default) or if no solution was found (\texttt{UNSAT}).
+The stop condition during incremental solving can be set.
+\begin{itemize}
+ \item SAT    : Stop the search process if a solution was found.
+ \item UNSAT  : Stop the search process if no solution was found.
+\end{itemize}
+The default for this option is SAT.
 \paragraph{\texttt{--ilearnt=<arg>}}
@@
 The building blocks of the language are terms.
-Each constant $c \in \mathcal{C}$ and each function symbol $f \in \mathcal{F}$ is a term.
-Furthermore $f(t_1, \dots, t_n)$ is a term if $f/n \in \mathcal{F}$ and $t_i \mid i \in \{1 \dots n\}$ are terms.
+Each variable $v \in \mathcal{V}$, constant $c \in \mathcal{C}$ and each function symbol $f \in \mathcal{F}$ is a term.
+Furthermore $f(t_1, \dots, t_n)$ is a term if $f/n \in \mathcal{F}$ and $t_i$ for $i \in \{1 \dots n\}$ are terms.
 Since a term is defined recursively, it can be nested.
-A term with the first parameter being a constant $a$, the others being terms $f(b,X)$ and $g(f(X), Y))$ would be expressed this way.
-\newline
-TODO: This is not a term following the definition, it contains variables!!! Is a variable a Term ?
+A term with the first parameter being a constant $a$, the others being terms $f(b,X)$ and $g(f(X), Y)$ would be expressed this way.
 \begin{verbatim}
@@
 In gringo2.0 every variable used in a rule must have a complete domain.
-This means that this variable has to occur in at least one predicate whichs domain is completed.
+This means that this variable has to occur in at least one predicate whose domain is completed.
 Therefore the predicate must be in the head of rules that are already domain restricted.
@@
 
-\subsection{Features}
-To ease the process of writing logic programs some features in the input syntax are introduced and are now explained in detail.
+\subsection{Basic Language}
+The base building blocks of the language are rules, facts and integrity constraints.
+\paragraph{Rule}
+A rule normally consists of a head and a body and ends with a dot.
+The body is separated from the head using ``:-'' as in
+\begin{verbatim}
+ head :- b1, ..., bN, not c1, ..., cN.
+\end{verbatim}
+The head consists of terms concatenated by a disjunction. For non disjunctive rules only one term is used.
+The body is a comma separated conjunction of terms and their negation.
+The variables used must be restricted to a domain in the body.
+Check (\ref{domainrestricted}) to see the restrictions on domains.
+\paragraph{Fact}
+A rule can occur without a body as a fact, therefore the body is omitted.
+\begin{verbatim}
+ head.
+\end{verbatim}
+\paragraph{Integrity Constraint}
+To express an integrity constraint, a rule with an empty body is used.
+\begin{verbatim}
+ :- body.
+\end{verbatim}
+
+\subsection{Advanced Language Features}
+To ease the process of writing logic programs some advanced features in the input syntax are introduced and are now explained in detail.
 For the most problems it is possible to write a much shorter version of the program with the extended syntax.
@@
 \end{verbatim}
-Furthermore the assignment operator can be used with aggregates, for example
+Furthermore the assignment operator can be used with aggregates. The program
 \begin{verbatim}
@@
 
-\paragraph{hide}
+\paragraph{hide and show}
 The hide construct is used to hide atoms from the output of the ASP solver.
@@
 \paragraph{comments}
-Comments included in the input are line based.
+Two kinds of comments were supported by gringo2.0.
+Line based comments start with a ``\texttt{\%}''.
 Whenever a ``\texttt{\%}'' occurs in the input, the rest of the line is treated as a comment and is therefore omitted.
-TODO: Block comments
+Block based comments can span over several lines.
+Everything between ``\texttt{\%*}'' and ``\texttt{*\%}'' is treated as a comment.
 \subsection{Meta-Syntax for Incremental Programs}
 Incremental programs must be splitted into 3 parts.
-The base part, for the information that hold in all steps ans is independent of the incremental constant.
+The base part, for the information that hold in all steps and is independent of the incremental constant.
 The cumulative part, where the cumulative rules must be set that hold in all steps, but depend on the incremental parameter.
-And finally the volatile part. In this part the queries to the program are given, that must hold in the last state.
-To seperate the parts, three statement exists. Everything beyond such a statement is considered to belong to the specific part.
+And finally the volatile part. In this part the queries to the program are given, that must hold only in the last state.
+To seperate the parts, three statements exists. Everything beyond such a statement is considered to belong to the specific part.
 To start the base part
@@
 \newline
-Usually you start with the with the generator.
+Usually you start with the generator.
 At first, we define the non instance domains, in our example the colors we use.
@@
 Now we have to color all the vertices, but we must give attention that each vertex has only one color.
-We can do this by simply saying that a vertex is colored with one color $C$, and that the same vertex does not have any of the other colors $D$ or $E$.
+We can do this by simply saying that a vertex $V$ is colored with one color $C$, and that the same vertex does not have any of the other colors $D$ or $E$.
 \begin{verbatim}
@@
 \begin{verbatim}
- :- edge(V,U), colored(V,C), colored(U,C),
-    vertex(V), vertex(U), color(C).
+ :- edge(V,U), colored(V,C), colored(U,C).
 \end{verbatim}
@@
 If we give the problem description and the instance to the grounder, it will generate a variable free version of the problem.
-This can then be solved by a solver and in the answer sets of the solver the ``\texttt{colored}'' predicate states a valid coloring of all vertices.
+This can then be solved by a solver and in the answer sets of the solver the ``\texttt{colored/2}'' predicate states a valid coloring of all vertices.
 We can do this by simply using the internal interface to clasp, to get a solution.
@@
 \]
+To improve the readability of the answerset, all but the \texttt{colored/2} predicates can be hidden with these two lines.
+\begin{verbatim}
+ #hide.
+ #show colored/2.
+\end{verbatim}
 

### 2008-09-03 app/options.cpp

@@ -245,7 +245,7 @@ void Options::initOptions(ProgramOptions::OptionGroup& allOpts, ProgramOptions::
 	common.addOptions()
 		("help,h"   , bool_switch(&help),   "Print help and exit")
 		("version,v", bool_switch(&version),"Print version and exit")
-		("verbose,V", bool_switch(&verbose), "Print extra information")
+		("verbose,V", bool_switch(&verbose), "Print additional information")
 		("stats"    , bool_switch(&stats),  "Print extended statistics")
 #ifdef WITH_CLASP
 		("solveonly"     , value<bool>(&grounder)->parser(mapMode), "Set the working mode\n"

### 2008-09-04 doc/documentation/appendix.tex
fixed some spelling mistakes in gringo errors, DOKU: redesign, fixed appendix, added new warnings <- description still missing
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@491 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 IDENTIFIER      = [LETTER | UNDERSCORE], {IN_STRING};
-STRING          = '"' {INSTRING-('"' | 'NL')} '"';
+STRING          = '"', {INSTRING-('"' | 'NL')}, '"';
 VARIABLE        = [CAP_LETTER],  {IN_STRING};
-COMMENT         = '%' {IN_STRING-NL}, NL;
+COMMENT         = '%', {IN_STRING-NL}, NL;
+BLOCK_COMMENT   = '%*', {IN_STRING | NL}, '*%';
 SHOW            = ['#'], 'show';
@@
 COMPUTE         = ['#'], 'compute';
+BASE            = ['#'], 'base';
+CUMULATIVE      = ['#'], 'cumulative';
+VOLATILE        = ['#'], 'volatile';
 DISJUNCTION     = '|';
@@
 PROGRAM = PROGRAM, CONST, IDENTIFIER, ASSIGN, CONST_TERM, DOT;
+PROGRAM = PROGRAM, BASE, DOT;
+PROGRAM = PROGRAM, CUMULATIVE, IDENTIFIER, DOT;
+PROGRAM = PROGRAM, VOLATILE, IDENTIFIER, DOT;
 PROGRAM = DOT;

### 2008-09-04 doc/documentation/paper.tex
Doku: forgot small changes
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@488 3e02f596-3131-0410-8352-e04eeaec37cc
fixed some spelling mistakes in gringo errors, DOKU: redesign, fixed appendix, added new warnings <- description still missing
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@491 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 Detailed information about incremental solving can be found in \cite{gekaosscth08a}.
-In this paper, at first, a brief description of how to use the three programs gringo2.0, clingo and iclingo and their options for the commandline is given. 
-In the next chapter the general restrictions to the input language are stated and the supported input language is described in detail.
-After that, a short introduction of how to write a logic program is given.
+In this paper, at first, the general restrictions to the input language are stated and the supported input language is described in detail.
+After that, a short introduction of how to write a logic program and how to write an incremental logic program is given.
+In the next chapter a brief description of how to use the three programs gringo2.0, clingo and iclingo and their options for the command line is given.
 The paper closes with an outlook of future work that has to be done in the area of grounding logic programs.
 The syntax of the gringo2.0 input language can be found in the appendix \ref{appendix}.
+
+\section{Syntax}
+The basic elements of a logic program and its syntax is now described.
+Disjunctive logic programs are based on an alphabet \cite{syntax} consisting of:
+\begin{itemize}
+\item a set $\mathcal{P}$ of predicate symbols,
+\item a set $\mathcal{F}$ of function symbols,
+\item a set $\mathcal{V}$ of variables,
+\item a set $\mathcal{C}$ of constants,
+\item a set \{ ``$\mid$'', ``,'' ,$\leftarrow$, not \} of connectives, and
+\item a set \{ ``('', ``)'', ``,'' \} of punctuation symbols.
+\end{itemize}
+
+\paragraph{Predicates and Function Symbols}
+Predicates and function symbols have a predicate symbol/name and an arity.
+For the function symbol $f$ with arity 3 we write $f/3$.
+The names of the function symbols and predicates have to start with either a lower case letter or an underscore and can contain digits and letters.
+With $\mathcal{P}$ we denote to the set of predicates and $\mathcal{F}$ denotes the set of function symbols.
+\paragraph{Constants}
+Constants are function symbols with an arity of zero.
+The name of a constant can also be encompassed by double quotation marks, which is sometimes helpful when using constants with spaces.
+It has to be remarked that a constant with quotation marks is never equal to a constant without, so ``\texttt{a}'' is not equal to \texttt{a}.
+$\mathcal{C}$ is the set of constants.
+\paragraph{Variables}
+Variables have to start with a capitalized letter and can contain digits and letters.
+The set of variables is denoted by $\mathcal{V}$.
+The language over $\mathcal{P}$, $\mathcal{F}$ and $\mathcal{C}$ is denoted by $\mathcal{L(P,F,C)}$.
+\paragraph{Terms}
+The building blocks of the language are terms.
+Each variable $v \in \mathcal{V}$, constant $c \in \mathcal{C}$ and each function symbol $f \in \mathcal{F}$ is a term.
+Furthermore $f(t_1, \dots, t_n)$ is a term if $f/n \in \mathcal{F}$ and $t_i$ for $i \in \{1 \dots n\}$ are terms.
+Since a term is defined recursively, it can be nested.
+A term with the first parameter being a constant $a$, the others being terms $f(b,X)$ and $g(f(X), Y)$ would be expressed this way.
+\begin{verbatim}
+ p(a,f(b,X), g(f(X),Y)).
+\end{verbatim}
+\paragraph{Atoms}
+An atom over $\mathcal{L(P, F, C)}$ is of the form $p(t_1, \dots, t_n)$ where $p/n$ is a predicate
+symbol with arity $n$, and $t_i \mid i \in \{1 \dots n\}$ are terms.
+\paragraph{Rules}
+A rule over the language $\mathcal{L(P,F,C)}$ has the form
+$$
+A_0 \mid \dots \mid A_l \leftarrow A_{l+1}, \dots, A_m, not\ A_{m+1}, \dots, not\ A_n.
+$$
+where $A_0$ to $A_n$ are atoms. \newline
+A logic program $\Pi$ over $\mathcal{L(P,F,C)}$ is a finite set of rules over $\mathcal{L(P,F,C)}$.
+
+\subsection{$\lambda$-Restriction}
+\label{domainrestricted}
+There exist two major grounders for ASP besides gringo1.0, lParse \cite{lparseManual} and  the DLV grounding module\cite{DLV}.
+To ensure the existence of an equivalent ground instantiation of a program both grounders except a subclass of logic programs.
+For lParse the logic programs must be $\omega$-restricted\cite{omega} while DLV can handle safe programs\cite{safe}.
+gringo2.0 verifies that the input program is $\lambda$-
+restricted \cite{gringopaper}, so that the existence of a finite equivalent ground instantiation is guaranteed. 
+$\lambda$-restrictedness extends the capabilities of $\omega$-restricted programs.
+The definition of $\lambda$-restrictedness can be found in \cite{gringopaper} and was expanded for aggregates and disjunctions.
+$\lambda$-restrictedness can not be compared with the safe programs of DLV since there exist programs that are safe and not $\lambda$-restricted and vice versa.
+\newline
+To give a gross impression of what $\lambda$-restriction means for writing a logic program an informal description is given.
+In gringo2.0 every variable used in a rule must have a complete domain.
+This means that this variable has to occur in at least one predicate whose domain is completed.
+Therefore the predicate must be in the head of rules that are already domain restricted.
+So the grounder has to find a valid order of the rules so that every predicate in the head is completed before it is used as a domain predicate for the next rules.
+To make things more clear, an example is given
+\begin{verbatim}
+ zig(0) :- not zag(0).   zag(0) :- not zig(0).
+ zig(1) :- not zag(1).   zag(1) :- not zig(1).
+ zigzag(X,Y) :- zig(X), zag(Y).
+\end{verbatim}
+The rules of this example can be ordered in the following way.
+At first, the rules with \texttt{zig/1} and \texttt{zag/1} are considered.
+Since these are the only rules \texttt{zig/1} and \texttt{zag/1} occurs in the head, the domains of the predicates are closed.
+The domains \{\texttt{zig(0)}, \texttt{zig(1)}\} and \{ \texttt{zag(0)}, \texttt{zag(1)}\} are complete, since no other predicates of the form \texttt{zig/1} or \texttt{zag/1} can be derived from the program.
+Now the third rule with \texttt{zigzag/2} in the head can be grounded.
+The variables \texttt{X} and \texttt{Y} both occur in a domain predicate, and the values $0$ respectively $1$ can substitute the variables.
+\newline
+To give a comparison to the $\omega$-restricted programs of lParse, an extra domain predicate has to be introduced to the example.
+\begin{verbatim}
+ domain(0..1).
+ zig(0) :- not zag(0).   zag(0) :- not zig(0).
+ zig(1) :- not zag(1).   zag(1) :- not zig(1).
+ zigzag(X,Y) :- zig(X), zag(Y), domain(X), domain(Y).
+\end{verbatim}
+\texttt{zig/1} and \texttt{zag/1} can not be used as a domain predicate, since their truth value can not be fully evaluated.
+
+\subsection{Basic Language}
+The base building blocks of the language are rules, facts and integrity constraints.
+\paragraph{Rule}
+A rule normally consists of a head and a body and ends with a dot.
+The body is separated from the head using ``:-'' as in
+\begin{verbatim}
+ head :- b1, ..., bN, not c1, ..., cN.
+\end{verbatim}
+The head consists of terms concatenated by a disjunction. For non disjunctive rules only one term is used.
+The body is a comma separated conjunction of terms and their negation.
+The variables used must be restricted to a domain in the body.
+Check (\ref{domainrestricted}) to see the restrictions on domains.
+\paragraph{Fact}
+A rule can occur without a body as a fact, therefore the body is omitted.
+\begin{verbatim}
+ head.
+\end{verbatim}
+\paragraph{Integrity Constraint}
+To express an integrity constraint, a rule with an empty body is used.
+\begin{verbatim}
+ :- body.
+\end{verbatim}
+
+\subsection{Advanced Language Features}
+To ease the process of writing logic programs some advanced features in the input syntax are introduced and are now explained in detail.
+For the most problems it is possible to write a much shorter version of the program with the extended syntax.
+%Even a speedup in solving the instances of the problem can happen TODO ?
+For a full syntax description in EBNF see appendix (\ref{appendix}).
+\paragraph{Function Symbols}
+Function symbols are fully supported with gringo2.0.
+They can be finitely nested and also be used as domain predicates.
+While in lParse function symbols are allowed in a very restricted way, no restrictions apply in gringo2.0.
+The following program can not be grounded with lParse since $X$ is an unrestricted variable. With gringo2.0 the program is now groundable.
+\begin{verbatim}
+a(X) :- f(1,g(X)).
+f(1,X) :- f(X).
+f(g(2)).
+\end{verbatim}
+From the second rule the domain \texttt{f(1,g(2))} is generated.
+This domain can be used for the first rule and results in the grounded rule
+\begin{verbatim}
+ a(2) :- f(1,g(2)).
+\end{verbatim}
+
+\paragraph{Range Operator}
+To define several atoms that stretch on one or more parameters over a certain range, two dots can be used to express an increasing integer range.
+Using the range operator in the head of a rule, it works as a conjunction of all the members in the range, so for every number the rule is duplicated.
+\begin{verbatim}
+ times(1..3, 1..2).
+\end{verbatim}
+will result in the facts
+\begin{verbatim}
+ times(1,1). times(2,1). times(3,1).
+ times(1,2). times(2,2). times(3,2).
+\end{verbatim}
+
+Using the range operator in the body of a rule, a disjunction of all members in the range is created. So the rules will be duplicated for every member.
+\begin{verbatim}
+ a :- p(1..3).
+\end{verbatim}
+will result in the rules
+\begin{verbatim}
+ a :- p(1).
+ a :- p(2).
+ a :- p(3).
+\end{verbatim}
+If the range operator is used in an aggregate, it creates a comma separated list of atoms, so
+\begin{verbatim}
+ a :- {b(1..3)}.
+\end{verbatim}
+results in
+\begin{verbatim}
+ a :- {b(1), b(2), b(3)}.
+\end{verbatim}
+
+\paragraph{Expander :}
+With the expander ``:'' a domain can be set.
+On the left side of the operator an atom with variables is expected and on the right side the domains for the variables are given.
+In the head, the atoms get connected via a disjunction, in the body they get connected via a conjunction, so no new rules are created with this operator.
+\begin{verbatim}
+ a(X,Y) : b(X) : c(Y) :- d(X) : e(X).
+ b(1..2).
+ c(3).
+ e(a;b;c).
+\end{verbatim}
+would become
+\begin{verbatim}
+ a(1,3) | a(2,3) :- d(a), d(b), d(c).
+\end{verbatim}
+This operator can be used in aggregates to create a comma separated list of atoms.
+For example
+\begin{verbatim}
+ n(1..9).
+ 1{a(X) : n(X) : X > 3}1.
+\end{verbatim}
+would become
+\begin{verbatim}
+ 1 {a(4), a(5), a(6), a(7), a(8), a(9)} 1.
+\end{verbatim}
+
+\paragraph{Conjunctor ;}
+The conjunctor ``;'' will create a conjunction of its elements.
+Used in the head of a rule means that the rule is duplicated.
+If used in the body instead, the atoms will be logically combined using a conjunction.
+\begin{verbatim}
+ a(X;1..2) :- b(X;a;b).
+ b(foo). b(bar).
+\end{verbatim}
+will results in the rules\footnote{gringo2.0 would simplify the grounded rules since some literals in the body are facts}
+\begin{verbatim}
+ a(foo,1) :- b(foo), b(a), b(b).
+ a(foo,2) :- b(foo), b(a), b(b).
+ a(bar,1) :- b(bar), b(a), b(b).
+ a(bar,2) :- b(bar), b(a), b(b).
+\end{verbatim}
+The conjunctor can not be used directly in aggregates, but for example in combination with the expander.
+\begin{verbatim}
+ { a(X,Y) : p(X;Y) }.
+ p(1..2).
+\end{verbatim}
+Results in
+\begin{verbatim}
+ {a(1,1), a(1,2), a(2,1), a(2,2)}.
+ p(1).
+ p(2).
+\end{verbatim}
+
+\paragraph{Arithmetic Expressions}
+The built-in arithmetics can be used for integer arithmetic.
+As operators addition (+), subtraction (-), multiplication (*), division (/), modulo (mod) and the absolute value operator (abs) can be used.
+Furthermore the bitwise operators and(\&), or(?), xor(\textasciicircum) and invert($\sim$) are supported.
+The bitwise operators are executed on signed integers using the binary complement for negative numbers.
+Arithmetic expressions can occur everywhere in the head and in the body, even in complex constructs.
+As parameters, integer constants or expressions that evaluate to integers are used.
+The simple example of using operators and range terms
+\begin{verbatim}
+ a(X-1 .. X+1) :- b(X,Y), c(abs(Y * 2 * -X)).
+ b(1,2).
+ c(4).
+\end{verbatim}
+would result in the grounded program
+\begin{verbatim}
+ a(0) :- b(1,2), c(4).
+ a(1) :- b(1,2), c(4).
+ a(2) :- b(1,2), c(4).
+ b(1,2).
+ c(4).
+\end{verbatim}
+
+\subparagraph{Relational Operators}
+The relational operators are ``less than'' ($<$),\\ ``greater than'' ($>$), ``less equal'' ($<=$), ``greater equal'' ($>=$) , ``equal'' ($==$) and ``not equal'' ($!=$).
+They can be used to constrain certain variables.
+For example, the following rule is only generated for all unifications where $X$ is greater or equal $Y-1$.
+\begin{verbatim}
+ a(X) :- b(X,Y), X >= Y-1.
+ b(0,2).
+ b(4,2).
+\end{verbatim}
+This would let $\texttt{b(4,2)}$ be a valid substitution, since $4 >= 2-1$ but $\texttt{b(0,2)}$ not, so the grounded program could be:
+\begin{verbatim}
+ a(4) :- b(4,2).
+ b(0,2).
+ b(4,2).
+\end{verbatim}
+
+\subparagraph{Assignment Operator}
+As assignment operator ``$=$'' is used.
+The variable on the left side is set to the value of the right side.
+\begin{verbatim}
+ a(X) :- X=Y*2, b(Y).
+ b(1..2).
+\end{verbatim}
+becomes
+\begin{verbatim}
+ a(2) :- b(1).
+ a(4) :- b(2).
+ b(1..2).
+\end{verbatim}
+A cyclic assignment that results in an unbound variable is not allowed and is rejected with an error . For example
+\begin{verbatim}
+ a(X) :- X=Y, Y=X.
+\end{verbatim}
+Furthermore the assignment operator can be used with aggregates. The program
+\begin{verbatim}
+ price(apple,5). prices(banana, 6). price(computer, 100).
+ fruit(apple).
+ fruit(banana).
+ max_cost(1..200).
+ costs(C) :- C = [price(X,Y) : fruit(X) : max_cost(Y) = Y].
+\end{verbatim}
+would lead to the fact \texttt{costs(11)} as the price of all fruits.
+$C$ in this example becomes all values that have to be generated for the aggregate.
+
+\paragraph{Classical Negation}
+For classical negation the atom is preceded by a minus sign.
+\begin{verbatim}
+ -a :- not  a.
+  a :- not -a. 
+\end{verbatim}
+\paragraph{Disjunction}
+A disjunction can be expressed using the pipe symbol ``$\mid$''.
+A disjunction can only occur in the head of a rule.
+\begin{verbatim}
+ red | green | blue :- colored.
+\end{verbatim}
+Using disjunctions in logic programs can increase expressiveness and complexity of the logic program.
+A solver for disjunctive logic programs such as claspD \cite{claspD} or cmodels \cite{cmodels} is needed to solve such programs.
+
+\paragraph{Aggregates}
+An aggregate is a set or a multiset of atoms and has a lower and an upper bound.
+The general syntax is
+\begin{verbatim}
+ lower aggregate_name [atom1=w1, ..., atomN=wN] upper
+\end{verbatim}
+If squared brackets are used, the atom list is interpreted as a multiset, otherwise, for curly brackets, as a set of atoms.
+The lower and upper bound is optional, and can be omitted.
+
+\subparagraph{Count}
+A ``count'' aggregate is a set of atoms and has an optional lower and upper bound.
+It evaluates to true if the number of true atoms in the set is between the lower and upper bound. Only curly brackets may be used for a count aggregate as it represents a set of atoms.
+Weights are not allowed in the count aggregate.
+For this aggregate a shorthand syntax exists
+\begin{verbatim}
+ lower {atom1, ..., atomN} upper
+\end{verbatim}
+The count aggregate is equal to the cardinality constraint known from smodels \cite{smodels}.
+\subparagraph{Sum}
+The ``sum'' aggregate is a multiset of weighted atoms and has an optional lower and upper bound.
+It evaluates to true if the sum of the weights of all true atoms in the multiset is between the lower an upper bound.
+The sum aggregate is equal to the weight constraints known from smodels, and has therefore a shorthand syntax
+\begin{verbatim}
+ lower [atom1=w1, ..., atomN=wN] upper
+\end{verbatim}
+If the command line option ``-l'' is used the semantic may change if negative weights exist in the aggregate. The transformation used is described in \cite{lparseManual}.
+
+\subparagraph{Min}
+The ``min'' aggregate is determined by the minimum of all weights of atoms that evaluate to true.
+For the minimum aggregate a multiset with weighted atoms is used. It has no shorthand syntax.
+\subparagraph{Max}
+The ``max'' aggregate is similar to the minimum aggregate but considers the maximum of the weights.
+
+\paragraph{Optimize Statements}
+
+\subparagraph{Minimize/Maximize}
+An optimize statement is used to find a minimal or maximal answer set of the logic program.
+The statement can be weighted or not, therefore weights can be omitted in the following syntax.
+\begin{verbatim}
+ minimize [atom1 = w1, ..., atomN = wN].
+\end{verbatim}
+The maximize statement uses the keyword ``maximize'' instead.
+If there were several minimize/maximize statements in the logic program the latter will be preferenced.
+Only after the n'th statement results in the same value for two different answer set, the n-1'th statement is considered.
+
+\subsection{Meta-Syntax}
+This section is about constructs that do not originally belong to the logic program but can be used to give additional information for example to the solver.
+The ``\#'' sign in front of all meta expressions is optional but is it strongly recommend to use it.
+It separates the meta directives from the logic program more clearly and can prevent severe mistakes,
+ confusing those constructs with atoms as shown in the example.
+\begin{verbatim}
+ hide.
+ hidden :- hide.
+\end{verbatim}
+This program hides all predicates from the output of the ASP solver instead of deriving the facts \texttt{hide} and \texttt{hidden}.
+
+\paragraph{hide and show}
+The hide construct is used to hide atoms from the output of the ASP solver.
+Two different syntaxes are supported by gringo2.0.
+The first 
+\begin{verbatim}
+ #hide name/arity.
+\end{verbatim}
+and the second
+\begin{verbatim}
+ #hide name(X1, ... , Xn).
+\end{verbatim}
+Both variants hide the predicate ``\texttt{name}''.
+The arity of the predicate can be given explicit or implicit with a list of variables.
+The variables are just placeholder and do not unify with other variables.
+For example \texttt{hide p(X,X)} would also hide the fact \texttt{p(a,b)}.
+The second syntax is implemented for compatibility reasons only.
+If the predicate that has to be hidden has arity zero the syntax is
+\begin{verbatim}
+ #hide name.
+\end{verbatim}
+If you want to express that all but a few predicates shall be hidden the empty hide statement
+in conjunction with the ``\texttt{show}'' statement can be is used.
+With ``\texttt{show}'' all predicates are named that shall be visible.
+So to hide all predicates but $p/3$ the following example can be used.
+\begin{verbatim}
+ #hide.
+ #show p/3.
+\end{verbatim}
+
+\paragraph{const}
+Within the ``\texttt{const}'' statement, a default value for constants can be defined in the logic program.
+while they are normally defined via the command line options.
+A default value for a constant $n$ can be easily be set.
+\begin{verbatim}
+ #const n=5.
+\end{verbatim}
+\paragraph{domain}
+A domain statement states that a specific variable has a global domain restriction.
+The statement
+\begin{verbatim}
+ #domain time(T).
+\end{verbatim}
+makes sure that the variable ``\texttt{T}'' is always bound to the domain of ``\texttt{time}''.
+This has the same effect as adding the domain ``\texttt{time(T)}'' to every rule the variable ``\texttt{T}'' occurs in.
+\paragraph{compute}
+The compute statement is a hint for the solver which models to compute.
+It has the syntax
+\begin{verbatim}
+ #compute number {a1, ..., aN, not aN+1, ..., not aM}.
+\end{verbatim}
+$number$ is the number of models that shall be computed by the answer set solver and can be omitted. If it is zero, no restriction to the number of models is applied.
+All positive atoms in the compute statement must be in the computed models, and all negated atoms do not have to be in the models of the program.
+The compute statement is solver dependent and not supported by every solver.
+\paragraph{comments}
+Two kinds of comments were supported by gringo2.0.
+Line based comments start with a ``\texttt{\%}''.
+Whenever a ``\texttt{\%}'' occurs in the input, the rest of the line is treated as a comment and is therefore omitted.
+Block based comments can span over several lines.
+Everything between ``\texttt{\%*}'' and ``\texttt{*\%}'' is treated as a comment.
+\subsection{Meta-Syntax for Incremental Programs}
+\label{metasyntaxinc}
+Incremental programs must be splitted into 3 parts.
+The base part, for the information that hold in all steps and is independent of the incremental constant.
+The cumulative part, where the cumulative rules must be set that hold in all steps, but depend on the incremental parameter.
+And finally the volatile part. In this part the queries to the program are given, that must hold only in the last state.
+To separate the parts, three statements exists. Everything beyond such a statement is considered to belong to the specific part.
+To start the base part
+\begin{verbatim}
+ #base.
+\end{verbatim}
+must be written. Now all rules of the base part can follow.
+The use of the incremental constant is not allowed in this part.
+For the cumulative part an additional constant has to be passed to the statement.
+\begin{verbatim}
+ #cumulative k.
+\end{verbatim}
+Now all incremental rules follow. The incremental constant used for grounding is $k$.
+The same holds for the volatile part.
+\begin{verbatim}
+ #volatile k.
+\end{verbatim}
+The ``\#''-sign can not be omitted for these meta statements.
+These statements can also be mixed, only the last statement is important.
+If no statement is given at the beginning of the file, base is assumed.
+
+\section{Examples}
+In this section some encodings for common problems are presented and explained.
+
+\subsection{How to Write a Logic Program}
+Writing a logic program to solve a problem is rather describing a problem than describing an algorithm to solve a problem.
+Therefore the problem must be encoded as a set of rules.
+Typically these are divided into a problem description and an instance of the problem.
+The solution to a problem can then be read from the answer sets of the program.
+The procedure of writing a logic program shall now be explained with a little example.
+We want to write a problem encoding for the 3-colorability problem of graph coloring.
+Given an arbitrary graph, no two connected edges may have the same color.
+The most common technique is the ``generate \& test'' approach.
+Therefore, in the generator, rules are created that can derive a superset of the solution to the problem.
+In the tester all solutions that do not fit the constraints are omitted.
+\newline
+Usually you start with the generator.
+At first, we define the non instance domains, in our example the colors we use.
+\begin{verbatim}
+ color(red). color(green). color(blue).
+\end{verbatim}
+Now we have to color all the vertices, but we must give attention that each vertex has only one color.
+We can do this by simply saying that a vertex $V$ is colored with one color $C$, and that the same vertex does not have any of the other colors $D$ or $E$.
+\begin{verbatim}
+ colored(V,C) :- not colored(V,D), not colored(V,E),
+                 C != D, C != E, D != E,
+                 color(E), color(C), color(D), vertex(V).
+\end{verbatim}
+As you can see, you have to give the domain of the variables in the rule.
+The variable $V$ is used for vertices and the variables $C$, $D$ and $E$ are used for colors.
+After every vertex now has a color we would get all possible solutions of all vertices colorfully painted.
+We now need to restrict our solutions in the tester by saying that two vertices that are connected by an edge should not have the same color.
+We can do this by using an integrity constraint.
+An integrity constraint is a rule with an empty head, and if an assignment of variables fulfills the body, the assignment is not a solution of our problem.
+\begin{verbatim}
+ :- edge(V,U), colored(V,C), colored(U,C).
+\end{verbatim}
+Here we have the edge between two vertices $V$ and $U$, and if both are colored with the same color $C$ the assignment is not a solution to our problem.
+\newline
+We have finally written our problem description, all we need now is a problem instance. This step is fairly simple.
+We just need to define several vertices and edges that do connect the vertices.
+For the following simple graph
+\[
+ \xymatrix{1\ar@{-}[dr]\ar@{-}[d]\ar@{-}[r]&2\ar@{-}[dl]\\
+           3&4}
+\]
+
+we could just write
+\begin{verbatim}
+ vertex(1..4).
+ edge(1,2). edge(1,4).
+ edge(3,1). edge(3,2).
+\end{verbatim}
+
+If we give the problem description and the instance to the grounder, it will generate a variable free version of the problem.
+This can then be solved by a solver and in the answer sets of the solver the ``\texttt{colored/2}'' predicate states a valid coloring of all vertices.
+We can do this by simply using the internal interface to clasp, to get a solution.
+\begin{verbatim}
+ ./clingo 3colorability.lp instance1.lp
+\end{verbatim}
+In the first answer set we find the ``\texttt{colored}'' predicates
+\begin{verbatim}
+ colored(1,green) colored(2,red) colored(3,blue) colored(4,red)
+\end{verbatim}
+and we can interpret them as the colors of the vertices
+
+\[
+ \xymatrix{\color{green}1\ar@{-}[dr]\ar@{-}[d]\ar@{-}[r]&\color{red}2\ar@{-}[dl]\\
+           \color{blue}3&\color{red}4}
+\]
+To improve the readability of the answer set, all but the \texttt{colored/2} predicates can be hidden with these two lines.
+\begin{verbatim}
+ #hide.
+ #show colored/2.
+\end{verbatim}
+
+\subsection{How to Write an Incremental Program}
+As in writing a non incremental logic problem, we describe a problem rather than an algorithm.
+The difference between a non incremental and an incremental program is that we can use an incremental constant.
+This constant is set to zero at the beginning and the solver tries to find a solution of the problem.
+If it can not find a solution it starts to increment the constant by one.
+At each step the solver searches for a solution to the problem and increments the constant until he finds a valid answer set for the logic program.
+The incremental constant can therefore be used among other things to emulate time steps.
+For example, if the solution for a planning problem can not be found in 3 steps, it is tried to find one in more steps.
+To make it concrete, an incremental encoding for the blocksworld problem is presented.
+This is a very simple problem, where $n$ blocks are stacked on top of each other and you can move only one block at the same time. The goal is to stack the blocks in the inverse order.
+Since an incremental logic program is divided into three parts, we start with the base part.
+All rules that hold in all time steps that are independent of the incremental constant are put here.
+At first we declare $n$ blocks and that each block is a location where a block can be put on.
+Furthermore we define an extra location, the table. That is the location where the bottom most block is put.
+\begin{verbatim}
+ #base.
+
+ block(1..n).
+ location(X) :- block(X).
+ location(table).
+\end{verbatim}
+Now we stack all blocks on top of each other, so block 1 is on block 2, block 2 is on block 3 etc.
+The last block we put on the table.
+This all holds in the initial state, so a $0$ is appended to the ``\texttt{on}'' predicate.
+\begin{verbatim}
+ on(X,X+1,0) :- block(X), X < n.
+ on(n,table,0).
+\end{verbatim}
+After declaring the base part and describing the initial situation we have to describe valid moves for the blocks.
+This is done in the cumulative part, since moves can be done in each step.
+For the incremental constant we choose the name ``\texttt{k}''.
+At first we want to execute at least one move every step, but not more than one. This can be done with a cardinality constraint.
+\begin{verbatim}
+ #cumulative k.
+
+ 1{move(X,Y,k) : block(X) : location(Y) : X != Y}1.
+\end{verbatim}
+A move is of the form ``\texttt{move(X,Y,k)}'' where \texttt{X} is the block that is moved, \texttt{Y} the new position of the block and \texttt{k} the step when the moved is executed.
+\newline
+Now that we have exactly one move per step we need to define what happens if we move a block.
+Understandably if we move block \texttt{X} to position \texttt{Y} in step \texttt{k}, the new location of the block shall be \texttt{Y}.
+\begin{verbatim}
+ on(X,Y,k) :- move(X,Y,k).
+\end{verbatim}
+For all other blocks that have not moved, the position shall be the same.
+In ASP a little trick is used to describe such a condition.
+We first define a little helper predicate that states that a block has moved.
+\begin{verbatim}
+ moved(X, k) :- move(X, Y, k).
+\end{verbatim}
+With this helping predicate we can define that a block is on the same position in step \texttt{k}, if it was not moved.
+\begin{verbatim}
+ on(X,Y,k) :- on(X,Y,k-1), block(X), location(Y), not moved(X, k).
+\end{verbatim}
+
+Now we have to restrict our moves, because only one block can be moved at the same time and it can only be moved if no other block is on top of it.
+First we introduce a predicate called ``\texttt{blocked}'' for each block with another block on top of it.
+\begin{verbatim}
+ blocked(Y, k) :- block(Y), on(X, Y, k - 1).
+\end{verbatim}
+Now we can restrict the moves with an integrity constraint.
+\begin{verbatim}
+ :- move(X, Y, k), blocked(X, k).
+\end{verbatim}
+So it is not allowed to move block \texttt{X} in step k, if it is blocked.
+Also you can not move a block to a position where another block is on, so only one block on top of another is allowed.
+\begin{verbatim}
+ :- move(X, Y, k), blocked(Y, k).
+\end{verbatim}
+The description of the problem is nearly finished. All that has to be done is to formulate a query.
+As a constraint to our solution we want to stop if all blocks are on top of each other in reversed order.
+This query shall only hold in the last step, therefore we write it in the volatile part.
+As always for constraints, we use an integrity constraint to express that all solutions are false where there is not block $n+1$ on the block $n$.
+\begin{verbatim}
+ #volatile k.
+
+ :- not on(X+1,X,k), block(X), X < n.
+\end{verbatim}
+To run this example, iclingo has to be used and the constant $n$ has to be set via command line.
+A potential command line call could look like
+\begin{verbatim}
+ iclingo --const n=10 blocksworld.lp
+\end{verbatim}
+and the answer to the problem can be read from the answer set.
+To clean up the answer the ``\texttt{show} and \texttt{hide}'' feature of gringo can be used.
+If the following lines were added to the program, the answer set only consists of \texttt{move/3} predicates.
+\begin{verbatim}
+ #hide.
+ #show move/3.
+\end{verbatim}
 
@@
 \paragraph{\texttt{--lparse,-l}}
-This mode prints lparse conform output. For a definition of the output format see \cite{lparseManual}.
+This mode prints lParse conform output. For a definition of the output format see \cite{lparseManual}.
 Since some features of gringo2.0 are not natively supported for this output language, new atoms may be introduced to create an equivalent program.\newline
-Sum aggregates with negative weights are handled like weight constraint in lparse for this output format.
+Sum aggregates with negative weights are handled like weight constraint in lParse for this output format.
 So a transformation as explained in \cite{lparseManual} is used to get rid of the negative weights.
@@
 \paragraph{\texttt{--bindersplitting=<arg>}}
-With this option you can enable or disable the bindersplitting used by gringo2.0
+With this option you can enable or disable the binder splitting used by gringo2.0
 \begin{itemize}
- \item yes : Enable bindersplitting
- \item no  : Disable bindersplitting
+ \item yes : Enable binder splitting
+ \item no  : Disable binder splitting
 \end{itemize}
@@
 \paragraph{\texttt{--imin=<num>}}
-At least \texttt{num} incremental steps are solved, wether an answer set has been found or not.
+At least \texttt{num} incremental steps are solved, whether an answer set has been found or not.
 \paragraph{\texttt{--imax=<num>}}
@@
 Because trying to ground an already variable free program could need a huge amount of memory.
-
+\item
+\begin{verbatim}
+ "Warning: There are no #base, #lambda or #delta sections."
+\end{verbatim}
+If writing an incremental program, the incremental sections\ref{metasyntaxinc} have to be used.
+\item
+\begin{verbatim}
+ "Warning: There are statements not within a #base, #lambda or #delta section."
+ "         These Statements are put into the #base section."
+\end{verbatim}
+If writing an incremental program, all statements that are not written in an incremental section are automatically considered to belong to the \texttt{base} part of the program.
+\item
+\begin{verbatim}
+ "Warning: Classical negation is not handled correctly in combination with the incremental output."
+ "         You have to add rules like: :- a, -a. on your own! (at least for now)"
+\end{verbatim}
+To use classical negation in an incremental program the described rules have to be added manually. This will be fixed in a later version of gringo.
 \end{itemize}
@@
 The parser can not read the input format. Some unexpected token occurred in the input file.
-\newline
-\newline
+\item
+\begin{verbatim}
+ "sorry clasp cannot handle disjunctive rules!"
+\end{verbatim}
+If using clingo or iclingo, this happens if disjunctive rules are used. Please choose another solver that support disjunctive rules.
+\item
+\begin{verbatim}
+ "A fixed number of incremental steps is needed to ground the program."
+\end{verbatim}
+TODO: I dont know.
+\item
+\begin{verbatim}
+ "cyclic constant definition."
+\end{verbatim}
+TODO: I dont know.
+\item
+\begin{verbatim}
+ "The following statement cant be used with the incremental interface: X."
+ "unsupported statement."
+\end{verbatim}
+TODO: I dont know.
+\item
+\begin{verbatim}
+ "output does not support getIncUid."
+\end{verbatim}
+TODO: I dont know.
+\item
+\begin{verbatim}
+ "comparing different types"
+\end{verbatim}
+TODO: I dont know.
+\item
+\begin{verbatim}
+ "comparing function symbols"
+\end{verbatim}
+TODO: I dont know.
+
+\item
 The next errors all belong to the ASPils output format.
@@
 
-%\subsection{Building Gringo2.0 From Source}
-%For compiling gringo2.0, a binary version of ``re2c'' is needed to compile the lexer and the parser.
-%TODO: What about lemon.
-%As the build environment ``cmake'' is used.
-%TODO: How to build gringo2.0, clingo, iclingo etc...
-
-\section{Syntax}
-The basic elements of a logic program and its syntax is now described.
-Disjunctive logic programs are based on an alphabet \cite{syntax} consisting of:
-\begin{itemize}
-\item a set $\mathcal{P}$ of predicate symbols,
-\item a set $\mathcal{F}$ of function symbols,
-\item a set $\mathcal{V}$ of variables,
-\item a set $\mathcal{C}$ of constants,
-\item a set \{ ``$\mid$'', ``,'' ,$\leftarrow$, not \} of connectives, and
-\item a set \{ ``('', ``)'', ``,'' \} of punctuation symbols.
-\end{itemize}
-
-\paragraph{Predicates and Function Symbols}
-Predicates and function symbols have a predicate symbol/name and an arity.
-For the function symbol $f$ with arity 3 we write $f/3$.
-The names of the function symbols and predicates have to start with either a lower case letter or an underscore and can contain digits and letters.
-With $\mathcal{P}$ we denote to the set of predicates and $\mathcal{F}$ denotes the set of function symbols.
-\paragraph{Constants}
-Constants are function symbols with an arity of zero.
-The name of a constant can also be encompassed by double quotation marks, which is sometimes helpful when using constants with spaces.
-It has to be remarked that a constant with quotation marks is never equal to a constant without, so ``\texttt{a}'' is not equal to \texttt{a}.
-$\mathcal{C}$ is the set of constants.
-\paragraph{Variables}
-Variables have to start with a capitalized letter and can contain digits and letters.
-The set of variables is denoted by $\mathcal{V}$.
-The language over $\mathcal{P}$, $\mathcal{F}$ and $\mathcal{C}$ is denoted by $\mathcal{L(P,F,C)}$.
-\paragraph{Terms}
-The building blocks of the language are terms.
-Each variable $v \in \mathcal{V}$, constant $c \in \mathcal{C}$ and each function symbol $f \in \mathcal{F}$ is a term.
-Furthermore $f(t_1, \dots, t_n)$ is a term if $f/n \in \mathcal{F}$ and $t_i$ for $i \in \{1 \dots n\}$ are terms.
-Since a term is defined recursively, it can be nested.
-A term with the first parameter being a constant $a$, the others being terms $f(b,X)$ and $g(f(X), Y)$ would be expressed this way.
-\begin{verbatim}
- p(a,f(b,X), g(f(X),Y)).
-\end{verbatim}
-\paragraph{Atoms}
-An atom over $\mathcal{L(P, F, C)}$ is of the form $p(t_1, \dots, t_n)$ where $p/n$ is a predicate
-symbol with arity $n$, and $t_i \mid i \in \{1 \dots n\}$ are terms.
-\paragraph{Rules}
-A rule over the language $\mathcal{L(P,F,C)}$ has the form
-$$
-A_0 \mid \dots \mid A_l \leftarrow A_{l+1}, \dots, A_m, not\ A_{m+1}, \dots, not\ A_n.
-$$
-where $A_0$ to $A_n$ are atoms. \newline
-A logic program $\Pi$ over $\mathcal{L(P,F,C)}$ is a finite set of rules over $\mathcal{L(P,F,C)}$.
-
-\subsection{$\lambda$-Restriction}
-\label{domainrestricted}
-There exist two major grounders for ASP besides gringo1.0, lParse \cite{lparseManual} and  the DLV grounding module\cite{DLV}.
-To ensure the existence of an equivalent ground instantiation of a program both grounders except a subclass of logic programs.
-For lParse the logic programs must be $\omega$-restricted\cite{omega} while DLV can handle safe programs\cite{safe}.
-gringo2.0 verifies that the input program is $\lambda$-
-restricted \cite{gringopaper}, so that the existence of a finite equivalent ground instantiation is guaranteed. 
-$\lambda$-restrictedness extends the capabilities of $\omega$-restricted programs.
-The definition of $\lambda$-restrictedness can be found in \cite{gringopaper} and was expanded for aggregates and disjunctions.
-$\lambda$-restrictedness can not be compared with the safe programs of DLV since there exist programs that are safe and not $\lambda$-restricted and vice versa.
-\newline
-To give a gross impression of what $\lambda$-restriction means for writing a logic program an informal description is given.
-In gringo2.0 every variable used in a rule must have a complete domain.
-This means that this variable has to occur in at least one predicate whose domain is completed.
-Therefore the predicate must be in the head of rules that are already domain restricted.
-So the grounder has to find a valid order of the rules so that every predicate in the head is completed before it is used as a domain predicate for the next rules.
-To make things more clear, an example is given
-\begin{verbatim}
- zig(0) :- not zag(0).   zag(0) :- not zig(0).
- zig(1) :- not zag(1).   zag(1) :- not zig(1).
- zigzag(X,Y) :- zig(X), zag(Y).
-\end{verbatim}
-The rules of this example can be ordered in the following way.
-At first, the rules with \texttt{zig/1} and \texttt{zag/1} are considered.
-Since these are the only rules \texttt{zig/1} and \texttt{zag/1} occurs in the head, the domains of the predicates are closed.
-The domains \{\texttt{zig(0)}, \texttt{zig(1)}\} and \{ \texttt{zag(0)}, \texttt{zag(1)}\} are complete, since no other predicates of the form \texttt{zig/1} or \texttt{zag/1} can be derived from the program.
-Now the third rule with \texttt{zigzag/2} in the head can be grounded.
-The variables \texttt{X} and \texttt{Y} both occur in a domain predicate, and the values $0$ respectively $1$ can substitute the variables.
-\newline
-To give a comparison to the $\omega$-restricted programs of lparse, an extra domain predicate has to be introduced to the example.
-\begin{verbatim}
- domain(0..1).
- zig(0) :- not zag(0).   zag(0) :- not zig(0).
- zig(1) :- not zag(1).   zag(1) :- not zig(1).
- zigzag(X,Y) :- zig(X), zag(Y), domain(X), domain(Y).
-\end{verbatim}
-\texttt{zig/1} and \texttt{zag/1} can not be used as a domain predicate, since their truth value can not be fully evaluated.
-
-%\paragraph{Function Symbols}
-%A constant is a representation of an entity in the problem domain.
-%A constant is either a name, starting with a lower case letter or an underscore, or a positive integer.
-%It also can be encompassed by double quotation marks.
-When using quotation marks, which is sometimes helpful when using %constants with spaces, it has to be remarked that a constant with quotation marks is never equal to a constant without, so ``$a$'' is not $a$.
-%\paragraph{Variable}
-%A variable is programming construct to simplify writing logic programs.
-%It always starts with a capitalized letter and can contain digits and letters.
-%In the grounding process a variable is replaced by one or more constants.
-%Each variable occurring in a rule must be domain restricted as described in section \ref{domainrestricted}.
-%\paragraph{Atom}
-%An atom is a predicate with a name. The name can consist of letters, digits and the underscore.
-%It has to start with a lower case letter or an underscore.
-%An atom can have several parameters, encapsulated by parenthesis and separated with commas.
-%The parameters are either constants, variables or function symbols.
-%Function symbols do have the same syntax as atoms. This way they can be finitely nested.
-%As an example, an atom with three parameters is given:
-%The first parameter being a constant $a$, the others being function symbols $f(b,X)$ and $g(f(X), Y))$.
-%\begin{verbatim}
-% p(a,f(b,X), g(f(X),Y)).
-%\end{verbatim}
-
-%\paragraph{Rule}
-%A rule normally consists of a head and a body and ends with a dot.
-%The body is separated from the head using ``:-'' as in
-%\begin{verbatim}
-% head :- body.
-%\end{verbatim}
-%The body is a comma separated conjunction of expressions.
-%The supported expressions are introduced in the next section.
-%The variables used in the head must be restricted to a domain in the body.
-%Check (\ref{domainrestricted}) to see the restrictions on domains.
-%\newline
-%A rule can occur without a body as a fact, therefore the body is omitted.
-%\begin{verbatim}
-% head.
-%\end{verbatim}
-%The same can be done with the body to express an integrity constraint.
-%\begin{verbatim}
-% :- body.
-%\end{verbatim}
-
-\subsection{Basic Language}
-The base building blocks of the language are rules, facts and integrity constraints.
-\paragraph{Rule}
-A rule normally consists of a head and a body and ends with a dot.
-The body is separated from the head using ``:-'' as in
-\begin{verbatim}
- head :- b1, ..., bN, not c1, ..., cN.
-\end{verbatim}
-The head consists of terms concatenated by a disjunction. For non disjunctive rules only one term is used.
-The body is a comma separated conjunction of terms and their negation.
-The variables used must be restricted to a domain in the body.
-Check (\ref{domainrestricted}) to see the restrictions on domains.
-\paragraph{Fact}
-A rule can occur without a body as a fact, therefore the body is omitted.
-\begin{verbatim}
- head.
-\end{verbatim}
-\paragraph{Integrity Constraint}
-To express an integrity constraint, a rule with an empty body is used.
-\begin{verbatim}
- :- body.
-\end{verbatim}
-
-\subsection{Advanced Language Features}
-To ease the process of writing logic programs some advanced features in the input syntax are introduced and are now explained in detail.
-For the most problems it is possible to write a much shorter version of the program with the extended syntax.
-%Even a speedup in solving the instances of the problem can happen TODO ?
-For a full syntax description in EBNF see appendix (\ref{appendix}).
-\paragraph{Function Symbols}
-Function symbols are fully supported with gringo2.0.
-They can be finitely nested and also be used as domain predicates.
-While in lParse function symbols are allowed in a very restricted way, no restrictions apply in gringo2.0.
-The following program can not be grounded with lParse since $X$ is an unrestricted variable. With gringo2.0 the program is now groundable.
-\begin{verbatim}
-a(X) :- f(1,g(X)).
-f(1,X) :- f(X).
-f(g(2)).
-\end{verbatim}
-From the second rule the domain \texttt{f(1,g(2))} is generated.
-This domain can be used for the first rule and results in the grounded rule
-\begin{verbatim}
- a(2) :- f(1,g(2)).
-\end{verbatim}
-
-\paragraph{Range Operator}
-To define several atoms that stretch on one or more parameters over a certain range, two dots can be used to express an increasing integer range.
-Using the range operator in the head of a rule, it works as a conjunction of all the members in the range, so for every number the rule is duplicated.
-\begin{verbatim}
- times(1..3, 1..2).
-\end{verbatim}
-will result in the facts
-\begin{verbatim}
- times(1,1). times(2,1). times(3,1).
- times(1,2). times(2,2). times(3,2).
-\end{verbatim}
-
-Using the range operator in the body of a rule, a disjunction of all members in the range is created. So the rules will be duplicated for every member.
-\begin{verbatim}
- a :- p(1..3).
-\end{verbatim}
-will result in the rules
-\begin{verbatim}
- a :- p(1).
- a :- p(2).
- a :- p(3).
-\end{verbatim}
-If the range operator is used in an aggregate, it creates a comma separated list of atoms, so
-\begin{verbatim}
- a :- {b(1..3)}.
-\end{verbatim}
-results in
-\begin{verbatim}
- a :- {b(1), b(2), b(3)}.
-\end{verbatim}
-
-\paragraph{Expander :}
-With the expander ``:'' a domain can be set.
-On the left side of the operator an atom with variables is expected and on the right side the domains for the variables are given.
-In the head, the atoms get connected via a disjunction, in the body they get connected via a conjunction, so no new rules are created with this operator.
-\begin{verbatim}
- a(X,Y) : b(X) : c(Y) :- d(X) : e(X).
- b(1..2).
- c(3).
- e(a;b;c).
-\end{verbatim}
-would become
-\begin{verbatim}
- a(1,3) | a(2,3) :- d(a), d(b), d(c).
-\end{verbatim}
-This operator can be used in aggregates to create a comma separated list of atoms.
-For example
-\begin{verbatim}
- n(1..9).
- 1{a(X) : n(X) : X > 3}1.
-\end{verbatim}
-would become
-\begin{verbatim}
- 1 {a(4), a(5), a(6), a(7), a(8), a(9)} 1.
-\end{verbatim}
-
-\paragraph{Conjunctor ;}
-The conjunctor ``;'' will create a conjunction of its elements.
-Used in the head of a rule means that the rule is duplicated.
-If used in the body instead, the atoms will be logically combined using a conjunction.
-\begin{verbatim}
- a(X;1..2) :- b(X;a;b).
- b(foo). b(bar).
-\end{verbatim}
-will results in the rules\footnote{gringo2.0 would simplify the grounded rules since some literals in the body are facts}
-\begin{verbatim}
- a(foo,1) :- b(foo), b(a), b(b).
- a(foo,2) :- b(foo), b(a), b(b).
- a(bar,1) :- b(bar), b(a), b(b).
- a(bar,2) :- b(bar), b(a), b(b).
-\end{verbatim}
-The conjunctor can not be used directly in aggregates, but for example in combination with the expander.
-\begin{verbatim}
- { a(X,Y) : p(X;Y) }.
- p(1..2).
-\end{verbatim}
-Results in
-\begin{verbatim}
- {a(1,1), a(1,2), a(2,1), a(2,2)}.
- p(1).
- p(2).
-\end{verbatim}
-
-\paragraph{Arithmetic Expressions}
-The built-in arithmetics can be used for integer arithmetic.
-As operators addition (+), subtraction (-), multiplication (*), division (/), modulo (mod) and the absolute value operator (abs) can be used.
-Furthermore the bitwise operators and(\&), or(?), xor(\textasciicircum) and invert($\sim$) are supported.
-The bitwise operators are executed on signed integers using the binary complement for negative numbers.
-Arithmetic expressions can occur everywhere in the head and in the body, even in complex constructs.
-As parameters, integer constants or expressions that evaluate to integers are used.
-The simple example of using operators and range terms
-\begin{verbatim}
- a(X-1 .. X+1) :- b(X,Y), c(abs(Y * 2 * -X)).
- b(1,2).
- c(4).
-\end{verbatim}
-would result in the grounded program
-\begin{verbatim}
- a(0) :- b(1,2), c(4).
- a(1) :- b(1,2), c(4).
- a(2) :- b(1,2), c(4).
- b(1,2).
- c(4).
-\end{verbatim}
-
-\subparagraph{Relational Operators}
-The relational operators are ``less than'' ($<$),\\ ``greater than'' ($>$), ``less equal'' ($<=$), ``greater equal'' ($>=$) , ``equal'' ($==$) and ``not equal'' ($!=$).
-They can be used to constrain certain variables.
-For example, the following rule is only generated for all unifications where $X$ is greater or equal $Y-1$.
-\begin{verbatim}
- a(X) :- b(X,Y), X >= Y-1.
- b(0,2).
- b(4,2).
-\end{verbatim}
-This would let $\texttt{b(4,2)}$ be a valid substitution, since $4 >= 2-1$ but $\texttt{b(0,2)}$ not, so the grounded program could be:
-\begin{verbatim}
- a(4) :- b(4,2).
- b(0,2).
- b(4,2).
-\end{verbatim}
-
-\subparagraph{Assignment Operator}
-As assignment operator ``$=$'' is used.
-The variable on the left side is set to the value of the right side.
-\begin{verbatim}
- a(X) :- X=Y*2, b(Y).
- b(1..2).
-\end{verbatim}
-becomes
-\begin{verbatim}
- a(2) :- b(1).
- a(4) :- b(2).
- b(1..2).
-\end{verbatim}
-A cyclic assignment that results in an unbound variable is not allowed and is rejected with an error . For example
-\begin{verbatim}
- a(X) :- X=Y, Y=X.
-\end{verbatim}
-Furthermore the assignment operator can be used with aggregates. The program
-\begin{verbatim}
- price(apple,5). prices(banana, 6). price(computer, 100).
- fruit(apple).
- fruit(banana).
- max_cost(1..200).
- costs(C) :- C = [price(X,Y) : fruit(X) : max_cost(Y) = Y].
-\end{verbatim}
-would lead to the fact \texttt{costs(11)} as the price of all fruits.
-$C$ in this example becomes all values that have to be generated for the aggregate.
-
-\paragraph{Classical Negation}
-For classical negation the atom is preceded by a minus sign.
-\begin{verbatim}
- -a :- not  a.
-  a :- not -a. 
-\end{verbatim}
-\paragraph{Disjunction}
-A disjunction can be expressed using the pipe symbol ``$\mid$''.
-A disjunction can only occur in the head of a rule.
-\begin{verbatim}
- red | green | blue :- colored.
-\end{verbatim}
-Using disjunctions in logic programs can increase expressiveness and complexity of the logic program.
-A solver for disjunctive logic programs such as claspD \cite{claspD} or cmodels \cite{cmodels} is needed to solve such programs.
-
-\paragraph{Aggregates}
-An aggregate is a set or a multiset of atoms and has a lower and an upper bound.
-The general syntax is
-\begin{verbatim}
- lower aggregate_name [atom1=w1, ..., atomN=wN] upper
-\end{verbatim}
-If squared brackets are used, the atom list is interpreted as a multiset, otherwise, for curly brackets, as a set of atoms.
-The lower and upper bound is optional, and can be omitted.
-
-\subparagraph{Count}
-A ``count'' aggregate is a set of atoms and has an optional lower and upper bound.
-It evaluates to true if the number of true atoms in the set is between the lower and upper bound. Only curly brackets may be used for a count aggregate as it represents a set of atoms.
-Weights are not allowed in the count aggregate.
-For this aggregate a shorthand syntax exists
-\begin{verbatim}
- lower {atom1, ..., atomN} upper
-\end{verbatim}
-The count aggregate is equal to the cardinality constraint known from smodels \cite{smodels}.
-\subparagraph{Sum}
-The ``sum'' aggregate is a multiset of weighted atoms and has an optional lower and upper bound.
-It evaluates to true if the sum of the weights of all true atoms in the multiset is between the lower an upper bound.
-The sum aggregate is equal to the weight constraints known from smodels, and has therefore a shorthand syntax
-\begin{verbatim}
- lower [atom1=w1, ..., atomN=wN] upper
-\end{verbatim}
-If the command line option ``-l'' is used the semantic may change if negative weights exist in the aggregate. The transformation used is described in \cite{lparseManual}.
-
-\subparagraph{Min}
-The ``min'' aggregate is determined by the minimum of all weights of atoms that evaluate to true.
-For the minimum aggregate a multiset with weighted atoms is used. It has no shorthand syntax.
-\subparagraph{Max}
-The ``max'' aggregate is similar to the minimum aggregate but considers the maximum of the weights.
-
-\paragraph{Optimize Statements}
-
-\subparagraph{Minimize/Maximize}
-An optimize statement is used to find a minimal or maximal answer set of the logic program.
-The statement can be weighted or not, therefore weights can be omitted in the following syntax.
-\begin{verbatim}
- minimize [atom1 = w1, ..., atomN = wN].
-\end{verbatim}
-The maximize statement uses the keyword ``maximize'' instead.
-If there were several minimize/maximize statements in the logic program the latter will be preferenced.
-Only after the n'th statement results in the same value for two different answer set, the n-1'th statement is considered.
-
-\subsection{Meta-Syntax}
-This section is about constructs that do not originally belong to the logic program but can be used to give additional information for example to the solver.
-The ``\#'' sign in front of all meta expressions is optional but is it strongly recommend to use it.
-It separates the meta directives from the logic program more clearly and can prevent severe mistakes,
- confusing those constructs with atoms as shown in the example.
-\begin{verbatim}
- hide.
- hidden :- hide.
-\end{verbatim}
-This program hides all predicates from the output of the ASP solver instead of deriving the facts \texttt{hide} and \texttt{hidden}.
-
-\paragraph{hide and show}
-The hide construct is used to hide atoms from the output of the ASP solver.
-Two different syntaxes are supported by gringo2.0.
-The first 
-\begin{verbatim}
- #hide name/arity.
-\end{verbatim}
-and the second
-\begin{verbatim}
- #hide name(X1, ... , Xn).
-\end{verbatim}
-Both variants hide the predicate ``\texttt{name}''.
-The arity of the predicate can be given explicit or implicit with a list of variables.
-The variables are just placeholder and do not unify with other variables.
-For example \texttt{hide p(X,X)} would also hide the fact \texttt{p(a,b)}.
-The second syntax is implemented for compatibility reasons only.
-If the predicate that has to be hidden has arity zero the syntax is
-\begin{verbatim}
- #hide name.
-\end{verbatim}
-If you want to express that all but a few predicates shall be hidden the empty hide statement
-in conjunction with the ``\texttt{show}'' statement can be is used.
-With ``\texttt{show}'' all predicates are named that shall be visible.
-So to hide all predicates but $p/3$ the following example can be used.
-\begin{verbatim}
- #hide.
- #show p/3.
-\end{verbatim}
-
-\paragraph{const}
-Within the ``\texttt{const}'' statement, a default value for constants can be defined in the logic program.
-while they are normally defined via the command line options.
-A default value for a constant $n$ can be easily be set.
-\begin{verbatim}
- #const n=5.
-\end{verbatim}
-\paragraph{domain}
-A domain statement states that a specific variable has a global domain restriction.
-The statement
-\begin{verbatim}
- #domain time(T).
-\end{verbatim}
-makes sure that the variable ``\texttt{T}'' is always bound to the domain of ``\texttt{time}''.
-This has the same effect as adding the domain ``\texttt{time(T)}'' to every rule the variable ``\texttt{T}'' occurs in.
-\paragraph{compute}
-The compute statement is a hint for the solver which models to compute.
-It has the syntax
-\begin{verbatim}
- #compute number {a1, ..., aN, not aN+1, ..., not aM}.
-\end{verbatim}
-$number$ is the number of models that shall be computed by the answer set solver and can be omitted. If it is zero, no restriction to the number of models is applied.
-All positive atoms in the compute statement must be in the computed models, and all negated atoms do not have to be in the models of the program.
-The compute statement is solver dependent and not supported by every solver.
-\paragraph{comments}
-Two kinds of comments were supported by gringo2.0.
-Line based comments start with a ``\texttt{\%}''.
-Whenever a ``\texttt{\%}'' occurs in the input, the rest of the line is treated as a comment and is therefore omitted.
-Block based comments can span over several lines.
-Everything between ``\texttt{\%*}'' and ``\texttt{*\%}'' is treated as a comment.
-\subsection{Meta-Syntax for Incremental Programs}
-Incremental programs must be splitted into 3 parts.
-The base part, for the information that hold in all steps and is independent of the incremental constant.
-The cumulative part, where the cumulative rules must be set that hold in all steps, but depend on the incremental parameter.
-And finally the volatile part. In this part the queries to the program are given, that must hold only in the last state.
-To seperate the parts, three statements exists. Everything beyond such a statement is considered to belong to the specific part.
-To start the base part
-\begin{verbatim}
- #base.
-\end{verbatim}
-must be written. Now all rules of the base part can follow.
-The use of the incremental constant is not allowed in this part.
-For the cumulative part an additional constant has to be passed to the statement.
-\begin{verbatim}
- #cumulative k.
-\end{verbatim}
-Now all incremental rules follow. The incremental constant used for grounding is $k$.
-The same holds for the volatile part.
-\begin{verbatim}
- #volatile k.
-\end{verbatim}
-The ``\#''-sign can not be omitted for these meta statements.
-These statements can also be mixed, only the last statement is important.
-If no statement is given at the beginning of the file, base is assumed.
-
-\section{Examples}
-In this section some encodings for common problems are presented and explained.
-
-\subsection{How to Write a Logic Program}
-Writing a logic program to solve a problem is rather describing a problem than describing an algorithm to solve a problem.
-Therefore the problem must be encoded as a set of rules.
-Typically these are divided into a problem description and an instance of the problem.
-The solution to a problem can then be read from the answer sets of the program.
-The procedure of writing a logic program shall now be explained with a little example.
-We want to write a problem encoding for the 3-colorability problem of graph coloring.
-Given an arbitrary graph, no two connected edges may have the same color.
-The most common technique is the ``generate \& test'' approach.
-Therefore, in the generator, rules are created that can derive a superset of the solution to the problem.
-In the tester all solutions that do not fit the constraints are omitted.
-\newline
-Usually you start with the generator.
-At first, we define the non instance domains, in our example the colors we use.
-\begin{verbatim}
- color(red). color(green). color(blue).
-\end{verbatim}
-Now we have to color all the vertices, but we must give attention that each vertex has only one color.
-We can do this by simply saying that a vertex $V$ is colored with one color $C$, and that the same vertex does not have any of the other colors $D$ or $E$.
-\begin{verbatim}
- colored(V,C) :- not colored(V,D), not colored(V,E),
-                 C != D, C != E, D != E,
-                 color(E), color(C), color(D), vertex(V).
-\end{verbatim}
-As you can see, you have to give the domain of the variables in the rule.
-The variable $V$ is used for vertices and the variables $C$, $D$ and $E$ are used for colors.
-After every vertex now has a color we would get all possible solutions of all vertices colorfully painted.
-We now need to restrict our solutions in the tester by saying that two vertices that are connected by an edge should not have the same color.
-We can do this by using an integrity constraint.
-An integrity constraint is a rule with an empty head, and if an assignment of variables fulfills the body, the assignment is not a solution of our problem.
-\begin{verbatim}
- :- edge(V,U), colored(V,C), colored(U,C).
-\end{verbatim}
-Here we have the edge between two vertices $V$ and $U$, and if both are colored with the same color $C$ the assignment is not a solution to our problem.
-\newline
-We have finally written our problem description, all we need now is a problem instance. This step is fairly simple.
-We just need to define several vertices and edges that do connect the vertices.
-For the following simple graph
-\[
- \xymatrix{1\ar@{-}[dr]\ar@{-}[d]\ar@{-}[r]&2\ar@{-}[dl]\\
-           3&4}
-\]
-
-we could just write
-\begin{verbatim}
- vertex(1..4).
- edge(1,2). edge(1,4).
- edge(3,1). edge(3,2).
-\end{verbatim}
-
-If we give the problem description and the instance to the grounder, it will generate a variable free version of the problem.
-This can then be solved by a solver and in the answer sets of the solver the ``\texttt{colored/2}'' predicate states a valid coloring of all vertices.
-We can do this by simply using the internal interface to clasp, to get a solution.
-\begin{verbatim}
- ./gringo -c 3colorability.lp instance1.lp
-\end{verbatim}
-In the first answer set we find the ``\texttt{colored}'' predicates
-\begin{verbatim}
- colored(1,green) colored(2,red) colored(3,blue) colored(4,red)
-\end{verbatim}
-and we can interpret them as the colors of the vertices
-
-\[
- \xymatrix{\color{green}1\ar@{-}[dr]\ar@{-}[d]\ar@{-}[r]&\color{red}2\ar@{-}[dl]\\
-           \color{blue}3&\color{red}4}
-\]
-To improve the readability of the answerset, all but the \texttt{colored/2} predicates can be hidden with these two lines.
-\begin{verbatim}
- #hide.
- #show colored/2.
-\end{verbatim}
-
-\subsection{How to Write an Incremental Program}
-As in writing a non incremental logic problem, we describe a problem rather than an algorithm.
-The difference between a non incremental and an incremental program is that we can use an incremental constant.
-This constant is set to zero at the beginning and the solver tries to find a solution of the problem.
-If it can not find a solution it starts to increment the constant by one.
-At each step the solver searches for a solution to the problem and increments the constant until he finds a valid answer set for the logic program.
-The incremental constant can therefore be used among other things to emulate time steps.
-For example, if the solution for a planning problem can not be found in 3 steps, it is tried to find one in more steps.
-To make it concrete, an incremental encoding for the blocksworld problem is presented.
-This is a very simple problem, where $n$ blocks are stacked on top of each other and you can move only one block at the same time. The goal is to stack the blocks in the inverse order.
-Since an incremental logic program is divided into three parts, we start with the base part.
-All rules that hold in all timesteps that are independent of the incremental constant are put here.
-At first we declare $n$ blocks and that each block is a location where a block can be put on.
-Furthermore we define an extra location, the table. That is the location where the bottom most block is put.
-\begin{verbatim}
- #base.
-
- block(1..n).
- location(X) :- block(X).
- location(table).
-\end{verbatim}
-Now we stack all blocks on top of each other, so block 1 is on block 2, block 2 is on block 3 etc.
-The last block we put on the table.
-This all holds in the initial state, so a $0$ is appended to the ``\texttt{on}'' predicate.
-\begin{verbatim}
- on(X,X+1,0) :- block(X), X < n.
- on(n,table,0).
-\end{verbatim}
-After declaring the base part and describing the initial situation we have to describe valid moves for the blocks.
-This is done in the cumulative part, since moves can be done in each step.
-For the incremental constant we choose the name ``\texttt{k}''.
-At first we want to execute at least one move every step, but not more than one. This can be done with a cardinality constraint.
-\begin{verbatim}
- #cumulative k.
-
- 1{move(X,Y,k) : block(X) : location(Y) : X != Y}1.
-\end{verbatim}
-A move is of the form ``\texttt{move(X,Y,k)}'' where \texttt{X} is the block that is moved, \texttt{Y} the new position of the block and \texttt{k} the step when the moved is executed.
-\newline
-Now that we have exactly one move per step we need to define what happens if we move a block.
-Understandably if we move block \texttt{X} to position \texttt{Y} in step \texttt{k}, the new location of the block shall be \texttt{Y}.
-\begin{verbatim}
- on(X,Y,k) :- move(X,Y,k).
-\end{verbatim}
-For all other blocks that have not moved, the position shall be the same.
-In ASP a little trick is used to describe such a condition.
-We first define a little helper predicate that states that a block has moved.
-\begin{verbatim}
- moved(X, k) :- move(X, Y, k).
-\end{verbatim}
-With this helping predicate we can define that a block is on the same position in step \texttt{k}, if it was not moved.
-\begin{verbatim}
- on(X,Y,k) :- on(X,Y,k-1), block(X), location(Y), not moved(X, k).
-\end{verbatim}
-
-Now we have to restrict our moves, because only one block can be moved at the same time and it can only be moved if no other block is on top of it.
-First we introduce a predicate called ``\texttt{blocked}'' for each block with another block on top of it.
-\begin{verbatim}
- blocked(Y, k) :- block(Y), on(X, Y, k - 1).
-\end{verbatim}
-Now we can restrict the moves with an integrity constraint.
-\begin{verbatim}
- :- move(X, Y, k), blocked(X, k).
-\end{verbatim}
-So it is not allowed to move block \texttt{X} in step k, if it is blocked.
-Also you can not move a block to a position where another block is on, so only one block on top of another is allowed.
-\begin{verbatim}
- :- move(X, Y, k), blocked(Y, k).
-\end{verbatim}
-The description of the problem is nearly finished. All that has to be done is to formulate a query.
-As a constraint to our solution we want to stop if all blocks are on top of each other in reversed order.
-This query shall only hold in the last step, therefore we write it in the volatile part.
-As always for constraints, we use an integrity constraint to express that all solutions are false where there is not block $n+1$ on the block $n$.
-\begin{verbatim}
- #volatile k.
-
- :- not on(X+1,X,k), block(X), X < n.
-\end{verbatim}
-To run this example, iclingo has to be used and the constant $n$ has to be set via command line.
-A potential command line call could look like
-\begin{verbatim}
- iclingo --const n=10 blocksworld.lp
-\end{verbatim}
-and the answer to the problem can be read from the answer set.
-To clean up the answer the ``\texttt{show} and \texttt{hide}'' feature of gringo can be used.
-If the following lines were added to the program, the answer set only consists of \texttt{move/3} predicates.
-\begin{verbatim}
- #hide.
- #show move/3.
-\end{verbatim}
-%\paragraph{Fastfood}
-%This example needs a proper instance to work.
-%\begin{verbatim}
-%#hide.
-%#show altdepot/2.
-%#show costs/1.
-%
-%% calculate the number of depots
-%depots(X) :- X = { depot(DN, DK) :  depot(DN, DK) }.
-%
-%% calculate the costs for the initial depots
-%% (the grounder can calculate this on its own)
-%costs(C) :- C = [ serves(RN, Dist) : restaurant(RN, RK) : dist(Dist) = Dist ].
-%
-%% choose the closest depot for a restaurant
-%% (the grounder can calculate this on its own)
-%serves(Rname,Dist) :- restaurant(Rname,RK),
-%   Dist = min [ depot(DN1, DK1) : depot(DN1, DK1) = abs(DK1-RK) ].
-%
-%dist(abs(X-Y)) :- restaurant(A, X), restaurant(B, Y).
-%
-%% Each restaurant may be an alternative depot or not.
-%X { altdepot(R,K) : restaurant(R,K) } X :- depots(X).
-%
-%% choose the closest depot for a restaurant
-%altserves(Rname,Dist) :- restaurant(Rname,RK),
-%   Dist = min [ altdepot(DN1, DK1) : restaurant(DN1, DK1) = abs(DK1-RK) ].
-%
-%% fail if we have not found a better solution
-%:- costs(C), C [ altserves(R,Dist) : dist(Dist) : restaurant(R,RK) = Dist ].
-%
-%% thats the easiest way to show the calculated costs
-%minimize [ altserves(R,Dist) : dist(Dist) : restaurant(R,RK) = Dist ].
-%\end{verbatim}
-%
-%\paragraph{Queens}
-%\begin{verbatim}
-%#hide.
-%#show q(X,Y).
-%
-%% this can be set via the command line
-%#const queens = 200.
-%
-%#const diags  = queens * 2 - 1.
-%d(1..queens).
-%diag(1..diags).
-%
-%% enumerate the diagonals
-%diagX(X,Y,X - Y + queens) :- d(X), d(Y).
-%diagY(X,Y,X + Y - 1) :- d(X), d(Y).
-%
-%% exactly 1 queen in each row/col
-%1 { q(X,Y) : d(Y) } 1 :- d(X).
-%1 { q(X,Y) : d(X) } 1 :- d(Y).
-%
-%% at most 1 queen on each diagonal
-%{ q(X,Y) : diagX(X,Y,D) } 1 :- diag(D).
-%{ q(X,Y) : diagY(X,Y,D) } 1 :- diag(D).
-%
-%\end{verbatim}
-
-%\section{Benchmarks}
-%The benchmarks were run on an Athlon XP 1700+ with 1GB of RAM.
-%TODO: Official (non internal wiki) link to the instances.
-
-%\input{benchmark2.result}
 \section{Future Work}
@@
 \appendix
-%todo: input grammar in EBNF
 \end{document}

### 2008-09-04 lib/gringo/src/claspoutput.cpp

@@ -161,7 +161,7 @@ void ClaspOutput::printMinimizeRule(const IntVector &pos, const IntVector &neg, 
 
 void ClaspOutput::printDisjunctiveRule(const IntVector &head, const IntVector &pos, const IntVector &neg)
 {
-	throw GrinGoException("error: sorry clasp cannot handle disjunctive rules!");
+	throw GrinGoException("Error: sorry clasp cannot handle disjunctive rules!");
 }
 
 void ClaspOutput::printComputeRule(int models, const IntVector &pos, const IntVector &neg)

### 2008-09-04 lib/gringo/src/grounder.cpp

@@ -187,7 +187,7 @@ void Grounder::ground()
 	else
 	{
 		if(incParts_.size() > 0 && opts_.ifixed < 0)
-			throw GrinGoException("Error: A fixed number of incremtal steps is needed to ground the program.");
+			throw GrinGoException("Error: A fixed number of incremental steps is needed to ground the program.");
 		if(incParts_.size() > 0)
 		{
 			do
@@ -387,7 +387,7 @@ void Grounder::addTrueNegation(int id, int arity)
 		if(!warn)
 			return;
 		warn = false;
-		std::cerr << "Warning: Classical negation is not handled correctly in combination with the icremental output." << std::endl;
+		std::cerr << "Warning: Classical negation is not handled correctly in combination with the incremental output." << std::endl;
 		std::cerr << "         You have to add rules like: :- a, -a. on your own! (at least for now)" << std::endl;
 		return;
 	}

### 2008-09-04 doc/documentation/paper.tex
Doku: added explanation of new errors and warnings
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@492 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 \end{verbatim}
-TODO: I dont know.
+This error occurs if an incremental program is given to gringo without using the option \texttt{--ifixed} or iclingo.
 \item
@@
 \end{verbatim}
-TODO: I dont know.
+A cyclic definition of a constant is not allowed as shown in the example.
+\begin{verbatim}
+ #const x=y.
+ #const y=x.
+\end{verbatim}
 \item
@@
 \end{verbatim}
-TODO: I dont know.
-\item
-\begin{verbatim}
- "output does not support getIncUid."
-\end{verbatim}
-TODO: I dont know.
+It is not allowed to use an optimize (\texttt{minimize} or \texttt{maximize}) or \texttt{compute} statement in an incremental program.
 \item
@@
  "comparing different types"
-\end{verbatim}
-TODO: I dont know.
-\item
-\begin{verbatim}
  "comparing function symbols"
 \end{verbatim}
-TODO: I dont know.
+This error occurs if trying to compare non compareable values or function symbols as in
+\begin{verbatim}
+ p :- 1 < s.
+ p :- f(x) < f(y).
+\end{verbatim}
 

### 2008-09-04 doc/documentation/paper.tex
Doku: removed all /hbox's
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@493 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 An atom over $\mathcal{L(P, F, C)}$ is of the form $p(t_1, \dots, t_n)$ where $p/n$ is a predicate
-symbol with arity $n$, and $t_i \mid i \in \{1 \dots n\}$ are terms.
+symbol with arity $n$, and $t_i$ for $i \in \{1 \dots n\}$ are terms.
 \paragraph{Rules}
@@
 \begin{verbatim}
- on(X,Y,k) :- on(X,Y,k-1), block(X), location(Y), not moved(X, k).
+ on(X,Y,k) :- on(X,Y,k-1), block(X), location(Y), not moved(X,k).
 \end{verbatim}
@@
 \paragraph{\texttt{--ground,-g}}
-If this parameter is used, the grounder expects an already grounded program as input.
-For converting an already grounded logic program into an output format supported by gringo2.0 this is the right option. This mode saves a lot of memory but does not ground the program.
+If this parameter is used, gringo expects an already grounded program as input.
+For converting an already grounded logic program into an output format 
+supported by gringo2.0 this is the right option.
+ This mode saves a lot of memory but does not ground the program.
 \paragraph{\texttt{--const,-c C=v}}
@@
 \paragraph{Default Parameters}
-The default command line is ``\texttt{./gringo --lparse --bindersplitting=yes}'' to read from standard input and for smodels output.
+The default command line is
+\begin{verbatim}
+ gringo --lparse --bindersplitting=yes
+\end{verbatim}
+to read from standard input and for smodels output.
 
@@
 \begin{verbatim}
-  iclingo 1 --incremental --istop=SAT --ilearnt=forget --iheuristic=keep
-            --trans-ext=no --eq=5 
+ iclingo 1 --incremental --istop=SAT --ilearnt=forget 
+           --iheuristic=keep --trans-ext=no --eq=5 
            --lookahead=no --lookback=yes --heuristic=Berkmin
@@
 \begin{verbatim}
- "Warning: There are statements not within a #base, #lambda or #delta section."
+ "Warning: There are statements not within a #base, #lambda"
+ "         or #delta section."
  "         These Statements are put into the #base section."
@@
 \begin{verbatim}
- "Warning: Classical negation is not handled correctly in combination with the incremental output."
- "         You have to add rules like: :- a, -a. on your own! (at least for now)"
+ "Warning: Classical negation is not handled correctly in"
+ "         combination with the incremental output."
+ "         You have to add rules like:"
+ "         :- a, -a. on your own! (at least for now)"
 \end{verbatim}
@@
 \begin{verbatim}
- "A fixed number of incremental steps is needed to ground the program."
+ "A fixed number of incremental steps is needed"
+ "to ground the program."
 \end{verbatim}
@@
 \begin{verbatim}
- "The following statement cant be used with the incremental interface: X."
+ "The following statement cant be used with the"
+ "incremental interface: X."
  "unsupported statement."

### 2008-09-04 app/options.cpp
- implemented the option --iquery   it allows to specify when to start grounding the queries - removed the option --ibase   i think block comments will do the job
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@494 3e02f596-3131-0410-8352-e04eeaec37cc

@@ -213,7 +213,6 @@ void Options::setDefaults() {
 	keepLearnts     = true;
 	keepHeuristic   = false;
 	iclaspOut       = false;
-	ibase           = false;
 	imin            = 1;
 	imax            = std::numeric_limits<int>::max();
 	iunsat          = false;
@@ -279,7 +278,7 @@ void Options::initOptions(ProgramOptions::OptionGroup& allOpts, ProgramOptions::
 		("imin"        , value<int>(&imin)    , "Minimum number of incremental steps", "<num>")
 		("imax"        , value<int>(&imax)    , "Maximum number of incremental steps", "<num>")
 		("incremental,i"   , bool_switch(&iclaspOut), "Use incremental clasp interface")
-		("ibase"       , bool_switch(&ibase)  , "(not implemented yet)")
+		("iquery"      , value<int>(&grounderOptions.iquery)  , "Start solving beginning with step num", "<num>")
 		("istop"      , value<bool>(&iunsat)->parser(mapStop), "Stop condition during incremental solving\n"
 			"\tDefault: SAT\n"
 			"\t  UNSAT : Stop if no solution found\n"
@@ -482,6 +481,8 @@ void Options::checkCommonOptions(const OptionValues& vm) {
 #endif
 	if(textOut)
 		outf = TEXT_OUT;
+
+	grounderOptions.iquery = max(grounderOptions.iquery, 0);
 
 	if(grounderOptions.ifixed >= 0 && outf == ICLASP_OUT)
 	{

### 2008-09-04 lib/gringo/include/grounder.h

@@ -35,8 +35,9 @@ namespace NS_GRINGO
 	public:
 		struct Options
 		{
-			Options() : binderSplit(true), ifixed(-1) {}
+			Options() : binderSplit(true), iquery(1), ifixed(-1) {}
 			bool binderSplit;
+			int iquery;
 			int ifixed;
 		};
 	public:

### 2008-09-04 lib/gringo/include/normalrule.h

@@ -67,9 +67,10 @@ namespace NS_GRINGO
 		struct
 		{
 			unsigned int ground_   : 1;
-			unsigned int once_     : 1;
-			unsigned int next_     : 1;
+			unsigned int base_     : 1;
 			unsigned int last_     : 1;
+			unsigned int lambda_   : 1;
+			unsigned int delta_    : 1;
 			unsigned int isGround_ : 1;
 		};
 	};

### 2008-09-04 lib/gringo/src/grounder.cpp

@@ -29,7 +29,7 @@
 
 using namespace NS_GRINGO;
 
-Grounder::Grounder(const Options &opts) : opts_(opts), incremental_(false), incStep_(1), internalVars_(0), output_(0), eval_(0)
+Grounder::Grounder(const Options &opts) : opts_(opts), incremental_(false), incStep_(0), internalVars_(0), output_(0), eval_(0)
 {
 }
 
@@ -180,6 +180,11 @@ void Grounder::ground()
 {
 	if(incremental_)
 	{
+		for(; incStep_ + 1 <= options().iquery; incStep_++)
+		{
+			reset();
+			ground_();
+		}
 		reset();
 		ground_();
 		incStep_++;
@@ -227,7 +232,7 @@ void Grounder::addProgram(Program *scc)
 
 void Grounder::ground_()
 {
-	if(incStep_ == 1)
+	if(incStep_ == 0)
 		output_->initialize(this, getPred());
 	else
 		output_->reinitialize();
@@ -242,7 +247,7 @@ void Grounder::ground_()
 		for(StatementVector::iterator it = rules->begin(); it !=rules->end(); it++)
 		{
 			Statement *rule = *it;
-			if(incStep_ == 1 || !incremental_)
+			if(incStep_ == 0 || !incremental_)
 				rule->ground(this, PREPARE);
 			else
 				rule->ground(this, REINIT);

### 2008-09-04 lib/gringo/src/literalstatement.cpp

@@ -118,7 +118,7 @@ LiteralStatement::~LiteralStatement()
 
 void LiteralStatement::setIncPart(Grounder *g, IncPart part, const Value &v)
 {
-	std::cerr << "The following statement cant be used with the incremental interface: " << std::endl << "\t" << this << std::endl;
+	std::cerr << "The following statement cant be used with the incremental interface: " << std::endl << "\t" << pp(g, this) << std::endl;
 	throw GrinGoException("Error: unsupported statement.");
 }
 

### 2008-09-04 lib/gringo/src/normalrule.cpp

@@ -32,7 +32,7 @@
 
 using namespace NS_GRINGO;
 
-NormalRule::NormalRule(Literal *head, LiteralVector *body) : Statement(), head_(head), body_(body), grounder_(0), ground_(1), once_(0), next_(0), last_(0), isGround_(0)
+NormalRule::NormalRule(Literal *head, LiteralVector *body) : Statement(), head_(head), body_(body), grounder_(0), ground_(1), base_(0), last_(0), lambda_(0), delta_(0), isGround_(0)
 {
 }
 
@@ -199,9 +199,26 @@ namespace
 bool NormalRule::ground(Grounder *g, GroundStep step)
 {
 	//std::cerr << "grounding: " << this << "(" << step << ")"<< std::endl;
+	// ground the query in the last step
 	if(last_ && g->getIncStep() == g->options().ifixed)
 	{
 		last_   = 0;
+		ground_ = 1;
+		if(step == REINIT)
+			step = PREPARE;
+	}
+	// start grounding cumulative from timestep 1
+	if(lambda_ && g->getIncStep() > 0)
+	{
+		lambda_ = 0;
+		ground_ = 1;
+		if(step == REINIT)
+			step = PREPARE;
+	}
+	// skip queries until iquery is reached
+	if(delta_ && g->getIncStep() >= g->options().iquery)
+	{
+		delta_  = 0;
 		ground_ = 1;
 		if(step == REINIT)
 			step = PREPARE;
@@ -272,7 +289,8 @@ bool NormalRule::ground(Grounder *g, GroundStep step)
 				}
 			}
 			groundOther(g, step, head_, body_);
-			if(once_)
+			// ground the base only one time
+			if(base_)
 				ground_ = 0;
 			break;
 	}
@@ -350,7 +368,6 @@ void NormalRule::grounded(Grounder *g)
 	}
 }
 
-// this looks odd so i hide it :)
 namespace
 {
 	class NormalRuleExpander : public Expandable
@@ -485,7 +502,7 @@ void NormalRule::setIncPart(Grounder *g, IncPart part, const Value &v)
 	switch(part)
 	{
 		case BASE:
-			once_ = 1;
+			base_ = 1;
 			break;
 		case DELTA:
 			if(g->options().ifixed >= 0)
@@ -493,9 +510,19 @@ void NormalRule::setIncPart(Grounder *g, IncPart part, const Value &v)
 				ground_ = 0;
[…]

### 2008-09-04 lib/gringo/src/output.cpp

@@ -176,11 +176,12 @@ void Atom::print_plain(Output *o, std::ostream &out)
 void Atom::print(Output *o, std::ostream &out)
 {
 	if(o->addAtom(this))
+	{
 		if (o->isVisible(predUid_))
 			out << "4" << " 3 " << uid_ << " " << o->atomToString(predUid_, values_) << " " << "1" << END_ENTRY << NL;
 		else
 			out << "4" << " 2 " << uid_ << " " << o->atomToString(predUid_, values_) << END_ENTRY << NL;
-
+	}
 }
 	
 // =============== NS_OUTPUT::Rule ===============

### 2008-09-04 lib/gringo/src/conditionalliteral.cpp
- fixed the bug listed in the TODO list
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@495 3e02f596-3131-0410-8352-e04eeaec37cc

@@ -25,6 +25,7 @@
 #include "literaldependencygraph.h"
 #include "statementchecker.h"
 #include "aggregateliteral.h"
+#include "domain.h"
 
 using namespace NS_GRINGO;
 		
@@ -84,12 +85,71 @@ SDGNode *ConditionalLiteral::createNode(SDG *dg, SDGNode *prev, DependencyAdd to
 	return n;
 }
 
+namespace
+{
+	// TODO: its better reimplement the LDGBuilder to work similar like the SDGBuilder
+	//       to avoid things like this!!!
+	class HeadNodeHelper : public Literal
+	{
+	public:
+		HeadNodeHelper(PredicateLiteral *pred, const VarSet &vars) : pred_(pred), vars_(vars) {}
+		void createNode(LDGBuilder *dg, bool head) 
+		{
+			assert(head);
+			VarSet needed = vars_, provided;
+			if(head || pred_->getNeg() || !pred_->getDomain()->complete())
+			{
+				if(pred_->getArgs())
+					for(TermVector::iterator it = pred_->getArgs()->begin(); it != pred_->getArgs()->end(); it++)
+						(*it)->getVars(needed);
+			}
+			else
+			{
+				if(pred_->getArgs())
+					for(TermVector::iterator it = pred_->getArgs()->begin(); it != pred_->getArgs()->end(); it++)
+						if((*it)->isComplex())
+							(*it)->getVars(needed);
+						else
+							(*it)->getVars(provided);
+			}
+			dg->createNode(pred_, head, needed, provided);
+			delete this;
+		}
+
+		// not used!
+		void getVars(VarSet &vars) const { assert(false); };
+		bool checkO(LiteralVector &unsolved) { assert(false); }
+		void preprocess(Grounder *g, Expandable *e) { assert(false); }
+		void reset() { assert(false); }
+		void finish() { assert(false); }
+		bool solved() { assert(false); }
+		bool isFact(Grounder *g) { assert(false); }
+		Literal* clone() const { assert(false); }
+		IndexedDomain *createIndexedDomain(Grounder *g, VarSet &index) { assert(false); }
+		bool match(Grounder *g) { assert(false); }
+		NS_OUTPUT::Object *convert() { assert(false); }
+		SDGNode *createNode(SDG *dg, SDGNode *prev, DependencyAdd todo) { assert(false); }
[…]

### 2008-09-04 lib/gringo/src/predicateliteral.cpp

@@ -77,7 +77,6 @@ SDGNode *PredicateLiteral::createNode(SDG *dg, SDGNode *prev, DependencyAdd todo
 
 void PredicateLiteral::createNode(LDGBuilder *dg, bool head)
 {
-	// TODO: take the parameter check into consideration
 	VarSet needed, provided;
 	if(head || getNeg() || !predNode_->complete())
 	{

### 2008-09-05 doc/documentation/paper.tex
added grounder statistics, don't know where to output them
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@496 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 At most \texttt{num} incremental steps are done. After them iclingo stops the search for answer sets.
-\paragraph{\texttt{--ibase}}
-With this option enabled, iclingo solves only the base program and ignores the cumulative and volatile part of the program.
-This option can be helpful when debugging.
+\paragraph{\texttt{--iquery=<num>}}
+With this option you can set in which step to start with solving the volatile part of the incremental program.
 \paragraph{\texttt{--istop=<arg>}}

### 2008-09-05 lib/gringo/include/output.h

@@ -51,6 +51,18 @@ namespace NS_GRINGO
 			
 			// must be called if predicates are added after initialize has been called
 			void addSignature();
+			struct Stats
+			{
+				unsigned int rules;
+				unsigned int atoms;
+				unsigned int count;
+				unsigned int sum;
+				unsigned int max;
+				unsigned int min;
+				unsigned int compute;
+				unsigned int optimize;
+			};
+			Stats stats_;
 		protected:
 			int uids_;
 			std::ostream *out_;

### 2008-09-05 lib/gringo/src/output.cpp

@@ -77,6 +77,7 @@ bool Output::addAtom(NS_OUTPUT::Atom *r)
 	{
 		res = atoms_[id].insert(std::make_pair(r->values_, newUid())).first;
 		r->uid_     = res->second;
+		++stats_.atoms;
 		return true;
 	}
 	else
@@ -195,6 +196,7 @@ void Rule::print_plain(Output *o, std::ostream &out)
 	out << " :- ";
 	body_->print_plain(o, out);
 	out << "." << NL;
+	++(o->stats_).rules;
 }
 
 void Rule::print(Output *o, std::ostream &out)
@@ -203,6 +205,7 @@ void Rule::print(Output *o, std::ostream &out)
 	body_->print(o, out);
 	uid_ = o->newUid();
 	out << "5" << " 3 " << uid_ << " " << head_->getUid() << " " << body_->getUid() << END_ENTRY << NL;
+	++(o->stats_).rules;
 }
 
 Rule::~Rule()
@@ -225,6 +228,7 @@ void Fact::print_plain(Output *o, std::ostream &out)
 {
 	head_->print_plain(o, out);
 	out << "." << NL;
+	++(o->stats_).rules;
 }
 
 void Fact::print(Output *o, std::ostream &out)
@@ -232,6 +236,7 @@ void Fact::print(Output *o, std::ostream &out)
 	head_->print(o, out);
 	uid_ = o->newUid();
 	out << "6" << " 2 " << uid_ << " " << head_->getUid() << END_ENTRY << NL;
+	++(o->stats_).rules;
 }
 
 Fact::~Fact()
@@ -254,6 +259,7 @@ void Integrity::print_plain(Output *o, std::ostream &out)
 	out << " :- ";
 	body_->print_plain(o, out);
 	out << "." << NL;
+	++(o->stats_).rules;
 
 }
 
@@ -262,6 +268,7 @@ void Integrity::print(Output *o, std::ostream &out)
 	body_->print(o, out);
 	uid_ = o->newUid();
 	out << "7" << " 2 " << uid_ << " " << body_->getUid() << END_ENTRY << NL;
+	++(o->stats_).rules;
 }
 
 void Integrity::addDomain(bool fact)
@@ -414,15 +421,19 @@ void Aggregate::print_plain(Output *o, std::ostream &out)
 	{
 		case SUM:
 			out << " [";
+			++(o->stats_).sum;
 			break;
 		case COUNT:
 			out << " {";
[…]

### 2008-09-05 lib/clasp/src/program_builder.cpp
- clasp bugfix for cardinality constraints
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@497 3e02f596-3131-0410-8352-e04eeaec37cc

@@ -313,7 +313,12 @@ bool PrgBodyNode::simplifyBody(ProgramBuilder& prg, uint32 bodyId, std::pair<uin
 					pre.setSimplifyHeads(bodyId);
 				}
 			}
-			else if (extended_) { extended_->bound_ -= weight(i); }
+			else if (extended_) { 
+				// subgoal is true: remove from rule and decrease necessary lower bound
+				weight_t w = weight(i);
+				extended_->sumWeights_	-= w;
+				extended_->bound_				-= w; 
+			}
 		}
 		else if (!mark || !prg.vars_.marked(p)) {
 			if (mark) prg.vars_.mark(p);

### 2008-09-05 app/main.cpp
added statistics for the output languages, still incomplete
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@498 3e02f596-3131-0410-8352-e04eeaec37cc

@@ -78,6 +78,7 @@ private:
 	enum State { start_read, start_ground, start_pre, start_solve, end_read, end_ground, end_pre, end_solve };
 	void setState(State s);
 	void ground(Output &output);
+	void printGrounderStats(Output& output) const;
 #ifdef WITH_CLASP
 	void printSatStats() const;
 	void printLpStats() const;
@@ -257,18 +258,24 @@ int MainApp::run(int argc, char **argv)
 		{
 			SmodelsOutput output(&std::cout);
 			ground(output);
+			if (options.verbose | options.stats)
+				printGrounderStats(output);
 			break;
 		}
 		case Options::GRINGO_OUT:
 		{
 			PilsOutput output(&std::cout, options.aspilsOut);
 			ground(output);
+			if (options.verbose | options.stats)
+				printGrounderStats(output);
 			break;
 		}
 		case Options::TEXT_OUT:
 		{
 			LparseOutput output(&std::cout);
 			ground(output);
+			if (options.verbose | options.stats)
+				printGrounderStats(output);
 			break;
 		}
 #ifdef WITH_CLASP
@@ -344,6 +351,21 @@ void MainApp::ground(Output &output)
 		setState(end_ground);
 	}
 }
+
+
+void MainApp::printGrounderStats(Output& output) const
+{
+	cerr << "=== Grounder Statistics ===" << std::endl;
+	cerr << "#rules   : " << output.stats_.rules << std::endl;
+	cerr << "#atoms   : " << output.stats_.atoms << std::endl;
+	cerr << "#count   : " << output.stats_.count << std::endl;
+	cerr << "#sum     : " << output.stats_.sum << std::endl;
+	cerr << "#min     : " << output.stats_.min << std::endl;
+	cerr << "#max     : " << output.stats_.max << std::endl;
+	cerr << "#compute : " << output.stats_.compute << std::endl;
+	cerr << "#optimize: " << output.stats_.optimize << std::endl;
+}
+
 
 #ifdef WITH_CLASP
 void MainApp::printSatStats() const

### 2008-09-05 lib/gringo/include/output.h

@@ -62,7 +62,6 @@ namespace NS_GRINGO
 				unsigned int compute;
 				unsigned int optimize;
 			};
-			Stats stats_;
 		protected:
 			int uids_;
 			std::ostream *out_;
@@ -72,6 +71,8 @@ namespace NS_GRINGO
 			std::map<Signature, bool> hide_;
 			std::vector<bool> visible_;
 			GlobalStorage *g_;
+		public:
+			Stats stats_;
 		};
 	
 		struct Object

### 2008-09-05 lib/gringo/src/output.cpp

@@ -25,8 +25,21 @@ using namespace NS_GRINGO;
 using namespace NS_OUTPUT;
 
 const char* END_ENTRY = " 0";
-Output::Output(std::ostream *out) : uids_(1), out_(out), pred_(0), hideAll_(false), g_(0)
+Output::Output(std::ostream *out) : uids_(1), out_(out), pred_(0), hideAll_(false), g_(0)//,
+//												stats_.rules(0)//, stats_.atoms(0), stats_.count(0),
+//												stats_.sum(0), stats_.max(0), stats_.min(0),
+//												stats_.compute(0), stats_.optimize(0)
+//TODO: Why does this not work
 {
+	stats_.rules    = 0;
+	stats_.atoms    = 0;
+	stats_.count    = 0;
+	stats_.sum      = 0;
+	stats_.count    = 0;
+	stats_.max      = 0;
+	stats_.min      = 0;
+	stats_.compute  = 0;
+	stats_.optimize = 0;
 	
 }
 

### 2008-09-05 lib/gringo/src/smodelsoutput.cpp

@@ -40,6 +40,7 @@ void SmodelsOutput::printBasicRule(int head, const IntVector &pos, const IntVect
 	for(IntVector::const_iterator it = pos.begin(); it != pos.end(); it++)
 		*out_ << " " << *it;
 	*out_ << NL;
+	++stats_.rules;
 }
 
 void SmodelsOutput::printConstraintRule(int head, int bound, const IntVector &pos, const IntVector &neg)
@@ -50,6 +51,8 @@ void SmodelsOutput::printConstraintRule(int head, int bound, const IntVector &po
 	for(IntVector::const_iterator it = pos.begin(); it != pos.end(); it++)
 		*out_ << " " << *it;
 	*out_ << NL;
+	++stats_.count;
+	++stats_.rules;
 }
 
 void SmodelsOutput::printChoiceRule(const IntVector &head, const IntVector &pos, const IntVector &neg)
@@ -63,6 +66,8 @@ void SmodelsOutput::printChoiceRule(const IntVector &head, const IntVector &pos,
 	for(IntVector::const_iterator it = pos.begin(); it != pos.end(); it++)
 		*out_ << " " << *it;
 	*out_ << NL;
+	++stats_.count;
+	++stats_.rules;
 }
 
 void SmodelsOutput::printWeightRule(int head, int bound, const IntVector &pos, const IntVector &neg, const IntVector &wPos, const IntVector &wNeg)
@@ -77,6 +82,8 @@ void SmodelsOutput::printWeightRule(int head, int bound, const IntVector &pos, c
 	for(IntVector::const_iterator it = wPos.begin(); it != wPos.end(); it++)
 		*out_ << " " << *it;
 	*out_ << NL;
+	++stats_.sum;
+	++stats_.rules;
 }
 
 void SmodelsOutput::printMinimizeRule(const IntVector &pos, const IntVector &neg, const IntVector &wPos, const IntVector &wNeg)
@@ -91,6 +98,7 @@ void SmodelsOutput::printMinimizeRule(const IntVector &pos, const IntVector &neg
 	for(IntVector::const_iterator it = wPos.begin(); it != wPos.end(); it++)
 		*out_ << " " << *it;
 	*out_ << NL;
+	++stats_.optimize;
 }
 
 void SmodelsOutput::printDisjunctiveRule(const IntVector &head, const IntVector &pos, const IntVector &neg)
[…]

### 2008-09-05 app/main.cpp
Hopefully finished grounding statistics
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@499 3e02f596-3131-0410-8352-e04eeaec37cc

@@ -334,6 +334,9 @@ void MainApp::ground(Output &output)
 		if(!parser.parse(&output))
 			throw NS_GRINGO::GrinGoException("Error: Parsing failed.");
 		setState(end_read);
+		// just an approximation
+		if (options.outf == Options::TEXT_OUT)
+			output.stats_.atoms = parser.getDomains()->size();
 	}
 	else
 	{
@@ -349,6 +352,9 @@ void MainApp::ground(Output &output)
 			cerr << "Grounding..." << endl;
 		grounder.ground();
 		setState(end_ground);
+		// just an approximation
+		if (options.outf == Options::TEXT_OUT)
+			output.stats_.atoms = grounder.getDomains()->size();
 	}
 }
 

### 2008-09-05 lib/gringo/src/smodelsoutput.cpp

@@ -149,6 +149,7 @@ void SmodelsOutput::finalize(bool last)
 		// number of models
 		*out_ << models_ << NL;
 		out_->flush();
+		stats_.atoms = uids_-1;
 	}
 }
 
